### 2021-08-24 paper/paper.tex
Initial commit

\title{Small induced universal graphs}
\author{
        James Trimble \\
                School of Computing Science\\
        University of Glasgow, Glasgow, Scotland \\
}
\date{\today}

\documentclass[12pt]{article}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage[x11names, svgnames, rgb]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{snakes,arrows,shapes}
\usepackage{amsmath}
%

\begin{document}
\newcommand{\calF}{\ensuremath{\mathcal{F}}}

\maketitle
%
\section{Introduction}

Given a collection $\calF$ of graphs, graph $G$ is \emph{induced
universal} for $\calF$ if and only if each element of $\calF$ is an induced
subgraph of $G$.  Graph $G$ is a \emph{minimal} induced universal graph
for $\calF$ is it has as few vertices as possible.  We write $f(k)$ to denote
    the order (number of vertices) of a minimal induced universal graph for the
    family of all graphs on $k$ vertices.  Moon showed that $f(k) \leq 2^{(k-1)/2}$
    \cite{moon_1965}, and Alon showed that $f(k) = (1 + o(1))2^{(k-1)/2}$
    \cite{alon2017asymptotically}.  There is an extensive literature on bounds
    on the order of minimal induced universal subgraphs for many families of
    graph; see the references in \cite{alon2017asymptotically}.  However, I am
    not aware of any existing systematic attempts to find exact values for
    families of small graphs.  This paper presents exact values of $f(k)$ for
    $k \leq 5$, obtained by brute-force computational search.  The
    corresponding counts of minimal induced universal subgraphs, and results for
    families of trees, are also
    presented.

\section{Method}

A simple brute-force approach was used to find the set of
$n$-vertex graphs that are induced universal for the family $\calF$ of all $k$-vertex
graphs.  Each possible $n$-vertex graph $G$ from Brendan McKay's lists of
graphs\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/graphs.html}} was
tested in turn.  We used a Python implementation of the McSplit
algorithm \cite{DBLP:conf/ijcai/McCreeshPT17} to test whether each $k$-vertex
graph is an induced subgraph of $G$.  Although the McSplit algorithm was designed for
the more general maximum common induced subgraph problem, it performs well on
the subgraph isomorphism problem if both input graphs are small.

For those graphs $G$ that do not contain a copy of each element of $\calF$, it
is desirable to iterate over $\calF$ in an order that allows us to quickly
reject $G$ by finding a graph that is not an induced subgraph of $G$.  The approach we
used was to sort $\calF$ in descending order of number of isomorphisms.  In
particular, this means that the clique $K_k$ and the independent set $I_k$ are
the first two graphs tested.

The full set of experiments described in this paper, run sequentially, required
around 14 million calls to the subgraph isomorphism algorithm and took around
12 minutes to complete on a laptop.

\section{Results}

Table~\ref{tab:graphresults} shows, for $0 \leq k \leq 5$, the value of $f(k)$.
The table also shows $F(k)$, the number of minimal induced universal
graphs for the family of all $k$-vertex graphs.  To my knowledge, the value of
$f(10)$ and the values of $F(k)$ have not been published previously.

The values $f(4)$ and $f(5)$ are equal to a simple lower bound
given in a comment by ``bof'' on Mathematics Stack Exchange
\cite{math_se_question}: $f(k) \geq 2k$ if $k \geq 4$.  (For $k < 10$, this
bound improves upon Moon's lower bound.)  To briefly summarise the proof, if $k \leq 2k$
then $G$ must be a split graph (that is, a graph whose vertices can be
partitioned into a clique and an independent set); therefore, $G$ cannot
contain the cycle $C_4$ as an induced subgraph.  An example of an 8-vertex
induced universal graph for the family of 4-vertex graphs was given by ``Chain
Markov'' as a comment on the same question.

\begin{table}[h!]
\centering
\begin{tabular}{r r r}
 \toprule
 $k$ & $f(k)$ & $F(k)$ \\ [0.5ex]
 \midrule
 0 & 0 & 1 \\
 1 & 1 & 1 \\
 2 & 3 & 2 \\
 3 & 5 & 5 \\
 4 & 8 & 438 \\
 5 & 10 & 22 \\
 \bottomrule
\end{tabular}
\caption{For each $k$, $f(k)$ is the minimum order of a graph containing all $k$-vertex graphs as
induced subgraphs, and $F(k)$ is the number of distinct $f(k)$-vertex graphs that contain
all $k$-vertex graphs as induced subgraphs.}
\label{tab:graphresults}
\end{table}
%
%

Figure~\ref{fig:graphs} shows three examples of minimal induced universal graphs.  
These are for the families of all graphs with three, four and five vertices
respectively.

\begin{figure}[h]
    \centering

\begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
  \pgfsetlinewidth{.5bp}
%%
\pgfsetcolor{black}
  % Edge: 3 -- 0
  \draw [] (51.166bp,32.248bp) .. controls (44.492bp,40.742bp) and (36.044bp,51.496bp)  .. (29.344bp,60.024bp);
  % Edge: 4 -- 0
  \draw [] (71.737bp,82.653bp) .. controls (60.758bp,80.98bp) and (46.814bp,78.855bp)  .. (35.882bp,77.189bp);
  % Edge: 4 -- 1
  \draw [] (104.4bp,96.268bp) .. controls (113.37bp,102.97bp) and (124.85bp,111.55bp)  .. (133.86bp,118.28bp);
  % Edge: 4 -- 3
  \draw [] (82.931bp,68.405bp) .. controls (78.728bp,58.107bp) and (73.39bp,45.028bp)  .. (69.205bp,34.773bp);
  % Node: 0
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (18.0bp,74.46bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 1
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (148.63bp,129.31bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 2
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (148.0bp,57.0bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 3
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (62.36bp,18.0bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 4
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (89.87bp,85.42bp) ellipse (18.0bp and 18.0bp);
\end{scope}
%
\end{tikzpicture}
\qquad
\qquad
\begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
  \pgfsetlinewidth{.5bp}
%%
\pgfsetcolor{black}
  % Edge: 4 -- 0
  \draw [] (44.956bp,147.01bp) .. controls (40.082bp,141.1bp) and (34.42bp,134.24bp)  .. (29.546bp,128.33bp);
  % Edge: 5 -- 1
  \draw [] (166.49bp,80.229bp) .. controls (168.44bp,90.759bp) and (170.89bp,103.99bp)  .. (172.84bp,114.47bp);
  % Edge: 6 -- 0
  \draw [] (81.617bp,113.74bp) .. controls (68.067bp,113.87bp) and (49.632bp,114.04bp)  .. (36.095bp,114.17bp);
  % Edge: 6 -- 1
  \draw [] (117.47bp,117.89bp) .. controls (129.81bp,120.9bp) and (146.19bp,124.9bp)  .. (158.51bp,127.9bp);
  % Edge: 6 -- 2
  \draw [] (96.884bp,95.599bp) .. controls (94.148bp,78.511bp) and (90.056bp,52.957bp)  .. (87.325bp,35.896bp);
  % Edge: 6 -- 4
  \draw [] (87.484bp,127.04bp) .. controls (81.59bp,133.51bp) and (74.549bp,141.23bp)  .. (68.67bp,147.68bp);
  % Edge: 7 -- 0
  \draw [] (56.422bp,85.611bp) .. controls (48.962bp,91.188bp) and (39.886bp,97.973bp)  .. (32.429bp,103.55bp);
  % Edge: 7 -- 2
  \draw [] (75.094bp,57.147bp) .. controls (76.715bp,50.372bp) and (78.564bp,42.641bp)  .. (80.189bp,35.849bp);
  % Edge: 7 -- 4
  \draw [] (67.894bp,92.673bp) .. controls (65.43bp,107.47bp) and (61.945bp,128.39bp)  .. (59.482bp,143.18bp);
  % Edge: 7 -- 5
  \draw [] (89.1bp,72.32bp) .. controls (105.38bp,70.096bp) and (129.09bp,66.857bp)  .. (145.27bp,64.648bp);
  % Edge: 7 -- 6
  \draw [] (81.628bp,89.246bp) .. controls (84.002bp,92.431bp) and (86.52bp,95.808bp)  .. (88.898bp,99.0bp);
  % Node: 0
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (18.0bp,114.33bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 1
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (176.12bp,132.19bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 2
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (84.46bp,18.0bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 3
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (210.0bp,194.0bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 4
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (56.51bp,161.02bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 5
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (163.15bp,62.21bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 6
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (99.76bp,113.58bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 7
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (70.87bp,74.81bp) ellipse (18.0bp and 18.0bp);
\end{scope}
%
\end{tikzpicture}
\qquad
\qquad
\begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
  \pgfsetlinewidth{.5bp}
%%
\pgfsetcolor{black}
  % Edge: 5 -- 0
  \draw [] (50.975bp,158.84bp) .. controls (45.582bp,156.23bp) and (39.597bp,153.34bp)  .. (34.222bp,150.74bp);
  % Edge: 5 -- 1
  \draw [] (84.643bp,161.37bp) .. controls (96.964bp,157.48bp) and (113.5bp,152.27bp)  .. (125.86bp,148.36bp);
  % Edge: 6 -- 0
  \draw [] (21.443bp,105.03bp) .. controls (20.863bp,111.4bp) and (20.211bp,118.58bp)  .. (19.631bp,124.95bp);
  % Edge: 6 -- 2
  \draw [] (31.856bp,71.036bp) .. controls (37.972bp,59.88bp) and (46.093bp,45.065bp)  .. (52.198bp,33.928bp);
  % Edge: 6 -- 5
  \draw [] (31.841bp,102.81bp) .. controls (39.568bp,116.7bp) and (50.773bp,136.84bp)  .. (58.539bp,150.8bp);
  % Edge: 7 -- 1
  \draw [] (135.92bp,66.785bp) .. controls (137.53bp,83.477bp) and (139.91bp,108.08bp)  .. (141.51bp,124.75bp);
  % Edge: 7 -- 2
  \draw [] (117.55bp,41.664bp) .. controls (105.65bp,36.69bp) and (89.681bp,30.016bp)  .. (77.737bp,25.024bp);
  % Edge: 7 -- 3
  \draw [] (146.34bp,62.563bp) .. controls (149.79bp,66.518bp) and (153.54bp,70.819bp)  .. (156.97bp,74.761bp);
  % Edge: 8 -- 0
  \draw [] (79.976bp,126.94bp) .. controls (66.776bp,130.34bp) and (48.816bp,134.96bp)  .. (35.628bp,138.36bp);
  % Edge: 8 -- 1
  \draw [] (114.12bp,129.78bp) .. controls (118.2bp,131.62bp) and (122.58bp,133.58bp)  .. (126.67bp,135.42bp);
  % Edge: 8 -- 3
  \draw [] (114.19bp,114.53bp) .. controls (125.71bp,109.05bp) and (141.01bp,101.78bp)  .. (152.51bp,96.32bp);
  % Edge: 8 -- 5
  \draw [] (87.424bp,137.42bp) .. controls (84.273bp,142.05bp) and (80.806bp,147.15bp)  .. (77.657bp,151.78bp);
  % Edge: 8 -- 6
  \draw [] (81.103bp,114.54bp) .. controls (68.744bp,108.69bp) and (51.929bp,100.72bp)  .. (39.582bp,94.871bp);
  % Edge: 8 -- 7
  \draw [] (105.76bp,106.01bp) .. controls (111.81bp,93.786bp) and (120.04bp,77.152bp)  .. (126.09bp,64.937bp);
  % Edge: 9 -- 0
  \draw [] (66.194bp,101.08bp) .. controls (55.929bp,109.98bp) and (41.964bp,122.1bp)  .. (31.708bp,131.0bp);
  % Edge: 9 -- 1
  \draw [] (93.68bp,100.8bp) .. controls (104.16bp,109.69bp) and (118.53bp,121.89bp)  .. (129.11bp,130.87bp);
  % Edge: 9 -- 2
  \draw [] (75.241bp,71.56bp) .. controls (72.339bp,60.696bp) and (68.624bp,46.796bp)  .. (65.707bp,35.879bp);
  % Edge: 9 -- 3
  \draw [] (97.955bp,89.017bp) .. controls (113.3bp,88.906bp) and (135.24bp,88.748bp)  .. (150.67bp,88.636bp);
  % Edge: 9 -- 5
  \draw [] (77.041bp,107.16bp) .. controls (75.021bp,119.7bp) and (72.338bp,136.36bp)  .. (70.321bp,148.89bp);
  % Edge: 9 -- 6
  \draw [] (61.949bp,88.484bp) .. controls (55.311bp,88.239bp) and (47.792bp,87.962bp)  .. (41.147bp,87.717bp);
  % Edge: 9 -- 8
  \draw [] (88.59bp,105.38bp) .. controls (88.73bp,105.64bp) and (88.871bp,105.9bp)  .. (89.011bp,106.17bp);
  % Node: 0
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (18.0bp,142.9bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 1
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (143.26bp,142.87bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 2
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (60.93bp,18.0bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 3
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (168.96bp,88.5bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 4
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (210.0bp,18.0bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 5
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (67.44bp,166.8bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 6
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (23.08bp,87.05bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 7
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (134.17bp,48.61bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 8
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (97.65bp,122.39bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 9
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (79.94bp,89.15bp) ellipse (18.0bp and 18.0bp);
\end{scope}
%
\end{tikzpicture}
\label{fig:graphs}
\caption{Induced universal graphs for the families of all
graphs with 3, 4, and 5 vertices}
\end{figure}

\section{Trees}

Table~\ref{tab:treeresults} gives the order $t(k)$ of a minimal induced universal graph for
the family of $k$-vertex trees, and the number $T(k)$ of such graphs.  Figure~\ref{fig:trees}
shows one of the 66 minimal induced universal graphs for the family of 6-vertex trees.

\begin{table}[h!]
\centering
\begin{tabular}{r r r}
 \toprule
 $k$ & $t(k)$ & $T(k)$ \\ [0.5ex]
 \midrule
 0 & 0 & 1 \\
 1 & 1 & 1 \\
 2 & 2 & 1 \\
 3 & 3 & 1 \\
 4 & 5 & 2 \\
 5 & 7 & 18 \\
 6 & 9 & 66 \\
 \bottomrule
\end{tabular}
\caption{For each $k$, $t(k)$ is the minimum order of a graph containing all $k$-vertex trees as
induced subgraphs, and $T(k)$ is the number of distinct $t(k)$-vertex graphs that contain
all $k$-vertex trees as induced subgraphs.}
\label{tab:treeresults}
\end{table}

\begin{figure}[h]
    \centering
\begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
  \pgfsetlinewidth{.5bp}
%%
\pgfsetcolor{black}
  % Edge: 6 -- 0
  \draw [] (33.468bp,122.33bp) .. controls (44.467bp,128.89bp) and (59.197bp,137.66bp)  .. (70.288bp,144.27bp);
  % Edge: 6 -- 1
  \draw [] (30.834bp,100.1bp) .. controls (39.774bp,91.035bp) and (51.646bp,78.993bp)  .. (60.571bp,69.941bp);
  % Edge: 7 -- 0
  \draw [] (108.35bp,211.0bp) .. controls (103.69bp,198.96bp) and (97.434bp,182.81bp)  .. (92.754bp,170.73bp);
  % Edge: 7 -- 2
  \draw [] (128.78bp,216.26bp) .. controls (138.81bp,207.93bp) and (152.29bp,196.74bp)  .. (162.31bp,188.42bp);
  % Edge: 7 -- 3
  \draw [] (111.85bp,245.92bp) .. controls (109.9bp,257.68bp) and (107.35bp,272.94bp)  .. (105.4bp,284.67bp);
  % Edge: 8 -- 0
  \draw [] (134.88bp,109.4bp) .. controls (124.44bp,118.9bp) and (110.07bp,131.98bp)  .. (99.64bp,141.47bp);
  % Edge: 8 -- 1
  \draw [] (132.46bp,88.608bp) .. controls (119.73bp,81.799bp) and (101.96bp,72.297bp)  .. (89.227bp,65.491bp);
  % Edge: 8 -- 2
  \draw [] (154.41bp,114.43bp) .. controls (159.1bp,127.91bp) and (165.58bp,146.51bp)  .. (170.24bp,159.91bp);
  % Edge: 8 -- 4
  \draw [] (166.38bp,93.965bp) .. controls (179.82bp,91.611bp) and (198.11bp,88.409bp)  .. (211.53bp,86.058bp);
  % Edge: 8 -- 5
  \draw [] (152.57bp,79.558bp) .. controls (155.7bp,66.447bp) and (159.96bp,48.609bp)  .. (163.09bp,35.509bp);
  % Node: 0
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (86.17bp,153.73bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 1
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (73.34bp,56.99bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 2
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (176.16bp,176.91bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 3
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (102.42bp,302.6bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 4
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (229.48bp,82.92bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 5
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (167.27bp,18.0bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 6
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (18.0bp,113.12bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 7
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (114.87bp,227.81bp) ellipse (18.0bp and 18.0bp);
\end{scope}
  % Node: 8
\begin{scope}
  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
  \pgfsetstrokecolor{strokecol}
  \draw (148.38bp,97.12bp) ellipse (18.0bp and 18.0bp);
\end{scope}
%
\end{tikzpicture}

\label{fig:trees}
\caption{A universal graph for the family of all
trees with 6 vertices}
\end{figure}

\bibliographystyle{abbrv}
\bibliography{main}
%
\end{document}
%

### 2021-08-24 find_universal_graphs.py

@@ -1,0 +1,56 @@
+import sys
+
+from universalgraphs import Graph, induced_subgraph_isomorphism, from_graph6_bytes
+
+counter = [0]
+
+def read_all_graphs(n):
+    with open("graphs/graph{}.g6".format(n), "rb") as f:
+        for line in f:
+            g = line.strip()
+            yield (from_graph6_bytes(g), g)
+
+    
+def read_all_trees(n):
+    with open("graphs/tree{}.all.txt".format(n), "r") as f:
+        for line in f:
+            tokens = [int(tok) for tok in line.strip().split()]
+            edges = zip(tokens[::2], tokens[1::2])
+            G = Graph(n)
+            for v, w in edges:
+                G.add_edge(v, w)
+            yield G
+
+    
+def iso(P, T):
+    counter[0] += 1
+    return induced_subgraph_isomorphism(P, T)
+
+
+def start():
+    if len(sys.argv) != 4:
+        print("Usage: python3 {} tree|graph NP NT".format(sys.argv[0]))
+        exit(1)
+
+    graph_type = sys.argv[1]
+    nP = int(sys.argv[2])
+    nT = int(sys.argv[3])
+
+    if graph_type == "tree":
+        patterns = [G for G in read_all_trees(nP)]
+    else:
+        patterns = [G for G, _ in read_all_graphs(nP)]
+        if not isinstance(patterns, list):
+            patterns = [patterns]
+
+    patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
+    #patterns.sort(key=lambda G: -abs(sum(sum(row) for row in G._adj_mat) - G.number_of_nodes() * (G.number_of_nodes() - 1) / 2))
+
+    for T, graph6_graph in read_all_graphs(nT):
+        if all(iso(P, T) for P in patterns):
+            print(graph6_graph.decode('ascii'))
+    sys.stderr.write('{} calls\n'.format(counter[0]))
+
+
+if __name__ == "__main__":
+    start()

### 2021-08-24 get_graphs.sh

@@ -1,0 +1,21 @@
+#!/bin/bash
+
+set -eu -o pipefail
+
+cd graphs
+
+# Copy trees with 1-3 vertices
+cp ../extra-graphs/* .
+
+# Get McKay's lists of trees
+for i in $(seq 4 12); do
+    wget https://users.cecs.anu.edu.au/~bdm/data/tree$i.all.tar.gz; tar xvf tree$i.all.tar.gz
+    cat tree$i.*.txt > tree$i.all.txt
+done
+
+# Get McKay's lists of graphs
+for i in $(seq 1 9); do
+    wget https://users.cecs.anu.edu.au/~bdm/data/graph$i.g6
+done
+wget https://users.cecs.anu.edu.au/~bdm/data/graph10.g6.gz
+gunzip graph10.g6.gz

### 2021-08-24 graph6_to_dot.py

@@ -1,0 +1,29 @@
+import sys
+
+from universalgraphs import Graph, from_graph6_bytes
+
+def read_all_graphs(filename):
+    with open(filename, "rb") as f:
+        for line in f:
+            g = line.strip()
+            yield (from_graph6_bytes(g), g)
+
+    
+def start():
+    filename = sys.argv[1]
+    outfile = sys.argv[2]
+    for i, (G, g6) in enumerate(read_all_graphs(filename)):
+        with open(outfile.format(i), 'w') as f:
+            f.write('graph G {\n')
+            f.write('  node[label="", shape=circle]\n')
+            for j in range(G.number_of_nodes()):
+                f.write('  {};\n'.format(j))
+            for v in range(G.number_of_nodes()):
+                for w in range(v):
+                    if G._adj_mat[v][w]:
+                        f.write('  {} -- {};\n'.format(v, w))
+            f.write('}\n')
+
+
+if __name__ == "__main__":
+    start()

### 2021-08-24 run_all.sh

@@ -1,0 +1,12 @@
+#!/bin/bash
+
+set -eu -o pipefail
+
+rm -rf results
+mkdir results
+
+cat exhaustive_tests.txt | while read type p t; do
+    echo $type $p $t
+    pypy3 find_universal_graphs.py $type $p $t > results/$type-$p-$t.txt
+    pypy3 graph6_to_dot.py results/$type-$p-$t.txt dot-graphs/$type-$p-$t-{}.dot
+done

### 2021-08-24 universalgraphs/__init__.py

@@ -1,0 +1,171 @@
+__all__ = ['Graph', 'induced_subgraph_isomorphism', 'from_graph6_bytes']
+
+
+class Graph(object):
+    def __init__(self, node_count = 0):
+        self._adj_mat = [[0] * node_count for _ in range(node_count)]
+
+    def add_node(self):
+        self._adj_mat.append([0] * len(self._adj_mat))
+        for row in self._adj_mat:
+            row.append(0)
+
+    def add_edge(self, v, w):
+        self._adj_mat[v][w] = True
+        self._adj_mat[w][v] = True
+
+    def del_edge(self, v, w):
+        self._adj_mat[v][w] = False
+        self._adj_mat[w][v] = False
+
+    def nodes(self):
+        return list(range(len(self._adj_mat)))
+
+    def number_of_nodes(self):
+        return len(self._adj_mat)
+
+    def adj_row(self, v):
+        return self._adj_mat[v]
+
+    def induced_subgraph(self, vv):
+        result = Graph(len(vv))
+        for i, v in enumerate(vv):
+            for j, w in enumerate(vv):
+                if self._adj_mat[v][w]:
+                    result.add_edge(i, j)
+        return result
+
+    def density(self):
+        n = self.number_of_nodes()
+        if n < 2:
+            return -1
+        return float(sum(sum(row) for row in self._adj_mat)) / (n * (n - 1))
+
+
+class LabelClass(object):
+    """A label class, used by the McSplit algorithm.
+
+    A label class contains a list of nodes from graph G and a list of nodes
+    from graph H to which these may be mapped.
+    """
+
+    def __init__(self, G_nodes, H_nodes):
+        self.G_nodes = G_nodes
+        self.H_nodes = H_nodes
+
+
+class PartitioningSIFinder(object):
+    def __init__(self, G, H, find_all):
+        self.G = G
+        self.H = H
+        self.find_all = find_all
+        self.list_of_mcs = []
+
+    def refine_label_classes(self, label_classes, v, w):
+        new_label_classes = []
+        for lc in label_classes:
+            G_nodes = lc.G_nodes
+            H_nodes = lc.H_nodes
+            G_adj_row_v = self.G.adj_row(v)
+            H_adj_row_w = self.H.adj_row(w)
+            new_lc_0_G_nodes = [u for u in G_nodes if not G_adj_row_v[u]]
[…]

### 2021-08-24 paper/paper.tex
Get upper and lower bounds for f(6)

@@
 \usepackage{tikz}
+\usepackage{amsthm}
 \usetikzlibrary{snakes,arrows,shapes}
@@
 
+\newtheorem{proposition}{Proposition}
+
 \maketitle
 %
+\begin{abstract}
+Using the results of brute-force computational experiments, we give for $0 \leq
+k \leq 5$ the minimum number of vertices $f(k)$ in a graph containing all
+$k$-vertex graphs as induced subgraphs.  We give the corresponding results for
+$0 \leq k \leq 6$ for graphs containing all trees on $k$ vertices.  We also show
+that $13 \leq f(6) \leq 15$.
+\end{abstract}
+
 \section{Introduction}
@@
 subgraph of $G$.  Graph $G$ is a \emph{minimal} induced universal graph
-for $\calF$ is it has as few vertices as possible.  We write $f(k)$ to denote
+for $\calF$ if it has as few vertices as possible.  We write $f(k)$ to denote
     the order (number of vertices) of a minimal induced universal graph for the
@@
     on the order of minimal induced universal subgraphs for many families of
-    graph; see the references in \cite{alon2017asymptotically}.  However, I am
+    graphs; see the references in \cite{alon2017asymptotically}.  However, I am
     not aware of any existing systematic attempts to find exact values for
@@
 
-A simple brute-force approach was used to find the set of
+Given values of $k$ and $n$,
+a simple brute-force approach was used to find the set of
 $n$-vertex graphs that are induced universal for the family $\calF$ of all $k$-vertex
@@
 around 14 million calls to the subgraph isomorphism algorithm and took around
-12 minutes to complete on a laptop.
+12 minutes to complete on a laptop.  \footnote{The code and results from this
+paper, including lists of minimal induced universal graphs graph6 format, are
+available from \url{https://github.com/jamestrimble/small-universal-graphs}}.
 
@@
 \label{fig:graphs}
-\caption{Induced universal graphs for the families of all
+\caption{Minimal induced universal graphs for the families of all
 graphs with 3, 4, and 5 vertices}
 \end{figure}
+
+\section{Bounds on $f(6)$}
+
+To give an upper bound of 15 on $f(6)$, Figure~\ref{fig:adjmat15} shows the
+adjacency matrix of a 15-vertex graph that is induced universal for the family
+of all graphs on six vertices.  This was generated using a specialised random
+graph generator.  We number the vertices from zero; the generator makes the
+$k$ vertices numbered $0$ to $k-1$ a clique, and the $k$ vertices numbered $k-1$ to $2k-2$ an
+independent set.  Each possible edge that is not involved in either the clique
+or the independent set is added with probability $1/2$.  It was only necessary
+to generate 53 graphs with this generator to find a solution.
+
+\begin{figure}[h]
+\centering
+\verb|0 1 1 1 1 1 0 0 0 1 1 0 0 0 1| \\
+\verb|1 0 1 1 1 1 0 0 0 0 1 1 1 1 1| \\
+\verb|1 1 0 1 1 1 1 0 0 1 0 0 1 0 0| \\
+\verb|1 1 1 0 1 1 1 1 1 1 0 0 0 0 0| \\
+\verb|1 1 1 1 0 1 1 1 0 1 1 1 1 0 0| \\
+\verb|1 1 1 1 1 0 0 0 0 0 0 0 0 0 1| \\
+\verb|0 0 1 1 1 0 0 0 0 0 0 0 0 1 0| \\
+\verb|0 0 0 1 1 0 0 0 0 0 0 1 1 1 1| \\
+\verb|0 0 0 1 0 0 0 0 0 0 0 1 1 1 0| \\
+\verb|1 0 1 1 1 0 0 0 0 0 0 1 0 0 0| \\
+\verb|1 1 0 0 1 0 0 0 0 0 0 0 1 0 1| \\
+\verb|0 1 0 0 1 0 0 1 1 1 0 0 0 1 0| \\
+\verb|0 1 1 0 1 0 0 1 1 0 1 0 0 1 1| \\
+\verb|0 1 0 0 0 0 1 1 1 0 0 1 1 0 0| \\
+\verb|1 1 0 0 0 1 0 1 0 0 1 0 1 0 0|
+\caption{The adjacency matrix of a 15-vertex induced universal graph for the family of all
+six-vertex graphs}
+\label{fig:adjmat15}
+\end{figure}
+
+For $k > 5$, it is straightforward to increase by 1 the lower bound by ``bof'' from the previous
+section.
+
+\begin{proposition}
+    $f(k) > 2k + 1$ for all $k > 5$.
+\end{proposition}
+\begin{proof}
+    Suppose, for contradition, that $G$ is an induced universal graph on $2k$
+    vertices for the family of all graphs on $k$ vertices.  Graph $G$ must have
+    $K_k$ and $I_k$ as induced sugraphs.  This clique and independent set must
+    overlap by exactly one vertex (otherwise, as in the proof by ``bof'' from
+    the previous section, $G$ would be a split graph and would not have $C_4$
+    as an induced subgraph).   Thus we can partition the vertices of $G$ into
+    three sets: a clique $S_1$, an independent set $S_2$, and a set $S_3$
+    containing the single remaining vertex.
+
+    Let $S$ be a set of vertices such that the subgraph of $G$ induced by $S$
+    is isomorphic to the cycle graph $C_k$.  $S$ may contain at most two vertices
+    from $S_1$, and these must be adjacent in the induced cycle.  Trivially,
+    $S$ can contain at most one vertex from $S_3$.  Hence, $S$ must contain at
+    least $k-3$ vertices from $S_2$.  At least two of these vertices must be adjacent
+    in the induced $C_k$.  Since $S_2$ is an independent set in $G$, we have a contradiction.
+\end{proof}
+
+Thus, we have that $13 \leq f(6) \leq 15$.
 

### 2021-08-24 find_universal_graphs_heur.py

@@ -1,0 +1,60 @@
+import sys
+import random
+
+from universalgraphs import Graph, induced_subgraph_isomorphism, from_graph6_bytes
+
+counter = [0]
+
+def read_all_graphs(n):
+    with open("graphs/graph{}.g6".format(n), "rb") as f:
+        for line in f:
+            g = line.strip()
+            yield (from_graph6_bytes(g), g)
+
+    
+def random_graph(n, nP):
+    G = Graph(n)
+    for v in range(n):
+        for w in range(v):
+            if v < nP and w < nP:
+                G.add_edge(v, w)
+            elif v >= nP-1 and w >= nP-1 and v < nP*2-1 and w < nP*2-1:
+                pass
+            elif random.random() < .5:
+                G.add_edge(v, w)
+    return G
+
+    
+def iso(P, T):
+    counter[0] += 1
+    return induced_subgraph_isomorphism(P, T)
+
+
+def start():
+    if len(sys.argv) != 5:
+        print("Usage: python3 {} NP NT iters seed".format(sys.argv[0]))
+        exit(1)
+
+    nP = int(sys.argv[1])
+    nT = int(sys.argv[2])
+    iters = int(sys.argv[3])
+    random.seed(int(sys.argv[4]))
+
+    patterns = [G for G, _ in read_all_graphs(nP)]
+    if not isinstance(patterns, list):
+        patterns = [patterns]
+
+    patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
+
+    for i in range(iters):
+        T = random_graph(nT, nP)
+        if all(iso(P, T) for P in patterns):
+            for row in T._adj_mat:
+                print(" ".join(str(int(x)) for x in row))
+            print('success', nP, nT, i)
+            exit(0)
+    print("no luck", nP, nT)
+
+
+if __name__ == "__main__":
+    start()

### 2021-08-24 run_heur.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+time pypy3 find_universal_graphs_heur.py 6 15 5000 1

### 2021-08-24 paper/paper.tex
Add Chain Markov's lower bound

@@
 
-The values $f(4)$ and $f(5)$ are equal to a simple lower bound
-given in a comment by ``bof'' on Mathematics Stack Exchange
-\cite{math_se_question}: $f(k) \geq 2k$ if $k \geq 4$.  (For $k < 10$, this
-bound improves upon Moon's lower bound.)  To briefly summarise the proof, if $k \leq 2k$
-then $G$ must be a split graph (that is, a graph whose vertices can be
-partitioned into a clique and an independent set); therefore, $G$ cannot
-contain the cycle $C_4$ as an induced subgraph.  An example of an 8-vertex
-induced universal graph for the family of 4-vertex graphs was given by ``Chain
-Markov'' as a comment on the same question.
+The values $f(1)$, $f(2)$ and $f(3)$ are equal to to the simple lower bound $2k
+- 1$ given in a question by ``Chain Markov'' on Mathematics Stack Exchange
+\cite{math_se_question}.  The values $f(4)$ and $f(5)$ are equal to a lower
+bound given in a comment by ``bof'' on that question: $f(k) \geq 2k$ if $k \geq
+4$.  (For $k < 10$, this bound improves upon Moon's lower bound.)  To briefly
+summarise the proof, if $k \leq 2k$ then $G$ must be a split graph (that is, a
+graph whose vertices can be partitioned into a clique and an independent set);
+therefore, $G$ cannot contain the cycle $C_4$ as an induced subgraph.  An
+example of an 8-vertex induced universal graph for the family of 4-vertex
+graphs was given by ``Chain Markov'' as a comment on the same question.
 

### 2021-08-27 find_universal_graphs_heur.py
Add local search to heuristic algorithm

@@ -44,16 +44,56 @@ def start():
     if not isinstance(patterns, list):
         patterns = [patterns]
 
-    patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
+    #patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
 
+    print("{} patterns".format(len(patterns)))
+    overall_best = -1
     for i in range(iters):
+        score_cache = {}
         T = random_graph(nT, nP)
-        if all(iso(P, T) for P in patterns):
-            for row in T._adj_mat:
-                print(" ".join(str(int(x)) for x in row))
-            print('success', nP, nT, i)
-            exit(0)
-    print("no luck", nP, nT)
+        best = -1
+        for j in range(1000): #range(i * 10):
+            v, w = choose_edge_to_change(T, nP)
+            change_an_edge(T, v, w)
+            flat_adj_mat = tuple(item for row in T._adj_mat for item in row)
+            if flat_adj_mat in score_cache:
+                score = score_cache[flat_adj_mat]
+            else:
+                score = sum(len(iso(P, T)) for P in patterns)
+                score_cache[flat_adj_mat] = score
+            if score < best:
+                # The change lowered the score, so revert it.
+                change_an_edge(T, v, w)
+            elif score > best:
+                best = score
+                if best > overall_best:
+                    overall_best = best
+            print(score, best, overall_best, " ", len(patterns))
+            if score == len(patterns):
+                print()
+                for row in T._adj_mat:
+                    print(" ".join(str(int(x)) for x in row))
+                print('success', nP, nT, i)
+                exit(0)
+        print()
+
+
+def change_an_edge(G, v, w):
+    G._adj_mat[v][w] = not G._adj_mat[v][w]
+    G._adj_mat[w][v] = not G._adj_mat[w][v]
+
+
+def choose_edge_to_change(G, nP):
+    while True:
+        v = random.randrange(G.number_of_nodes())
+        w = random.randrange(G.number_of_nodes())
+        if v == w:
+            continue
+        if v < nP and w < nP:
+            continue
+        if v >= nP-1 and w >= nP-1 and v < nP*2-1 and w < nP*2-1:
+            continue
+        return v, w
 
 
 if __name__ == "__main__":

### 2021-08-27 paper/paper.tex
Start writing up better bounds for f(6)

@@
 \usepackage{amsmath}
+
+\usepackage{caption}
+\captionsetup{width=0.8\textwidth,font=footnotesize}
 %
@@
 
-To give an upper bound of 15 on $f(6)$, Figure~\ref{fig:adjmat15} shows the
-adjacency matrix of a 15-vertex graph that is induced universal for the family
+For $k > 6$, we can increase by 2 the lower bound by ``bof'' from the previous
+section.
+
+\begin{proposition}\label{f6proposition}
+    $f(k) \geq 2k + 2$ for all $k > 5$.
+\end{proposition}
+\begin{proof}
+
+    Suppose that $G$ is an induced universal graph
+    for the family of all graphs on $k$ vertices, and that $G$ has no
+    more than $2k + 1$ vertices.  Graph $G$ must have
+    $K_k$ and $I_k$ as induced sugraphs.  This clique and independent set may
+    overlap by no more than one vertex.
+    Therefore it is possible to partition the vertices of $G$ into
+    three sets: a clique $S_1$,
+    an independent set $S_2$, and a set $S_3$
+    containing at most 2 vertices.
+
+    We will show that that $G$ can contain as an induced subgraph at most
+    one of the complementary graphs in Figure~\ref{fig:boundproof}, which we
+    call $H$ and $H'$ respectively.
+    First, let $G_1$ be an induced
+    subgraph of $G$ that is isomorphic to $H$.  Since there are
+    no edges between the two three-vertex cliques in $G_1$, it must be the case that the
+    vertex set of at least one of these cliques does not intersect $S_1$.
+    Since $S_2$ is an independent set in $G$, this clique must have exactly
+    one vertex in $S_2$ and two vertices in $S_3$.  We can deduce, then, that
+    $S_3$ contains exactly two vertices, and that these vertices are adjacent in $G$.
+
+\begin{figure}[h]
+    \centering
+
+% \begin{tikzpicture}[>=latex',line join=bevel]
+%   \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.5cm]
+%   \pgfsetlinewidth{.5bp}
+%   \node at (0,2.9) (1) {};
+%   \node at (0,2) (2) {};
+%   \node at (0,1.1) (3) {};
+%   \node at (1.1,2.9) (4) {};
+%   \node at (1.1,2) (5) {};
+%   \node at (1.1,1.1) (6) {};
+%   \draw (1) -- (2) -- (3);
+%   \draw (3) to [out=140,in=220] (1);
+%   \draw (4) -- (5) -- (6);
+%   \draw (6) to [out=40,in=-40] (4);
+%   \draw (1) -- (4);
+%   \draw (1) -- (5);
+%   \draw (1) -- (6);
+%   \draw (2) -- (4);
+%   \draw (2) -- (5);
+%   \draw (2) -- (6);
+%   \draw (3) -- (4);
+%   \draw (3) -- (5);
+%   \draw (3) -- (6);
+% \end{tikzpicture}
+% \qquad \qquad
+% \begin{tikzpicture}
+%   \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.5cm]
+%   \pgfsetlinewidth{.5bp}
+%   \node at (0,2.9) (1) {};
+%   \node at (0,2) (2) {};
+%   \node at (0,1.1) (3) {};
+%   \node at (1.1,2.9) (4) {};
+%   \node at (1.1,2) (5) {};
+%   \node at (1.1,1.1) (6) {};
+% \end{tikzpicture}
+% \qquad \qquad
+\begin{tikzpicture}[>=latex',line join=bevel]
+  \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.5cm]
+  \pgfsetlinewidth{.5bp}
+  \node at (0,2.9) (1) {};
+  \node at (0,2) (2) {};
+  \node at (0,1.1) (3) {};
+  \node at (1.1,2.9) (4) {};
+  \node at (1.1,2) (5) {};
+  \node at (1.1,1.1) (6) {};
+  \draw (1) -- (2) -- (3);
+  \draw (3) to [out=140,in=220] (1);
+  \draw (4) -- (5) -- (6);
+  \draw (6) to [out=140,in=220] (4);
+\end{tikzpicture}
+\qquad \qquad
+\begin{tikzpicture}[>=latex',line join=bevel]
+  \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.5cm]
+  \pgfsetlinewidth{.5bp}
+  \node at (0,2.9) (1) {};
+  \node at (0,2) (2) {};
+  \node at (0,1.1) (3) {};
+  \node at (1.1,2.9) (4) {};
+  \node at (1.1,2) (5) {};
+  \node at (1.1,1.1) (6) {};
+  \draw (1) -- (4);
+  \draw (1) -- (5);
+  \draw (1) -- (6);
+  \draw (2) -- (4);
+  \draw (2) -- (5);
+  \draw (2) -- (6);
+  \draw (3) -- (4);
+  \draw (3) -- (5);
+  \draw (3) -- (6);
+\end{tikzpicture}
+%%
+\caption{If $k > 5$ and $G$ is a graph on $2k + 1$ or fewer vertices, then
+$G$ cannot have as induced subgraphs all four of the following graphs: the
+clique $K_k$, the independent set $I_k$, and
+the two graphs shown. See the the proof of Proposition~\ref{f6proposition}}.
+\label{fig:boundproof}
+\end{figure}
+
+    Now consider graph $H'$.  Since $H'$ is an induced subgraph of $G$, it
+    follows by taking complements of $H'$ and $G$ that $H$ is an induced
+    subgraph of $G'$ (the complement of $G$).  We can repeat the argument
+    of the previous paragraph with the roles of $S_1$ and $S_2$ reversed to
+    show that the two vertices in $S_3$ must be adjacent in the complement of
+    $G$, and therefore not adjacent in $G$.  Since we previously showed that
+    these vertices are adjacent in $G$, we have a contradiction.
+\end{proof}
+
+To give a lower bound of 14 on $f(6)$, Figure~\ref{fig:adjmat14} shows the
+adjacency matrix of a 14-vertex graph that is induced universal for the family
 of all graphs on six vertices.  This was generated using a specialised random
@@
 or the independent set is added with probability $1/2$.  It was only necessary
-to generate 53 graphs with this generator to find a solution.
+to generate 53 graphs with this generator to find a solution TODO fixme.
+
+TODO: give more details of generator
 
@@
 \centering
-\verb|0 1 1 1 1 1 0 0 0 1 1 0 0 0 1| \\
-\verb|1 0 1 1 1 1 0 0 0 0 1 1 1 1 1| \\
-\verb|1 1 0 1 1 1 1 0 0 1 0 0 1 0 0| \\
-\verb|1 1 1 0 1 1 1 1 1 1 0 0 0 0 0| \\
-\verb|1 1 1 1 0 1 1 1 0 1 1 1 1 0 0| \\
-\verb|1 1 1 1 1 0 0 0 0 0 0 0 0 0 1| \\
-\verb|0 0 1 1 1 0 0 0 0 0 0 0 0 1 0| \\
-\verb|0 0 0 1 1 0 0 0 0 0 0 1 1 1 1| \\
-\verb|0 0 0 1 0 0 0 0 0 0 0 1 1 1 0| \\
-\verb|1 0 1 1 1 0 0 0 0 0 0 1 0 0 0| \\
-\verb|1 1 0 0 1 0 0 0 0 0 0 0 1 0 1| \\
-\verb|0 1 0 0 1 0 0 1 1 1 0 0 0 1 0| \\
-\verb|0 1 1 0 1 0 0 1 1 0 1 0 0 1 1| \\
-\verb|0 1 0 0 0 0 1 1 1 0 0 1 1 0 0| \\
-\verb|1 1 0 0 0 1 0 1 0 0 1 0 1 0 0|
-\caption{The adjacency matrix of a 15-vertex induced universal graph for the family of all
+\small
+\verb|0 1 1 1 1 1 0 1 1 0 0 0 1 0| \\
+\verb|1 0 1 1 1 1 1 1 1 1 1 0 0 1| \\
+\verb|1 1 0 1 1 1 0 1 1 0 1 0 1 0| \\
+\verb|1 1 1 0 1 1 1 0 0 0 1 1 0 0| \\
+\verb|1 1 1 1 0 1 0 1 0 0 0 0 0 0| \\
+\verb|1 1 1 1 1 0 0 0 0 0 0 1 0 1| \\
+\verb|0 1 0 1 0 0 0 0 0 0 0 1 1 1| \\
+\verb|1 1 1 0 1 0 0 0 0 0 0 1 1 0| \\
+\verb|1 1 1 0 0 0 0 0 0 0 0 0 1 0| \\
+\verb|0 1 0 0 0 0 0 0 0 0 0 1 0 1| \\
+\verb|0 1 1 1 0 0 0 0 0 0 0 1 1 0| \\
+\verb|0 0 0 1 0 1 1 1 0 1 1 0 0 1| \\
+\verb|1 0 1 0 0 0 1 1 1 0 1 0 0 0| \\
+\verb|0 1 0 0 0 1 1 0 0 1 0 1 0 0|
+\caption{The adjacency matrix of a 14-vertex induced universal graph for the family of all
 six-vertex graphs}
-\label{fig:adjmat15}
+\label{fig:adjmat14}
 \end{figure}
 
-For $k > 5$, it is straightforward to increase by 1 the lower bound by ``bof'' from the previous
-section.
-
-\begin{proposition}
-    $f(k) > 2k + 1$ for all $k > 5$.
-\end{proposition}
-\begin{proof}
-    Suppose, for contradition, that $G$ is an induced universal graph on $2k$
-    vertices for the family of all graphs on $k$ vertices.  Graph $G$ must have
-    $K_k$ and $I_k$ as induced sugraphs.  This clique and independent set must
-    overlap by exactly one vertex (otherwise, as in the proof by ``bof'' from
-    the previous section, $G$ would be a split graph and would not have $C_4$
-    as an induced subgraph).   Thus we can partition the vertices of $G$ into
-    three sets: a clique $S_1$, an independent set $S_2$, and a set $S_3$
-    containing the single remaining vertex.
-
-    Let $S$ be a set of vertices such that the subgraph of $G$ induced by $S$
-    is isomorphic to the cycle graph $C_k$.  $S$ may contain at most two vertices
-    from $S_1$, and these must be adjacent in the induced cycle.  Trivially,
-    $S$ can contain at most one vertex from $S_3$.  Hence, $S$ must contain at
-    least $k-3$ vertices from $S_2$.  At least two of these vertices must be adjacent
-    in the induced $C_k$.  Since $S_2$ is an independent set in $G$, we have a contradiction.
-\end{proof}
-
-Thus, we have that $13 \leq f(6) \leq 15$.
+Thus, we have that $f(6) = 14$.
 
@@
 
-\label{fig:trees}
 \caption{A universal graph for the family of all
 trees with 6 vertices}
+\label{fig:trees}
 \end{figure}

### 2021-08-27 paper/paper.tex
Add better heuristic solver, and finish first draft of paper

@@
-\title{Small induced universal graphs}
+\title{Induced universal graphs for families of small graphs}
 \author{
@@
 \begin{abstract}
-Using the results of brute-force computational experiments, we give for $0 \leq
-k \leq 5$ the minimum number of vertices $f(k)$ in a graph containing all
-$k$-vertex graphs as induced subgraphs.  We give the corresponding results for
-$0 \leq k \leq 6$ for graphs containing all trees on $k$ vertices.  We also show
-that $13 \leq f(6) \leq 15$.
+For $0 \leq
+k \leq 6$, we give the minimum number of vertices $f(k)$ in a graph containing all
+$k$-vertex graphs as induced subgraphs, and show
+that $16 \leq f(7) \leq 18$.  For $0 \leq k \leq 5$, we also give the counts of
+such graphs, as generated by brute-force computer search.  We give additional
+results for small graphs containing all trees on $k$ vertices.
 \end{abstract}
@@
 
-Given a collection $\calF$ of graphs, graph $G$ is \emph{induced
-universal} for $\calF$ if and only if each element of $\calF$ is an induced
-subgraph of $G$.  Graph $G$ is a \emph{minimal} induced universal graph
-for $\calF$ if it has as few vertices as possible.  We write $f(k)$ to denote
-    the order (number of vertices) of a minimal induced universal graph for the
-    family of all graphs on $k$ vertices.  Moon showed that $f(k) \leq 2^{(k-1)/2}$
+Given a collection $\calF$ of graphs, graph $G$ is \emph{induced universal} for
+$\calF$ if and only if each element of $\calF$ is an induced subgraph of $G$.
+Graph $G$ is a \emph{minimal} induced universal graph for $\calF$ if it has as
+few vertices as possible.  We write $f(k)$ to denote the order (that is, the
+number of vertices) of a minimal induced universal graph for the family of all
+graphs on $k$ vertices.  Moon showed that $f(k) \leq 2^{(k-1)/2}$
 \cite{moon_1965}, and Alon showed that $f(k) = (1 + o(1))2^{(k-1)/2}$
-    \cite{alon2017asymptotically}.  There is an extensive literature on bounds
-    on the order of minimal induced universal subgraphs for many families of
-    graphs; see the references in \cite{alon2017asymptotically}.  However, I am
-    not aware of any existing systematic attempts to find exact values for
-    families of small graphs.  This paper presents exact values of $f(k)$ for
-    $k \leq 5$, obtained by brute-force computational search.  The
-    corresponding counts of minimal induced universal subgraphs, and results for
-    families of trees, are also
-    presented.
+\cite{alon2017asymptotically}.  There is an extensive literature on bounds on
+the order of minimal induced universal subgraphs for many families of graphs;
+see the references in \cite{alon2017asymptotically}.  However, I am not aware
+of any existing systematic attempts to find exact values for families of small
+graphs.  This paper presents exact values of $f(k)$ for $k \leq 6$ and bounds
+on $f(7)$; in each case, the upper bounds were determined by brute-force
+computational search and proofs are given for the lower bounds.  The
+corresponding counts of minimal induced universal subgraphs for $0 \leq k \leq
+5$, and results for families of trees, are also presented.
 
-\section{Method}
+\section{Method for generating all induced universal graphs}
 
-Given values of $k$ and $n$,
+Given $k$ and $n$,
 a simple brute-force approach was used to find the set of
-$n$-vertex graphs that are induced universal for the family $\calF$ of all $k$-vertex
+$n$-vertex graphs that are induced universal for the family $\calF(k)$ of all $k$-vertex
 graphs.  Each possible $n$-vertex graph $G$ from Brendan McKay's lists of
@@
 the more general maximum common induced subgraph problem, it performs well on
-the subgraph isomorphism problem if both input graphs are small.
+the subgraph isomorphism problem if both input graphs are small.  (Although
+we have described the application of this method to families of all graphs of a given order,
+our method could easily be applied to any family $\calF$ of graphs, and indeed
+Section~\ref{sec:trees} gives results from the application of this method
+to families of trees.)
 
-For those graphs $G$ that do not contain a copy of each element of $\calF$, it
-is desirable to iterate over $\calF$ in an order that allows us to quickly
-reject $G$ by finding a graph that is not an induced subgraph of $G$.  The approach we
-used was to sort $\calF$ in descending order of number of isomorphisms.  In
-particular, this means that the clique $K_k$ and the independent set $I_k$ are
-the first two graphs tested.
+For those graphs $G$ that do not contain a copy of each element of $\calF(k)$,
+it improves the speed of our algorithm if we can iterate over
+$\calF(k)$ in an order that allows us to quickly reject $G$ by finding a graph
+that is not an induced subgraph of $G$.  The approach we used was to sort
+$\calF(k)$ in descending order of number of isomorphisms.  In particular, this
+means that the clique $K_k$ and the independent set $I_k$ are the first two
+graphs tested.
 
 The full set of experiments described in this paper, run sequentially, required
-around 14 million calls to the subgraph isomorphism algorithm and took around
-12 minutes to complete on a laptop.  \footnote{The code and results from this
-paper, including lists of minimal induced universal graphs graph6 format, are
-available from \url{https://github.com/jamestrimble/small-universal-graphs}}.
+took less than an hour complete on a laptop with an Intel Core i5-6200U CPU
+and 8 GB RAM.\footnote{The code and
+results from this paper, including lists of minimal induced universal graphs
+in graph6 format, are available from
+\url{https://github.com/jamestrimble/small-universal-graphs}}
 
-\section{Results}
+\section{Results for $k \leq 5$}
 
@@
 graphs for the family of all $k$-vertex graphs.  To my knowledge, the value of
-$f(10)$ and the values of $F(k)$ have not been published previously.
+$f(5)$ and the values of $F(k)$ have not been published previously.
 
@@
   \cite{math_se_question}.  The values $f(4)$ and $f(5)$ are equal to a lower
-bound given in a comment by ``bof'' on that question: $f(k) \geq 2k$ if $k \geq
-4$.  (For $k < 10$, this bound improves upon Moon's lower bound.)  To briefly
-summarise the proof, if $k \leq 2k$ then $G$ must be a split graph (that is, a
-graph whose vertices can be partitioned into a clique and an independent set);
-therefore, $G$ cannot contain the cycle $C_4$ as an induced subgraph.  An
-example of an 8-vertex induced universal graph for the family of 4-vertex
-graphs was given by ``Chain Markov'' as a comment on the same question.
+  bound given in a comment by ``bof'' on the same question: $f(k) \geq 2k$ if $k
+  \geq 4$.  (For $k < 10$, this bound improves upon Moon's lower bound $f(k)
+  \leq 2^{(k-1)/2}$.) To briefly summarise the proof, if $k \leq 2k$ then $G$
+  must be a split graph (that is, a graph whose vertices can be partitioned
+  into a clique and an independent set); therefore, $G$ cannot contain the
+  cycle $C_4$ as an induced subgraph.  An example of an 8-vertex induced
+  universal graph for the family of 4-vertex graphs was given by ``Chain
+  Markov'' as a comment on the same question.
 
@@
 
-\begin{figure}[h]
+\begin{figure}[htb]
     \centering
@@
 \end{tikzpicture}
-\label{fig:graphs}
 \caption{Minimal induced universal graphs for the families of all
 graphs with 3, 4, and 5 vertices}
+\label{fig:graphs}
 \end{figure}
 
-\section{Bounds on $f(6)$}
+\section{$f(6) = 14$}\label{sec:f6}
 
-For $k > 6$, we can increase by 2 the lower bound by ``bof'' from the previous
-section.
+This section shows that $f(6) = 14$.  We begin with the lower bound.
+For $k \geq 6$, we can increase by 2 the lower bound by ``bof''.  In particular,
+this means that $f(6) \geq 14$.
 
 \begin{proposition}\label{f6proposition}
-    $f(k) \geq 2k + 2$ for all $k > 5$.
+    $f(k) \geq 2k + 2$ for all $k \geq 6$.
 \end{proposition}
@@
 
-    Suppose that $G$ is an induced universal graph
-    for the family of all graphs on $k$ vertices, and that $G$ has no
-    more than $2k + 1$ vertices.  Graph $G$ must have
-    $K_k$ and $I_k$ as induced sugraphs.  This clique and independent set may
-    overlap by no more than one vertex.
-    Therefore it is possible to partition the vertices of $G$ into
-    three sets: a clique $S_1$,
-    an independent set $S_2$, and a set $S_3$
-    containing at most 2 vertices.
+    Suppose that $G$ is an induced universal graph for the family of all graphs
+    on $k$ vertices, and that $G$ has no more than $2k + 1$ vertices.  Graph
+    $G$ must have $K_k$ and $I_k$ as induced sugraphs.  This clique and
+    independent set may overlap by no more than one vertex, so their union must
+    contain at least $2k - 1$ vertices.  Therefore it is possible to partition
+    the vertices of $G$ into three sets: a clique $S_1$, an independent set
+    $S_2$, and a third set $S_3$ containing at most 2 vertices.
 
-    We will show that that $G$ can contain as an induced subgraph at most
-    one of the complementary graphs in Figure~\ref{fig:boundproof}, which we
-    call $H$ and $H'$ respectively.
-    First, let $G_1$ be an induced
-    subgraph of $G$ that is isomorphic to $H$.  Since there are
-    no edges between the two three-vertex cliques in $G_1$, it must be the case that the
-    vertex set of at least one of these cliques does not intersect $S_1$.
-    Since $S_2$ is an independent set in $G$, this clique must have exactly
-    one vertex in $S_2$ and two vertices in $S_3$.  We can deduce, then, that
-    $S_3$ contains exactly two vertices, and that these vertices are adjacent in $G$.
+    We will show that $G$ cannot contain as induced subgraphs both
+    one of the graphs in Figure~\ref{fig:boundproof}.  These graphs are complements
+    of each other, and we refer to them as $H$ and $H'$ respectively.
 
-\begin{figure}[h]
+\begin{figure}[htb]
     \centering
@@
 
+    Let $G_1$ be an induced
+    subgraph of $G$ that is isomorphic to $H$.  Since there are
+    no edges between the two three-vertex cliques in $G_1$, it must be the case that the
+    vertex set of at least one of these cliques does not intersect $S_1$.
+    Since $S_2$ is an independent set in $G$, this clique must have exactly
+    one vertex in $S_2$ and two vertices in $S_3$.  We can deduce, then, that
+    $S_3$ contains exactly two vertices, and that these vertices are adjacent in $G$.
+
     Now consider graph $H'$.  Since $H'$ is an induced subgraph of $G$, it
@@
     show that the two vertices in $S_3$ must be adjacent in the complement of
-    $G$, and therefore not adjacent in $G$.  Since we previously showed that
+    $G$, and therefore must not be adjacent in $G$.  Since we previously showed that
     these vertices are adjacent in $G$, we have a contradiction.
@@
 
-To give a lower bound of 14 on $f(6)$, Figure~\ref{fig:adjmat14} shows the
+To give an upper bound of 14 on $f(6)$, Figure~\ref{fig:adjmat14} shows the
 adjacency matrix of a 14-vertex graph that is induced universal for the family
-of all graphs on six vertices.  This was generated using a specialised random
-graph generator.  We number the vertices from zero; the generator makes the
+of all graphs on six vertices.  This was generated using a simple local search
+algorithm.  We begin by generating a random graph on 14 vertices as follows.
+We number the vertices from zero; the generator makes the
 $k$ vertices numbered $0$ to $k-1$ a clique, and the $k$ vertices numbered $k-1$ to $2k-2$ an
 independent set.  Each possible edge that is not involved in either the clique
-or the independent set is added with probability $1/2$.  It was only necessary
-to generate 53 graphs with this generator to find a solution TODO fixme.
+or the independent set is added with probability $1/2$.
+We then repeatedly ``flip'' the status of a random edge from present to absent
+or vice versa, but always leave the large clique and independent set intact.
+After each flip, we count the number of 156 graphs on 6 vertices that are isomorphic
+to a subgraph of our 14-vertex graph.  If the most recent flip resulted in a decrease
+in this number, we revert it.  After each 1000 flips, we restart the algorithm
+with a new random graph.
 
-TODO: give more details of generator
-
-\begin{figure}[h]
+\begin{figure}[htb]
 \centering
@@
 
-Thus, we have that $f(6) = 14$.
+\section{Bounds on $f(7)$}
 
-\section{Trees}
+By Proposition~\ref{f6proposition}, we have $f(7) \geq 16$.  Figure~\ref{fig:adjmat18}
+shows the adjacency matrix of an 18-vertex induced universal
+graph for the family of all seven-vertex graphs. This was generated with 
+the heuristic described in Section~\ref{sec:f6}, with 10000 rather than 1000
+edge-flips permitted before each restart.
+
+Thus we have $16 \leq f(7) \leq 18$.
+
+\begin{figure}[htb]
+\centering
+\small
+\verb|0 1 1 1 1 1 1 1 1 1 0 1 1 1 1 1 0 0| \\
+\verb|1 0 1 1 1 1 1 0 1 1 1 0 1 1 0 1 0 0| \\
+\verb|1 1 0 1 1 1 1 1 0 1 0 1 0 0 0 0 0 1| \\
+\verb|1 1 1 0 1 1 1 0 1 0 0 1 0 0 0 0 0 1| \\
+\verb|1 1 1 1 0 1 1 1 1 0 0 0 0 0 1 0 0 0| \\
+\verb|1 1 1 1 1 0 1 0 1 0 0 0 0 1 1 1 1 1| \\
+\verb|1 1 1 1 1 1 0 0 0 0 0 0 0 1 0 0 0 1| \\
+\verb|1 0 1 0 1 0 0 0 0 0 0 0 0 0 1 1 1 0| \\
+\verb|1 1 0 1 1 1 0 0 0 0 0 0 0 1 1 1 0 1| \\
+\verb|1 1 1 0 0 0 0 0 0 0 0 0 0 0 1 1 1 1| \\
+\verb|0 1 0 0 0 0 0 0 0 0 0 0 0 1 0 0 1 0| \\
+\verb|1 0 1 1 0 0 0 0 0 0 0 0 0 0 1 1 1 0| \\
+\verb|1 1 0 0 0 0 0 0 0 0 0 0 0 0 0 1 1 1| \\
+\verb|1 1 0 0 0 1 1 0 1 0 1 0 0 0 0 0 0 1| \\
+\verb|1 0 0 0 1 1 0 1 1 1 0 1 0 0 0 0 0 1| \\
+\verb|1 1 0 0 0 1 0 1 1 1 0 1 1 0 0 0 1 1| \\
+\verb|0 0 0 0 0 1 0 1 0 1 1 1 1 0 0 1 0 1| \\
+\verb|0 0 1 1 0 1 1 0 1 1 0 0 1 1 1 1 1 0|
+\caption{The adjacency matrix of an 18-vertex induced universal graph for the family of all
+seven-vertex graphs}
+\label{fig:adjmat18}
+\end{figure}
+
+\section{Trees}\label{sec:trees}
 

### 2021-08-27 find_universal_graphs_heur.py

@@ -32,34 +32,34 @@ def iso(P, T):
 
 def start():
     if len(sys.argv) != 5:
-        print("Usage: python3 {} NP NT iters seed".format(sys.argv[0]))
+        print("Usage: python3 {} NP NT iters_per_T seed".format(sys.argv[0]))
         exit(1)
 
     nP = int(sys.argv[1])
     nT = int(sys.argv[2])
-    iters = int(sys.argv[3])
+    iters_per_T = int(sys.argv[3])
     random.seed(int(sys.argv[4]))
 
     patterns = [G for G, _ in read_all_graphs(nP)]
     if not isinstance(patterns, list):
         patterns = [patterns]
 
-    #patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
+    patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
 
-    print("{} patterns".format(len(patterns)))
+#    print("{} patterns".format(len(patterns)))
     overall_best = -1
-    for i in range(iters):
+    while True:
         score_cache = {}
         T = random_graph(nT, nP)
         best = -1
-        for j in range(1000): #range(i * 10):
+        for j in range(iters_per_T): #range(i * 10):
             v, w = choose_edge_to_change(T, nP)
             change_an_edge(T, v, w)
             flat_adj_mat = tuple(item for row in T._adj_mat for item in row)
             if flat_adj_mat in score_cache:
                 score = score_cache[flat_adj_mat]
             else:
-                score = sum(len(iso(P, T)) for P in patterns)
+                score = calc_score(T, patterns, best)
                 score_cache[flat_adj_mat] = score
             if score < best:
                 # The change lowered the score, so revert it.
@@ -68,14 +68,24 @@ def start():
                 best = score
                 if best > overall_best:
                     overall_best = best
-            print(score, best, overall_best, " ", len(patterns))
+#            print(score, best, overall_best, " ", len(patterns))
             if score == len(patterns):
                 print()
                 for row in T._adj_mat:
                     print(" ".join(str(int(x)) for x in row))
-                print('success', nP, nT, i)
+                print('success', nP, nT)
                 exit(0)
-        print()
+#        print()
+
+
+def calc_score(T, patterns, best):
+    score = len(patterns)
+    for P in patterns:
+        if not iso(P, T):
+            score -= 1
+            if score < best:
+                return -1
+    return score
 
 
 def change_an_edge(G, v, w):

### 2021-08-27 run_heur.sh

@@ -1,3 +1,4 @@
 #!/bin/bash
 
-time pypy3 find_universal_graphs_heur.py 6 15 5000 1
+time pypy3 find_universal_graphs_heur.py 6 14 1000 1
+time pypy3 find_universal_graphs_heur.py 7 18 10000 1

### 2021-08-31 paper/paper.tex
Add a couple of citations to the introduction

@@
+\documentclass[12pt]{article}
 \title{Induced universal graphs for families of small graphs}
@@
         University of Glasgow, Glasgow, Scotland \\
+        \href{mailto:j.trimble.1@research.gla.ac.uk}{j.trimble.1@research.gla.ac.uk}
 }
@@
 
-\documentclass[12pt]{article}
-\usepackage[a4paper, total={6in, 8in}]{geometry}
+\usepackage[letterpaper, total={6in, 8in}]{geometry}
 \usepackage[x11names, svgnames, rgb]{xcolor}
@@
 
-\begin{document}
 \newcommand{\calF}{\ensuremath{\mathcal{F}}}
@@
 
+\begin{document}
+
 \maketitle
@@
 \begin{abstract}
-For $0 \leq
-k \leq 6$, we give the minimum number of vertices $f(k)$ in a graph containing all
+For $0 \leq k \leq 6$, we give the minimum number of vertices $f(k)$ in a graph containing all
 $k$-vertex graphs as induced subgraphs, and show
@@
 Graph $G$ is a \emph{minimal} induced universal graph for $\calF$ if it has as
-few vertices as possible.  We write $f(k)$ to denote the order (that is, the
+few vertices as possible.  The problem of finding a minimal induced universal graph
+of a family of graphs generalises the minimum common supergraph problem
+\cite{DBLP:journals/computing/BunkeJK00}.
+
+We write $f(k)$ to denote the order (that is, the
 number of vertices) of a minimal induced universal graph for the family of all
@@
 the order of minimal induced universal subgraphs for many families of graphs;
-see the references in \cite{alon2017asymptotically}.  However, I am not aware
-of any existing systematic attempts to find exact values for families of small
-graphs.  This paper presents exact values of $f(k)$ for $k \leq 6$ and bounds
-on $f(7)$; in each case, the upper bounds were determined by brute-force
-computational search and proofs are given for the lower bounds.  The
-corresponding counts of minimal induced universal subgraphs for $0 \leq k \leq
-5$, and results for families of trees, are also presented.
+see the references in \cite{alon2017asymptotically}.
+However, to my knowledge the only
+existing systematic attempt to find exact values for families of small
+graphs is a Mathematics Stack Exchange answer by James Preen that presents results for families of small connected graphs obtained by brute-force search with the Maple programming language \cite{preen_math_se}.
 
-\section{Method for generating all induced universal graphs}
+This paper uses a brute-force approach similar to that of Preen to find minimal
+induced universal graphs for families of small \emph{general} graphs.
+Section~\ref{sec:method} describes our computer search method, including an
+optimisation to reduce run times.  Sections~\ref{sec:results5} presents exact
+values of $f(k)$ for $k \leq 5$ along with
+the corresponding counts of minimal induced universal subgraphs.
+Section~\ref{sec:f6} gives the value of $f(6)$ and and
+Section~\ref{sec:f7} gives bounds on $f(7)$; in each of these sections, the lower
+bound is proven and a graph obtained by heuristic search demonstrating the
+upper bound is given.  Finally, Section~\ref{sec:trees} gives results for
+families of trees.
+
+\section{Generating all induced universal graphs}\label{sec:method}
 
 Given $k$ and $n$,
-a simple brute-force approach was used to find the set of
+I used a simple brute-force approach to find the set of
 $n$-vertex graphs that are induced universal for the family $\calF(k)$ of all $k$-vertex
@@
 graphs\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/graphs.html}} was
-tested in turn.  We used a Python implementation of the McSplit
+tested in turn.  I used a Python implementation of the McSplit
 algorithm \cite{DBLP:conf/ijcai/McCreeshPT17} to test whether each $k$-vertex
@@
 the subgraph isomorphism problem if both input graphs are small.  (Although
-we have described the application of this method to families of all graphs of a given order,
-our method could easily be applied to any family $\calF$ of graphs, and indeed
+I have described the application of this method to families of all graphs of a given order,
+the method could easily be applied to any family $\calF$ of graphs, and indeed
 Section~\ref{sec:trees} gives results from the application of this method
@@
 $\calF(k)$ in an order that allows us to quickly reject $G$ by finding a graph
-that is not an induced subgraph of $G$.  The approach we used was to sort
+that is not an induced subgraph of $G$.  The approach I used was to sort
 $\calF(k)$ in descending order of number of isomorphisms.  In particular, this
@@
 
-\section{Results for $k \leq 5$}
+\section{Results for \texorpdfstring{$k \leq 5$}{k<=5}}\label{sec:results5}
 
@@
 
-Figure~\ref{fig:graphs} shows three examples of minimal induced universal graphs.  
-These are for the families of all graphs with three, four and five vertices
+Figure~\ref{fig:graphs} shows examples of minimal induced universal graphs
+for the families of all graphs with three, four and five vertices
 respectively.
@@
 
-\section{$f(6) = 14$}\label{sec:f6}
+\section{\texorpdfstring{$f(6) = 14$}{f(6)=14}}\label{sec:f6}
 
@@
 
-\section{Bounds on $f(7)$}
+\section{Bounds on \texorpdfstring{$f(7)$}{f(7)}}\label{sec:f7}
 

### 2021-08-31 paper/paper.tex
Delete an extra word
Fix grammar

@@
 
-The full set of experiments described in this paper, run sequentially, required
-took less than an hour complete on a laptop with an Intel Core i5-6200U CPU
+The full set of experiments described in this paper, run sequentially,
+took less than an hour to complete on a laptop with an Intel Core i5-6200U CPU
 and 8 GB RAM.\footnote{The code and
@@
     We will show that $G$ cannot contain as induced subgraphs both
-    one of the graphs in Figure~\ref{fig:boundproof}.  These graphs are complements
+    of the graphs in Figure~\ref{fig:boundproof}.  These graphs are complements
     of each other, and we refer to them as $H$ and $H'$ respectively.

### 2021-09-02 paper/paper.tex
Brendan McKay's suggestions

@@
   \geq 4$.  (For $k < 10$, this bound improves upon Moon's lower bound $f(k)
-  \leq 2^{(k-1)/2}$.) To briefly summarise the proof, if $k \leq 2k$ then $G$
+  \leq 2^{(k-1)/2}$.) To briefly summarise the proof, if $f(k) \leq 2k$ then $G$
   must be a split graph (that is, a graph whose vertices can be partitioned
@@
 $k$ vertices numbered $0$ to $k-1$ a clique, and the $k$ vertices numbered $k-1$ to $2k-2$ an
-independent set.  Each possible edge that is not involved in either the clique
+independent set.\footnote{The clique and the independent set thus have one vertex
+in common.  The proof of Proposition~\ref{f6proposition} can be modified straightforwardly
+to show that there is no 14-vertex induced universal graph for this family of graphs
+that contains a 6-vertex clique and a 6-vertex independent set as
+induced subgraphs with disjoint vertex sets.}
+Each possible edge that is not involved in either the clique
 or the independent set is added with probability $1/2$.
@@
 After each flip, we count the number of 156 graphs on 6 vertices that are isomorphic
-to a subgraph of our 14-vertex graph.  If the most recent flip resulted in a decrease
-in this number, we revert it.  After each 1000 flips, we restart the algorithm
+to a subgraph of our 14-vertex graph.  If the most recent flip decreased
+this number, we revert it.  After each 1000 flips, we restart the algorithm
 with a new random graph.
@@
 graph for the family of all seven-vertex graphs. This was generated with 
-the heuristic described in Section~\ref{sec:f6}, with 10000 rather than 1000
-edge-flips permitted before each restart.
+the heuristic described in Section~\ref{sec:f6}, with two modifications.
+First, 10000 rather than 1000
+edge-flips were permitted before each restart, as this was found to be more effective
+in a preliminary run of the experiment.  Second, the overlapping six-vertex clique
+and independent set were replaced with a clique on seven vertices and an independent
+set on seven vertices.  Again, these had one vertex in common.  (We also tried making
+the clique and independent set vertex-disjoint, but did not find an 18-vertex solution
+in four hours with this approach.)
 
@@
 shows one of the 66 minimal induced universal graphs for the family of 6-vertex trees.
+
+\section{Acknowledgements}
+
+I would like to thank Brendan McKay for helpful feedback on this paper, and Persi
+Diaconis for introducing me to induced universal graphs and for interesting email
+discussions on related topics.
 

### 2021-09-02 find_universal_graphs_heur.py

@@ -12,13 +12,16 @@ def read_all_graphs(n):
             yield (from_graph6_bytes(g), g)
 
     
-def random_graph(n, nP):
+def random_graph(n, nP, independent_set_start):
     G = Graph(n)
     for v in range(n):
         for w in range(v):
             if v < nP and w < nP:
                 G.add_edge(v, w)
-            elif v >= nP-1 and w >= nP-1 and v < nP*2-1 and w < nP*2-1:
+            elif (v >= independent_set_start and
+                  w >= independent_set_start and
+                  v < independent_set_start + nP and
+                  w < independent_set_start + nP):
                 pass
             elif random.random() < .5:
                 G.add_edge(v, w)
@@ -31,14 +34,18 @@ def iso(P, T):
 
 
 def start():
-    if len(sys.argv) != 5:
-        print("Usage: python3 {} NP NT iters_per_T seed".format(sys.argv[0]))
+    if len(sys.argv) < 5:
+        print("Usage: python3 {} NP NT iters_per_T seed [independent-set-start]".format(sys.argv[0]))
         exit(1)
 
     nP = int(sys.argv[1])
     nT = int(sys.argv[2])
     iters_per_T = int(sys.argv[3])
     random.seed(int(sys.argv[4]))
+    if len(sys.argv) == 6:
+        independent_set_start = int(sys.argv[5])
+    else:
+        independent_set_start = nP - 1
 
     patterns = [G for G, _ in read_all_graphs(nP)]
     if not isinstance(patterns, list):
@@ -50,10 +57,10 @@ def start():
     overall_best = -1
     while True:
         score_cache = {}
-        T = random_graph(nT, nP)
+        T = random_graph(nT, nP, independent_set_start)
         best = -1
         for j in range(iters_per_T): #range(i * 10):
-            v, w = choose_edge_to_change(T, nP)
+            v, w = choose_edge_to_change(T, nP, independent_set_start)
             change_an_edge(T, v, w)
             flat_adj_mat = tuple(item for row in T._adj_mat for item in row)
             if flat_adj_mat in score_cache:
@@ -68,7 +75,7 @@ def start():
                 best = score
                 if best > overall_best:
                     overall_best = best
-#            print(score, best, overall_best, " ", len(patterns))
+            print(score, best, overall_best, " ", len(patterns))
             if score == len(patterns):
                 print()
                 for row in T._adj_mat:
[…]

### 2021-10-10 get_graphs.sh
Use geng and gentreeg to generate graphs

@@ -2,20 +2,23 @@
 
 set -eu -o pipefail
 
-cd graphs
+mkdir -p graphs
+
+for i in $(seq 1 12); do
+    ./nauty/gentreeg $i > graphs/tree$i.g6
+done
+
+for i in $(seq 1 10); do
+    ./nauty/geng $i > graphs/graph$i.g6
+done
 
 # Copy trees with 1-3 vertices
-cp ../extra-graphs/* .
+cp extra-graphs/* graphs
+
+cd graphs
 
 # Get McKay's lists of trees
 for i in $(seq 4 12); do
     wget https://users.cecs.anu.edu.au/~bdm/data/tree$i.all.tar.gz; tar xvf tree$i.all.tar.gz
     cat tree$i.*.txt > tree$i.all.txt
 done
-
-# Get McKay's lists of graphs
-for i in $(seq 1 9); do
-    wget https://users.cecs.anu.edu.au/~bdm/data/graph$i.g6
-done
-wget https://users.cecs.anu.edu.au/~bdm/data/graph10.g6.gz
-gunzip graph10.g6.gz

### 2021-10-10 universalgraphs/__init__.py
Don't rely on NetworkX in __init__.py

@@ -159,7 +159,7 @@ def from_graph6_bytes(bytes_in):
     n, data = data_to_n(data)
     nd = (n * (n - 1) // 2 + 5) // 6
     if len(data) != nd:
-        raise NetworkXError(
+        raise ValueError(
             f"Expected {n * (n - 1) // 2} bits but got {len(data) * 6} in graph6"
         )
 

### 2021-10-12 paper/paper.tex
Tweak writing, and add example figure

@@
 \usepackage{amsthm}
-\usetikzlibrary{snakes,arrows,shapes}
+\usetikzlibrary{decorations, decorations.pathreplacing, decorations.pathmorphing,
+    calc, backgrounds, positioning, tikzmark, patterns, matrix, fit, graphs,
+    snakes, arrows, shapes, calligraphy}
+%\usetikzlibrary{snakes,arrows,shapes}
 \usepackage{amsmath}
+\usepackage{subfigure}
+\usepackage{cleveref}
 
@@
 \newtheorem{proposition}{Proposition}
+
+\definecolor{uofguniversityblue}{rgb}{0, 0.219608, 0.396078}
+
+\definecolor{uofgheather}{rgb}{0.356863, 0.32549, 0.490196}
+\definecolor{uofgaquamarine}{rgb}{0.603922, 0.72549, 0.678431}
+\definecolor{uofgslate}{rgb}{0.309804, 0.34902, 0.380392}
+\definecolor{uofgrose}{rgb}{0.823529, 0.470588, 0.709804}
+\definecolor{uofgmocha}{rgb}{0.709804, 0.564706, 0.47451}
+
+\definecolor{uofglawn}{rgb}{0.517647, 0.741176, 0}
+\definecolor{uofgcobalt}{rgb}{0, 0.615686, 0.92549}
+\definecolor{uofgturquoise}{rgb}{0, 0.709804, 0.819608}
+\definecolor{uofgsunshine}{rgb}{1.0, 0.862745, 0.211765}
+\definecolor{uofgpumpkin}{rgb}{1.0, 0.72549, 0.282353}
+\definecolor{uofgthistle}{rgb}{0.584314, 0.070588, 0.447059}
+\definecolor{uofgpillarbox}{rgb}{0.701961, 0.047059, 0}
+\definecolor{uofglavendar}{rgb}{0.356863, 0.301961, 0.580392}
+
+\definecolor{uofgsandstone}{rgb}{0.321569, 0.278431, 0.231373}
+\definecolor{uofgforest}{rgb}{0, 0.317647, 0.2}
+\definecolor{uofgburgundy}{rgb}{0.490196, 0.133333, 0.223529}
+\definecolor{uofgrust}{rgb}{0.603922, 0.227451, 0.023529}
 
@@
 
-We write $f(k)$ to denote the order (that is, the
-number of vertices) of a minimal induced universal graph for the family of all
-graphs on $k$ vertices.  Moon showed that $f(k) \leq 2^{(k-1)/2}$
+We write $\calF(k)$ to denote the family of all graphs on $k$ vertices,
+and we write $f(k)$ to denote the order (that is, the
+number of vertices) of a minimal induced universal graph for $\calF(k)$.
+We write $F(k)$ to denote the number of non-isomorphic graphs of order $f(k)$
+that are induced universal for $f(k)$.
+To give an example, we have $f(3)=5$ and $F(3)=5$. All five of the minimal induced universal graphs
+for $\calF(3)$ are shown in \Cref{fig:univ3a}. \Cref{fig:univ3b} shows the four
+graphs in $\calF(3)$ as induced subgraphs of a single induced universal graph.
+
+Moon showed that $f(k) \leq 2^{(k-1)/2}$
 \cite{moon_1965}, and Alon showed that $f(k) = (1 + o(1))2^{(k-1)/2}$
@@
 However, to my knowledge the only
-existing systematic attempt to find exact values for families of small
-graphs is a Mathematics Stack Exchange answer by James Preen that presents results for families of small connected graphs obtained by brute-force search with the Maple programming language \cite{preen_math_se}.
+existing systematic attempt to find \emph{exact} values for families of small
+graphs is an answer by James Preen on the Mathematics Stack Exchange website that
+presents results for families of small connected graphs obtained by brute-force
+search with the Maple programming language \cite{preen_math_se}.
+
+\begin{figure}[htb]
+    \centering
+
+    \subfigure[][The five induced universal graphs of order 5 for \calF(3) (that is, for the family of all 3-vertex graphs)] {
+      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \graph [nodes={draw, circle, minimum width=.52cm, inner sep=1pt}, circular placement, radius=0.82cm,
+                clockwise=5] {
+                    0,1,2,3,4;
+           3 -- 0;
+           4 -- 0;
+           4 -- 1;
+           4 -- 3;
+        };
+      \end{tikzpicture}
+      \qquad
+      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \graph [nodes={draw, circle, minimum width=.52cm, inner sep=1pt}, circular placement, radius=0.82cm,
+                clockwise=5] {
+                    0,1,2,3,4;
+          3 -- 0;
+          4 -- 0;
+          4 -- 1;
+          4 -- 2;
+          4 -- 3;
+        };
+      \end{tikzpicture}
+      \qquad
+      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \graph [nodes={draw, circle, minimum width=.52cm, inner sep=1pt}, circular placement, radius=0.82cm,
+                clockwise=5] {
+                    0,1,2,3,4;
+          3 -- 0;
+          3 -- 1;
+          4 -- 0;
+          4 -- 2;
+          4 -- 3;
+        };
+      \end{tikzpicture}
+      \qquad
+      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \graph [nodes={draw, circle, minimum width=.52cm, inner sep=1pt}, circular placement, radius=0.82cm,
+                clockwise=5] {
+                    0,1,2,3,4;
+          3 -- 0;
+          3 -- 1;
+          4 -- 0;
+          4 -- 1;
+          4 -- 3;
+        };
+      \end{tikzpicture}
+      \qquad
+      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \graph [nodes={draw, circle, minimum width=.52cm, inner sep=1pt}, circular placement, radius=0.82cm,
+                clockwise=5] {
+                    0,1,2,3,4;
+          3 -- 0;
+          3 -- 1;
+          4 -- 0;
+          4 -- 1;
+          4 -- 2;
+          4 -- 3;
+        };
+      \end{tikzpicture}
+      \label{fig:univ3a}
+    }
+
+    \par\bigskip
+    \subfigure[][A demonstration that the first graph in \Cref{fig:univ3a} is induced universal
+            for $\calF(3)$.  For each graph $G$ in $\calF(3)$ ($I_3$, $K_3$, $P_3$, and
+            a graph with a single edge), an induced subgraph isomorphic to $G$ is shown in a
+            single color.] {
+      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \graph [nodes={draw, circle, minimum width=.52cm, inner sep=1pt}, circular placement, radius=0.82cm,
+                clockwise=5] {
+                    0,1[fill=uofgpumpkin],2[fill=uofgpumpkin],3[fill=uofgpumpkin],4;
+           3 -- 0;
+           4 -- 0;
+           4 -- 1;
+           4 -- 3;
+        };
+      \end{tikzpicture}
+      \qquad
+      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \graph [nodes={draw, circle, minimum width=.52cm, inner sep=1pt}, circular placement, radius=0.82cm,
+                clockwise=5] {
+                    0[fill=uofgturquoise],1,2,3[fill=uofgturquoise],4[fill=uofgturquoise];
+           4 -- 1;
+           3 --[uofgturquoise, ultra thick] 0;
+           4 --[uofgturquoise, ultra thick] 0;
+           4 --[uofgturquoise, ultra thick] 3;
+        };
+      \end{tikzpicture}
+      \qquad
+      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \graph [nodes={draw, circle, minimum width=.52cm, inner sep=1pt}, circular placement, radius=0.82cm,
+                clockwise=5] {
+                    0,1[fill=uofgrose],2,3[fill=uofgrose],4[fill=uofgrose];
+           3 -- 0;
+           4 -- 0;
+           4 --[uofgrose, ultra thick] 1;
+           4 --[uofgrose, ultra thick] 3;
+        };
+      \end{tikzpicture}
+      \qquad
+      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \graph [nodes={draw, circle, minimum width=.52cm, inner sep=1pt}, circular placement, radius=0.82cm,
+                clockwise=5] {
+                    0,1[fill=uofgmocha],2[fill=uofgmocha],3,4[fill=uofgmocha];
+           3 -- 0;
+           4 -- 0;
+           4 --[uofgmocha, ultra thick] 1;
+           4 -- 3;
+        };
+      \end{tikzpicture}
+      \label{fig:univ3b}
+    }
+\caption{Induced universal graphs for the family of all graphs on 3 vertices}
+\label{fig:graphs}
+\end{figure}
 
@@
  \midrule
- 0 & 0 & 1 \\
  1 & 1 & 1 \\

### 2021-10-13 automorphism-experiment/plot.sh
Add automorphism experiment

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+set -eu -o pipefail
+
+gnuplot scatter.gnuplot
+gnuplot deg-scatter.gnuplot
+sed -i -e '19s/^\(\\path.*\)/\% \1/' *.tex # Ciaran's trick
+latexmk -pdf

### 2021-10-13 automorphism-experiment/run.sh

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+set -eu -o pipefail
+
+(
+cd ..
+time pypy3 find_universal_graphs.py graph 5 8 --automorphism-stats | tee automorphism-experiment/results.txt
+)

### 2021-10-14 find_universal_graphs.py
igraph check and additional options

@@ -1,6 +1,26 @@
+import random
 import sys
 
+try:
+    import igraph
+except:
+    pass
+
 from universalgraphs import Graph, induced_subgraph_isomorphism, from_graph6_bytes
+
+def to_igraph_graph(G):
+    g = igraph.Graph()
+    g.add_vertices(G.number_of_nodes())
+    for v in range(G.number_of_nodes()):
+        for w in range(v):
+            if G.adj_row(v)[w]:
+                g.add_edges([(v, w)])
+    return g
+
+def igraph_induced_subgraph_isomorphism(G, H):
+    g = to_igraph_graph(G)
+    h = to_igraph_graph(H)
+    return h.subisomorphic_lad(g, induced=True)
 
 counter = [0]
 
@@ -22,19 +42,72 @@ def read_all_trees(n):
             yield G
 
     
-def iso(P, T):
+def iso(P, T, check_with_igraph):
     counter[0] += 1
-    return induced_subgraph_isomorphism(P, T)
+    result = bool(induced_subgraph_isomorphism(P, T))
+    if check_with_igraph:
+        assert result == igraph_induced_subgraph_isomorphism(P, T)
+    return result
 
+
+def check_automorphism_counts(patterns):
+    for G in patterns:
+        automorphism_count1 = len(induced_subgraph_isomorphism(G, G, True))
+        automorphism_count2 = to_igraph_graph(G).count_isomorphisms_vf2()
+        if automorphism_count1 != automorphism_count2:
+            print("Unexpected automorphism count!")
+            exit(2)
+
+def print_usage_and_exit():
+    print("Usage: python3 {} tree|graph NP NT".format(sys.argv[0]))
+    exit(1)
+
+def print_automorphism_stats_and_exit(nT, patterns):
+    print("auto_count deg_score count")
+    automorphism_counts = [len(induced_subgraph_isomorphism(G, G, True)) for G in patterns]
+    deg_scores = []
+    for G in patterns:
+        deg_scores.append(abs(
+            sum(sum(row) for row in G._adj_mat) - G.number_of_nodes() * (G.number_of_nodes() - 1) // 2
+            ))
+    # c[i] will be number of targets that contain the i^th pattern as an induced subgraph
+    c = [0 for _ in patterns]
+    for T, graph6_graph in read_all_graphs(nT):
+        for i, P in enumerate(patterns):
+            if iso(P, T, False):
+                c[i] += 1
+    for auto_count, score, count in zip(automorphism_counts, deg_scores, c):
[…]

### 2021-10-15 automorphism-experiment/analyse-second-experiment.sh
Experiments with ordering strategies

@@ -1,0 +1,34 @@
+#!/bin/bash
+
+set -eu -o pipefail
+
+#echo "Results for automorphism and degree pattern orders:"
+#cat results-of-second-experiment.txt
+#echo
+for type in graph sample-of-graphs; do
+    echo "Results for "automorphisms" pattern order:"
+    cat results-of-second-experiment_automorphisms_$type.txt | cut -d' ' -f1 | \
+        datamash --header-out min 1 max 1 mean 1 | column -t
+    echo
+    echo "Results for "degree" pattern order:"
+    cat results-of-second-experiment_degree_$type.txt | cut -d' ' -f1 | \
+        datamash --header-out min 1 max 1 mean 1 | column -t
+    echo
+    echo "Results for random pattern order:"
+    cat results-of-second-experiment_random_$type.txt | cut -d' ' -f1 | \
+        datamash --header-out min 1 max 1 mean 1 | column -t
+    echo
+    echo "Results for almost-random pattern order:"
+    cat results-of-second-experiment_almost-random_$type.txt | cut -d' ' -f1 | \
+        datamash --header-out min 1 max 1 mean 1 | column -t
+    echo
+    echo
+done
+
+Rscript reshape.R
+
+gnuplot second-experiment-scatter.gnuplot
+latexmk -pdf second-experiment-plot
+
+gnuplot second-experiment-scatter-using-sample.gnuplot
+latexmk -pdf second-experiment-plot-using-sample

### 2021-10-15 automorphism-experiment/reshape.R

@@ -1,0 +1,22 @@
+library(tidyverse)
+
+for (type in c("graph", "sample-of-graphs")) {
+    autom <- read.delim(paste0("results-of-second-experiment_automorphisms_", type, ".txt"), header=FALSE, col.names=c("calls", "X"), sep=" ")$calls
+    degree <- read.delim(paste0("results-of-second-experiment_degree_", type, ".txt"), header=FALSE, col.names=c("calls", "X"), sep=" ")$calls
+    almost_random <- read.delim(paste0("results-of-second-experiment_almost-random_", type, ".txt"), header=FALSE, col.names=c("calls", "X"), sep=" ")$calls
+    r <- read.delim(paste0("results-of-second-experiment_random_", type, ".txt"), header=FALSE, col.names=c("calls", "X"), sep=" ")$calls
+
+    d <- data.frame(autom, degree, almost_random, r)
+
+    e <- d %>%
+        pivot_longer(everything(), names_to = "method", values_to = "calls")
+
+    e$methodnum <- match(e$method, c("autom", "degree", "almost_random", "r"))
+    e$jittered_methodnum <- e$methodnum + runif(nrow(e), -.0, .0)
+    write.table(e, paste0("results-of-second-experiment_", type, ".tsv"), sep="\t", row.names=FALSE, quote=FALSE)
+
+    f <- e %>%
+        group_by(methodnum) %>%
+        summarise(mean=mean(calls))
+    write.table(f, paste0("results-of-second-experiment-summary_", type, ".tsv"), sep="\t", row.names=FALSE, quote=FALSE)
+}

### 2021-10-15 automorphism-experiment/run-second-experiment.sh

@@ -1,0 +1,30 @@
+#!/bin/bash
+
+set -eu -o pipefail
+
+iters=50
+p=5
+t=9
+
+rm -f results-of-second-experiment*.txt
+(
+cd ..
+for type in graph sample-of-graphs; do
+    for i in $(seq 1 $iters); do
+        echo i = $i
+        time pypy3 find_universal_graphs.py $type $p $t >/dev/null 2>> automorphism-experiment/results-of-second-experiment_automorphisms_$type.txt
+    done
+    for i in $(seq 1 $iters); do
+        echo i = $i
+        time pypy3 find_universal_graphs.py $type $p $t --degree-pattern-order >/dev/null 2>> automorphism-experiment/results-of-second-experiment_degree_$type.txt
+    done
+    for i in $(seq 1 $iters); do
+        echo i = $i
+        time pypy3 find_universal_graphs.py $type $p $t --random-pattern-order >/dev/null 2>> automorphism-experiment/results-of-second-experiment_random_$type.txt
+    done
+    for i in $(seq 1 $iters); do
+        echo i = $i
+        time pypy3 find_universal_graphs.py $type $p $t --almost-random-pattern-order >/dev/null 2>> automorphism-experiment/results-of-second-experiment_almost-random_$type.txt
+    done
+done
+)

### 2021-10-15 paper/paper.tex
isomorphisms -> automorphisms (correction)
Start writing more about iteration orders

@@
 
-For those graphs $G$ that do not contain a copy of each element of $\calF(k)$,
-it improves the speed of our algorithm if we can iterate over
-$\calF(k)$ in an order that allows us to quickly reject $G$ by finding a graph
-that is not an induced subgraph of $G$.  The approach I used was to sort
-$\calF(k)$ in descending order of number of isomorphisms.  In particular, this
-means that the clique $K_k$ and the independent set $I_k$ are the first two
-graphs tested.
-
 The full set of experiments described in this paper, run sequentially,
@@
 \url{https://github.com/jamestrimble/small-universal-graphs}}
+
+\section{Iteration order for $\calF$}\label{sec:iteration-order}
+
+When deciding whether a graph $G$ contains any element of $\calF(k)$, are
+some iteration orders of $\calF(k)$ better than others?
+For a graph $G$ that contains a copy of each element of $\calF(k)$, the iteration
+order is of no importance; we much check every element of $\calF(k)$.
+For graphs $G$ that do \emph{not} contain a copy of each element of $\calF(k)$,
+however, we would like to find an order of $\calF(k)$ such that graphs
+that are likely to fail the subgraph isomorphism test are checked early.
+
+As Diaconis and Chatterjee note, \cite{chatterjee2021isomorphisms}, the complete
+graph $K_n$ is the `hardest' graph for a random graph to contain.  This led me
+to consider two possible orderings in which $K_k$ and its complement $I_k$ come
+first.  The first of these approaches is to the sort $\calF(k)$ such that
+graphs with large automorphism appear first.  The second is to place graphs
+with unusually high or low edge counts, as measured by the absolute value of
+${2|E(G)| - {|V(G)| \choose 2}}$, first.
+
+\Cref{fig:scatter} considers these two strategies, using at attempt to find
+an induced universal graph of order 8 $\calF(5)$ as an example. Each point
+represents one of the 34 graphs of order 5; only 17 points are visible as each
+graph and its complement appear in the same place on the charts.  On the vertical
+axis, we have the number of graphs in $\calF(8)$ containing $G$; we would like
+graphs for which this value is as small as possible to appear early in our order.
+On the horizontal axes we have each of our two measures: size of automorphism group
+and `extremeness' of edge count.  The latter measure has a strong correlation
+with the number of graphs containing $G$ as an induced subgraph, suggesting that
+it is a good proxy to use when sorting the list of graphs.
+
+\begin{figure}[htb]
+    \centering
+
+%    \subfigure[][TODO caption] {
+      \includegraphics*{img/automorphisms-scatter.pdf}
+%      \label{fig:scatter-a}
+%    }
+     \qquad
+%    \subfigure[][TODO caption] {
+      \includegraphics*{img/deg-scatter.pdf}
+%      \label{fig:scatter-b}
+%    }
+    \caption{For each graph $G \in \calF(5)$, we plot on the vertical axis
+    the number of graphs in $\calF(8)$ that contain $G$ as an induced subgraph.
+    This value has a strong negative correlation with the size of the automorphism
+    group of $G$ (left plot), but a weaker correlation with a measure of how
+    extreme the edge count of $G$ is (right plot).
+    All axis scales are logarithmic.}
+\label{fig:scatter}
+\end{figure}
 

### 2021-10-15 paper/paper.tex
Writing about strategies

--- 2021-10-12 paper/paper.tex
@@
 \usepackage{cleveref}
+\usepackage[ruled,vlined]{algorithm2e}
+\usepackage{algpseudocode}
 
@@
 
+\crefname{algorithm}{Algorithm}{Algorithms}
+\Crefname{algorithm}{Algorithm}{Algorithms}
+\crefname{algocf}{Algorithm}{Algorithms}
+\Crefname{algocf}{Algorithm}{Algorithms}
+\crefname{figure}{Figure}{Figures}
+\Crefname{figure}{Figure}{Figures}
+\crefname{table}{Table}{Tables}
+\Crefname{table}{Table}{Tables}
+\crefname{section}{Section}{Sections}
+\Crefname{section}{Section}{Sections}
+
 \newcommand{\calF}{\ensuremath{\mathcal{F}}}
+\newcommand{\calG}{\ensuremath{\mathcal{G}}}
+\newcommand{\AlgVar}[1]{\mathit{#1}}
+
+\newcommand{\lineref}[1]{line~\ref{#1}}
+\newcommand{\codelineref}[1]{line~\ref{#1}}
+\newcommand{\linerangeref}[2]{lines~\ref{#1} to~\ref{#2}}
+\newcommand{\Lineref}[1]{Line~\ref{#1}}
+\newcommand{\Linerangeref}[2]{Lines~\ref{#1} to~\ref{#2}}
+
+% https://tex.stackexchange.com/a/184829
+\algnewcommand{\LeftComment}[1]{\(\triangleright\) #1}
 
@@
 
-Given $k$ and $n$,
-I used a simple brute-force approach to find the set of
-$n$-vertex graphs that are induced universal for the family $\calF(k)$ of all $k$-vertex
-graphs.  Each possible $n$-vertex graph $G$ from Brendan McKay's lists of
-graphs\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/graphs.html}} was
-tested in turn.  I used a Python implementation of the McSplit
-algorithm \cite{DBLP:conf/ijcai/McCreeshPT17} to test whether each $k$-vertex
-graph is an induced subgraph of $G$.  Although the McSplit algorithm was designed for
-the more general maximum common induced subgraph problem, it performs well on
-the subgraph isomorphism problem if both input graphs are small.  (Although
-I have described the application of this method to families of all graphs of a given order,
-the method could easily be applied to any family $\calF$ of graphs, and indeed
-Section~\ref{sec:trees} gives results from the application of this method
-to families of trees.)
+Given $\calF$ and $n$,
+we use the brute-force approach described by Preen \cite{preen_math_se} and shown
+in \Cref{alg:brute-force} to find the set of
+$n$-vertex graphs that are induced universal for the family $\calF$.
+The function $\FuncSty{AllInducedUniversalGraphs}()$ is the entry point.
+\Lineref{TryEachG} tests in turn each candidate graph $G$ from the family
+of $n$-vertex graphs, and adds to the collection $\calG$ those that are induced universal
+for $\calF$.  The function $\FuncSty{IsInducedUniversal}()$ tests
+whether a graph $G$ is induced universal for $\calF$ by checking that
+$H$ is isomorphic to a subgraph of $G$ for each $G \in \calF$.
 
-For those graphs $G$ that do not contain a copy of each element of $\calF(k)$,
-it improves the speed of our algorithm if we can iterate over
-$\calF(k)$ in an order that allows us to quickly reject $G$ by finding a graph
-that is not an induced subgraph of $G$.  The approach I used was to sort
-$\calF(k)$ in descending order of number of isomorphisms.  In particular, this
-means that the clique $K_k$ and the independent set $I_k$ are the first two
-graphs tested.
+For the collection $\calF(n)$ of candidate graphs on \lineref{TryEachG}, our
+implementation uses Brendan McKay's lists of
+graphs\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/graphs.html}}.
+These are read in \texttt{graph6}
+format\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/formats.html}}
+from a text file as the program proceeds, and therefore only one candidate
+graph at a time needs to be stored in memory.
+
+For the call $\FuncSty{InducedSubIso}()$ to an algorithm for the induced
+subgraph isomorphism decision problem on line \lineref{CallSubIso},
+our program uses a Python implementation of the McSplit
+algorithm \cite{DBLP:conf/ijcai/McCreeshPT17}.
+McSplit was designed for
+the more general maximum common induced subgraph problem,
+but it can be used for induced subgraph isomorphism with a simple
+modification: we backtrack when the calculated upper bound
+on the order of a common subgraph is less the order of graph $H$.
+This method for solving
+the subgraph isomorphism problem is very fast if both input graphs
+are small, as the graphs considered in this paper are.
+While McSplit is suitable for our purposes in this paper,
+we do not claim that it is the the fastest induced subgraph isomorphism
+solver for very small graphs; it would be useful future work to perform
+an experimental comparison with other subgraph isomorphism solvers.
+%%% (Although
+%%% I have described the application of this method to families of all graphs of a given order,
+%%% the method could easily be applied to any family $\calF$ of graphs, and indeed
+%%% Section~\ref{sec:trees} gives results from the application of this method
+%%% to families of trees.)
+
+\begin{algorithm}[h!]
+\DontPrintSemicolon
+%%% \nl $\FuncSty{Search}(\AlgVar{future},M)$ \;
+%%% \nl \Begin{
+%%% %\nl \lIf {$\AlgVar{future} = \emptyset$ \bf{and} $|M| > |\AlgVar{incumbent}|$}
+%%% \nl \lIf {$|M| > |\AlgVar{incumbent}|$}{$\AlgVar{incumbent} \gets M$} \label{StoreIncumbent}
+%%% %\nl \lIf {$\AlgVar{future} = \emptyset$}{return}
+%%% \medskip
+%%% \nl $\AlgVar{bound} \gets |M|  + \sum_{\langle \setG,\setH \rangle \in \AlgVar{future}} \min(|\setG|,|\setH|)$ \label{CalcBound} \;
+%%% \nl \lIf {$\AlgVar{bound} \leq |\AlgVar{incumbent}|$}{\KwSty{return}} \label{PruneSearch}
+%%% \medskip
+%%% \nl $\langle \setG,\setH \rangle \gets \FuncSty{SelectLabelClass}(\AlgVar{future})$ \label{SelectClass} \;
+%%% \nl $v \gets \FuncSty{SelectVertex}(\setG)$ \label{SelectVertex} \;
+%%% \nl \For {$w \in \setH$ \label{WLoop}} {
+%%% \nl    $\AlgVar{future'} \gets \emptyset$ \label{NewFuture} \;
+%%% \nl    \For {$\langle \setG',\setH'\rangle \in future$ \label{InnerLoop}}{
+%%% \nl        $\setG'' \gets \setG' \cap \N(\graphG, v) \setminus \{v\}$ \label{NewPWithEdge} \;
+%%% \nl        $\setH'' \gets \setH' \cap \N(\graphH, w) \setminus \{w\}$ \;
+%%% \nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$\label{IfNonEmpty}}{
+%%% \nl            $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG'' , \setH'' \rangle\}$ \label{AddToFutureWithEdge}}
+%%% \nl        $\setG'' \gets \setG' \cap \invN(\graphG, v) \setminus \{v\}$ \label{NewPWithoutEdge}  \;
+%%% \nl        $\setH'' \gets \setH' \cap \invN(\graphH, w) \setminus \{w\}$ \;
+%%% \nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$\label{IfNonEmpty2}}{
+%%% \nl            $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG'' , \setH'' \rangle\}$} \label{InnerLoopEnd}
+%%%        }
+%%% \nl   $\FuncSty{Search}(\AlgVar{future'},M\cup \{(v,w)\})$ \label{ExpandWithV} \;
+%%%   }
+%%% \nl $\setG' \gets \setG \setminus \{v\}$ \label{RemoveV} \;
+%%% \nl $\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle \setG,\setH \rangle\}$\;
+%%% \nl \lIf {$\setG' \neq \emptyset$} {$\AlgVar{future} \gets \AlgVar{future} \cup \{\langle \setG',\setH \rangle \}$}
+%%% \nl $\FuncSty{Search}(\AlgVar{future},M)$ \label{ExpandWithoutV} \;
+%%% }
+%%% \;
+\nl $\FuncSty{IsInducedUniversal}(\calF,G)$ \label{BruteForceFun} \;
+\nl \KwData{A family of graphs $\calF$ and a graph $G$.}
+\nl \KwResult{A logical value indicating whether $G$ is induced universal for $\calF$.}
+\nl \Begin{
+\nl    \For {$H \in \calF$ \label{HLoop}}{
+\nl      \lIf {$\neg\FuncSty{InducedSubIso}(H, G)$\label{CallSubIso}}{$\KwSty{return}$~$\AlgVar{false}$}
+       }
+\nl   $\KwSty{return}$~$\AlgVar{true}$ \;
+    }
+\medskip
+\nl $\FuncSty{AllInducedUniversalGraphs}(\calF,n)$ \label{BruteForceFun} \;
+\nl \KwData{A family of graphs $\calF$ and a natural number $n$.}
+\nl \KwResult{The set of all graphs of order $n$ that are induced universal for $\calF$.}
+\nl \Begin{
+\nl    $\calG \gets \emptyset$ \;
+\nl    \For {$G \in \calF(n)$}{
+\nl      \lIf {$\FuncSty{IsInducedUniversal}(\calF,G)$\label{TryEachG}}{$\calG \gets \calG \cup \{G\}$}
+       }
+\nl   $\KwSty{return}$~$\calG$ \;
+    }
+\caption{A brute-force algorithm for finding all order-$n$ induced universal subgraphs of a family $\calF$ of graphs.
+	The entry point is $\FuncSty{AllInducedUniversalGraphs}(\calF,n)$.}
+\label{alg:brute-force}
+\end{algorithm}
 
@@
 \url{https://github.com/jamestrimble/small-universal-graphs}}
+
+\section{Iteration order for $\calF$}\label{sec:iteration-order}
+
+When deciding on \lineref{HLoop} of the algorithm whether a graph $G$ contains
+every element of $\calF$ as an induced subgraph, are some iteration orders of
+$\calF$ better than others?  For a graph $G$ that contains a copy of each
+element of $\calF$, the iteration order is of no importance; we much check
+every element of $\calF$ before returning $\AlgVar{true}$.  For a graph $G$ that
+does \emph{not} contain a copy of each element of $\calF$, however, we would like
+to iterate over $\calF$ in an order such that graphs that are likely to fail the
+subgraph isomorphism test are checked early; that way, we can quickly return
+$\AlgVar{false}$.
+
+We begin by considering the case where $\calF = \calF(k)$ for some $k$; that is,
+$\calF$ is the set of all graphs of order $k$.
+As Diaconis and Chatterjee note, \cite{chatterjee2021isomorphisms}, the complete
+graph $K_n$ is the `hardest' graph of order $n$ for a random graph to contain.
+This led me
+to consider two possible orderings in which $K_k$ and its complement $I_k$ come
+first.  The first of these approaches is to the sort $\calF(k)$ such that
+graphs with large automorphism groups appear first.  (Intuitively, if a graph
+has few automorphisms then we can generate many different labelled graphs
+by permuting the vertex labels.  This gives more `opportunities' for the
+graph to be an induced subgraph of $G$.)
+The second approach considered is to place graphs
+with unusually high or low edge counts, as measured by the absolute value of
+${2|E(G)| - {|V(G)| \choose 2}}$, first.
+
+\Cref{fig:scatter} examines whether graphs that appear early in the orderings
+of $\calF$
+generated by each of these two strategies are contained in few graphs with
+a given number of vertices, as we would hope to be the case.
+For this experiment, we let $\calF = \calF(5)$, and we consider the problem
+of finding induced universal graphs of order 8.
+Each point on the plots
+represents one of the 34 graphs of order 5 (fewer than 34 points are visible
+due to exactly-overlapping points, as we would expect).  On the vertical
+axis, we have the number of graphs in $\calF(8)$ containing $G$; we would like
+graphs for which this value is small to appear early in our order.
+On the horizontal axes we have each of our two measures: size of automorphism group
+and `extremeness' of edge count.  The first of these measures has a strong correlation
+with the number of graphs containing $G$ as an induced subgraph, suggesting that
+it is a good strategy to use when sorting the list of graphs.
+
+\begin{figure}[htb]
+    \centering
+
+%    \subfigure[][TODO caption] {
+      \includegraphics*{img/automorphisms-scatter.pdf}
+%      \label{fig:scatter-a}
+%    }
+     \qquad
+%    \subfigure[][TODO caption] {
+      \includegraphics*{img/deg-scatter.pdf}
+%      \label{fig:scatter-b}
+%    }
+    \caption{For each graph $G \in \calF(5)$, we plot on the vertical axis
+    the number of graphs in $\calF(8)$ that contain $G$ as an induced subgraph.
+    This value has a strong negative correlation with the size of the automorphism
+    group of $G$ (left plot), but a weaker correlation with a measure of how
+    extreme the edge count of $G$ is (right plot).
+    All axis scales are logarithmic.}
+\label{fig:scatter}
+\end{figure}
+
+\subsection{Testing the strategies: $\calF = \calF{5}$}
+
+We now test whether each of our ordering strategies results, as hoped,
+in a reduced number of calls to the subgraph isomorphism function
+in \Cref{alg:brute-force}.  As our first test case, we consider 
+$\FuncSty{AllInducedUniversalGraphs}(\calF(5),9)$ ---
+finding all induced universal graphs of order 9 for the family of
+all graphs of order 5.  We consider four strategies.
+The first two of these are the automorphism-count and `degree extremeness'
+measures described above.  The fourth strategy is a random order.
+Finally, we used one further strategy, which we call ``almost random''.
+Under this strategy, $I_5$ and $K_5$ are placed at the start of the list,
+and the remaining 32 graphs are in random order.  This strategy was not
+intended to be useful in practice, but to give insight into whether
+the success of the first two strategies was purely due to having
+$I_5$ and $K_5$ at the start of the list of graphs.
+For each of these strategies, we ran the program 50 times.
+
+The results are shown in \Cref{fig:second-experiment}.  Over 50 runs,
+the `automorphisms` strategy required a mean of 304,758 calls to the subgraph
+isomorphism solver, while the `degree' strategy required a very similar
+number of calls:  a mean of 304,701.  Surprisingly, the `almost random'
+strategy had the lowest average number of calls of all the strategies,
+at 304,431.  The `random' strategy fared poorly, requiring 822,649
+calls on average and a minimum of 451,287 calls in the best of the 50 runs.
+The variance in number of calls was tiny for all but the random strategy,
+with between 304,000 and 305,000 calls being required for each run.
+
+We note, first, that `degree' strategy appears to be at least as effective
+as the `automorphisms' strategy for this instance, despite the lower
+correlation observed in \Cref{fig:scatter}.  Perhaps most surprising is
+that the `almost random' strategy performed better than any of the other
+strategies.
+
+\begin{figure}[htb]
+    \centering
+
+    \includegraphics*[width=0.7\textwidth]{img/second-experiment-plot}
+
+    \caption{TODO}
+\label{fig:second-experiment}
+\end{figure}
+
+\subsection{Testing the strategies: $\calF \subset \calF{5}$}
+
+TODO abcdef
+
+\begin{figure}[htb]
+    \centering
+
+    \includegraphics*[width=0.7\textwidth]{img/second-experiment-plot-using-sample}
+
+    \caption{TODO}
+\label{fig:second-experiment-using-sample}
+\end{figure}
 

### 2021-10-15 find_universal_graphs.py
Add extra order

@@ -93,6 +93,8 @@ def start():
             print_automorphism_stats = True
         elif arg == "--random-pattern-order":
             pattern_order = "random"
+        elif arg == "--almost-random-pattern-order":
+            pattern_order = "almost-random"
         elif arg == "--degree-pattern-order":
             pattern_order = "degree"
         else:
@@ -115,17 +117,26 @@ def start():
         patterns = [G for G, _ in read_all_graphs(nP)]
         if not isinstance(patterns, list):
             patterns = [patterns]
+        if graph_type == "sample-of-graphs":
+            random.shuffle(patterns)
+            patterns = patterns[:len(patterns) // 2]
 
     if check_with_igraph:
         check_automorphism_counts(patterns)
         sys.stderr.write('Checked automorphism counts.\n')
 
-    if pattern_order == "random":
-        random.shuffle(patterns)
+    random.shuffle(patterns)
+    if pattern_order == "automorphisms":
+        patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
+    elif pattern_order == "almost-random":
+        # Move I_{nP} and K_{nP} to the start
+        patterns.sort(
+            key=lambda G: sum(sum(row) for row in G._adj_mat) == G.number_of_nodes() * (G.number_of_nodes() - 1),
+            reverse=True
+        )
+        patterns.sort(key=lambda G: sum(sum(row) for row in G._adj_mat) == 0, reverse=True)
     elif pattern_order == "degree":
         patterns.sort(key=lambda G: -abs(sum(sum(row) for row in G._adj_mat) - G.number_of_nodes() * (G.number_of_nodes() - 1) / 2))
-    else:
-        patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
 
     if print_automorphism_stats:
         print_automorphism_stats_and_exit(nT, patterns)

### 2021-10-15 find_universal_graphs.py
Randomise order of I_k and K_k

@@ -131,10 +131,9 @@ def start():
     elif pattern_order == "almost-random":
         # Move I_{nP} and K_{nP} to the start
         patterns.sort(
-            key=lambda G: sum(sum(row) for row in G._adj_mat) == G.number_of_nodes() * (G.number_of_nodes() - 1),
+            key=lambda G: sum(sum(row) for row in G._adj_mat) in [0, G.number_of_nodes() * (G.number_of_nodes() - 1)],
             reverse=True
         )
-        patterns.sort(key=lambda G: sum(sum(row) for row in G._adj_mat) == 0, reverse=True)
     elif pattern_order == "degree":
         patterns.sort(key=lambda G: -abs(sum(sum(row) for row in G._adj_mat) - G.number_of_nodes() * (G.number_of_nodes() - 1) / 2))
 

### 2021-10-15 paper/paper.tex
A bit more writing

@@
 
+\begin{figure}[htb]
+    \centering
+
+    \includegraphics*[width=0.7\textwidth]{img/second-experiment-plot}
+
+    \caption{Using each of our four order strategies, we ran
+        $\FuncSty{AllInducedUniversalGraphs}(\calF(5),9)$ 50 times. The figure
+        shows the number calls to the subgraph isomorphism algorithm, in thousands.
+        Horizontal lines show mean number of calls.}
+\label{fig:second-experiment}
+\end{figure}
+
 We note, first, that `degree' strategy appears to be at least as effective
@@
 strategies.
+\Cref{tab:graphs-by-autom-count}
+
+\input{img/graphs5}
+
+\subsection{Testing the strategies: $\calF \subset \calF{5}$}
+
+TODO abcdef
 
@@
 
-    \includegraphics*[width=0.7\textwidth]{img/second-experiment-plot}
-
-    \caption{TODO}
-\label{fig:second-experiment}
-\end{figure}
-
-\subsection{Testing the strategies: $\calF \subset \calF{5}$}
-
-TODO abcdef
-
-\begin{figure}[htb]
-    \centering
-
     \includegraphics*[width=0.7\textwidth]{img/second-experiment-plot-using-sample}
 
-    \caption{TODO}
+    \caption{Using each of our four order strategies, we ran
+        $\FuncSty{AllInducedUniversalGraphs}(\calF,9)$ 50 times, where
+        $\calF$ is a random subset of $\calF(5)$ containing 17 graphs. The figure
+        shows the number calls to the subgraph isomorphism algorithm, in thousands.
+        Horizontal lines show mean number of calls.}
 \label{fig:second-experiment-using-sample}
@@
 
+\begin{figure}[htb]
+    \centering
+      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.1cm]
+      \end{tikzpicture}
+\label{fig:heuristic-regions}
+\end{figure}
+
 \section{Bounds on \texorpdfstring{$f(7)$}{f(7)}}\label{sec:f7}

### 2021-10-15 show_family.py
Show family of 5-vertex graphs

@@ -1,0 +1,134 @@
+import random
+import sys
+import json
+import math
+
+from universalgraphs import Graph, induced_subgraph_isomorphism, from_graph6_bytes
+
+def read_all_graphs(n):
+    with open("graphs/graph{}.g6".format(n), "rb") as f:
+        for line in f:
+            g = line.strip()
+            yield (from_graph6_bytes(g), g)
+
+    
+def read_all_trees(n):
+    with open("graphs/tree{}.all.txt".format(n), "r") as f:
+        for line in f:
+            tokens = [int(tok) for tok in line.strip().split()]
+            edges = zip(tokens[::2], tokens[1::2])
+            G = Graph(n)
+            for v, w in edges:
+                G.add_edge(v, w)
+            yield G
+
+    
+def print_usage_and_exit():
+    print("Usage: python3 {} tree|graph NP".format(sys.argv[0]))
+    exit(1)
+
+
+def graph_summary(G, auto_count):
+    n = G.number_of_nodes()
+    edges = []
+    for i in range(n):
+        for j in range(i+1, n):
+            if G._adj_mat[i][j]:
+                edges.append((i, j))
+    return n, edges, auto_count
+
+
+table_template = r"""
+\begin{{table}}[h!]
+\centering
+\begin{{tabular}}{{c l}}
+ \toprule
+ Automorphisms & Graphs \\ [0.5ex]
+ \midrule
+{}
+ \bottomrule
+\end{{tabular}}
+\caption{{The 34 graphs of order 5, classified by automorphism group size}}
+\label{{tab:graphs-by-autom-count}}
+\end{{table}}
+"""
+
+graph_template = r"""\begin{{tikzpicture}}[>=latex',line join=bevel,scale=.3]
+    {{
+         \tikzstyle{{every node}}=[draw, circle, inner sep=1pt, minimum size=.1cm]
+         {}
+         {}
+    }};
+  \end{{tikzpicture}}"""
+
+def vertex_x(i, n):
+    return math.cos(math.radians(i / n * 360 + 90))
+
+
+def vertex_y(i, n):
+    return math.sin(math.radians(i / n * 360 + 90))
+
+
+def vertex_tikz(n):
+    retval = ""
+    for i in range(n):
+        retval += "\\node ({}) at ({},{}) {{}};".format(i, vertex_x(i, n), vertex_y(i, n))
+    return retval
+
+
+def edge_tikz(edges):
[…]

### 2021-10-15 find_universal_graphs.py
Add seed option

@@ -97,6 +97,8 @@ def start():
             pattern_order = "almost-random"
         elif arg == "--degree-pattern-order":
             pattern_order = "degree"
+        elif arg.startswith("--seed="):
+            random.seed(int(arg[7:]))
         else:
             if arg_count == 0:
                 graph_type = sys.argv[1]

### 2021-10-15 find_universal_graphs.py
Allow -- args before positional args

@@ -101,11 +101,11 @@ def start():
             random.seed(int(arg[7:]))
         else:
             if arg_count == 0:
-                graph_type = sys.argv[1]
+                graph_type = arg
             elif arg_count == 1:
-                nP = int(sys.argv[2])
+                nP = int(arg)
             elif arg_count == 2:
-                nT = int(sys.argv[3])
+                nT = int(arg)
             else:
                 print_usage_and_exit()
             arg_count += 1

### 2021-10-15 find_universal_graphs_heur_for_trees.py
Add a heuristic solver for trees

@@ -1,0 +1,126 @@
+import sys
+import random
+
+from universalgraphs import Graph, induced_subgraph_isomorphism, from_graph6_bytes
+
+counter = [0]
+
+def read_all_trees(n):
+    with open("graphs/tree{}.all.txt".format(n), "r") as f:
+        for line in f:
+            tokens = [int(tok) for tok in line.strip().split()]
+            edges = zip(tokens[::2], tokens[1::2])
+            G = Graph(n)
+            for v, w in edges:
+                G.add_edge(v, w)
+            yield G
+
+def read_all_graphs(n):
+    with open("graphs/graph{}.g6".format(n), "rb") as f:
+        for line in f:
+            g = line.strip()
+            yield (from_graph6_bytes(g), g)
+
+    
+def random_graph(n, nP):
+    G = Graph(n)
+    for v in range(n):
+        for w in range(v):
+            if random.random() < .1:
+                G.add_edge(v, w)
+    return G
+
+    
+def iso(P, T):
+    counter[0] += 1
+    return induced_subgraph_isomorphism(P, T)
+
+
+def start():
+    if len(sys.argv) < 5:
+        print("Usage: python3 {} NP NT iters_per_T seed".format(sys.argv[0]))
+        exit(1)
+
+    nP = int(sys.argv[1])
+    nT = int(sys.argv[2])
+    iters_per_T = int(sys.argv[3])
+    random.seed(int(sys.argv[4]))
+
+    patterns = [G for G in read_all_trees(nP)]
+
+#    for P in patterns:
+#        print(len(induced_subgraph_isomorphism(P, P, True)))
+    patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
+
+#    print("{} patterns".format(len(patterns)))
+    overall_best = -1
+    while True:
+        score_cache = {}
+        T = random_graph(nT, nP)
+        best = -1
+        for _ in range(iters_per_T):
+            v, w = choose_edge_to_change(T, nP)
+            change_an_edge(T, v, w)
+            flat_adj_mat = tuple(item for row in T._adj_mat for item in row)
+            if flat_adj_mat in score_cache:
+                score = score_cache[flat_adj_mat]
+            else:
+                score = calc_score(T, patterns, best)
+                score_cache[flat_adj_mat] = score
+            if score < best:
+                # The change lowered the score, so revert it.
+                change_an_edge(T, v, w)
+            elif score > best:
+                best = score
+                if best > overall_best:
[…]

### 2021-10-15 make-heuristic-regions-fig.py
Add heuristic regions figure

@@ -1,0 +1,32 @@
+n = 14
+k = 6
+
+adjmat = [[0] * n for _ in range(n)]
+
+for i in range(k):
+    for j in range(i):
+        adjmat[i][j] = "\\textbf{1}"
+        adjmat[j][i] = "\\textbf{1}"
+
+for i in range(k - 1, k * 2 - 1):
+    for j in range(k - 1, i):
+        adjmat[i][j] = "\\textbf{0}"
+        adjmat[j][i] = "\\textbf{0}"
+
+for i in range(n):
+    for j in range(n):
+        if i == j:
+            continue
+        if i < k and j < k:
+            continue
+        if i >= k - 1 and i < 2 * k - 1 and j >= k - 1 and j < 2 * k - 1:
+            continue
+        adjmat[i][j] = "$\\boldsymbol{\\cdot}$"
+
+print("\draw[uofgcobalt,fill=uofgcobalt,fill opacity=0.15, thick] (-.5,-5.5) rectangle (5.5,.5);")
+print("\draw[uofgpumpkin,fill=uofgpumpkin,fill opacity=0.15, thick] (4.5,-10.5) rectangle (10.5,-4.5);")
+
+for i in range(n):
+    for j in range(n):
+        print("\\node ({}_{}) at ({},{}) {{{}}};".format(i, j, i, -j, adjmat[i][j]))
+

### 2021-10-15 find_universal_graphs_heur.py
Print less to stdout

@@ -75,12 +75,11 @@ def start():
                 best = score
                 if best > overall_best:
                     overall_best = best
-            print(score, best, overall_best, " ", len(patterns))
+            sys.stderr.write("{} {} {}  {}\n".format(score, best, overall_best, len(patterns)))
             if score == len(patterns):
-                print()
                 for row in T._adj_mat:
                     print(" ".join(str(int(x)) for x in row))
-                print('success', nP, nT)
+                #print('success', nP, nT)
                 exit(0)
 #        print()
 

### 2021-10-15 find_universal_graphs_heur_for_trees.py

@@ -73,7 +73,7 @@ def start():
             elif score > best:
                 best = score
                 if best > overall_best:
-                    print("*", overall_best, len(patterns))
+                    sys.stderr.write("* {} {}\n".format(overall_best, len(patterns)))
                     overall_best = best
             if score == len(patterns):
 #                print()
@@ -87,15 +87,15 @@ def start():
 #                                print('GOOD')
                 for row in T._adj_mat:
                     print(" ".join(str(int(x)) for x in row))
-                print('success', nP, nT)
-                print('graph {')
-                for i in range(nT):
-                    print('  v{};'.format(i))
-                for i in range(nT):
-                    for j in range(i):
-                        if T._adj_mat[i][j]:
-                            print('  v{} -- v{};'.format(i, j))
-                print('}')
+                #### print('success', nP, nT)
+                #### print('graph {')
+                #### for i in range(nT):
+                ####     print('  v{};'.format(i))
+                #### for i in range(nT):
+                ####     for j in range(i):
+                ####         if T._adj_mat[i][j]:
+                ####             print('  v{} -- v{};'.format(i, j))
+                #### print('}')
                 exit(0)
 
 

### 2021-10-15 paper/paper.tex
More writing

@@
 of a family of graphs generalises the minimum common supergraph problem
-\cite{DBLP:journals/computing/BunkeJK00}.
+\cite{DBLP:journals/computing/BunkeJK00}, which applies only to families
+containing exactly two graphs.
 
@@
       \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
-        \graph [nodes={draw, circle, minimum width=.52cm, inner sep=1pt}, circular placement, radius=0.82cm,
+        \graph [nodes={draw, circle, minimum width=.49cm, inner sep=1pt}, circular placement, radius=0.78cm,
                 clockwise=5] {
@@
 For the collection $\calF(n)$ of candidate graphs on \lineref{TryEachG}, our
-implementation uses Brendan McKay's lists of
-graphs\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/graphs.html}}.
+implementation uses trees generated using Brendan McKay's \texttt{geng}
+program \cite{DBLP:journals/jal/McKay98}
+\footnote{The program \texttt{geng} is distributed
+with Nauty; we used version 27.r3.\url{https://pallini.di.uniroma1.it/}}.
 These are read in \texttt{graph6}
@@
 \DontPrintSemicolon
-%%% \nl $\FuncSty{Search}(\AlgVar{future},M)$ \;
-%%% \nl \Begin{
-%%% %\nl \lIf {$\AlgVar{future} = \emptyset$ \bf{and} $|M| > |\AlgVar{incumbent}|$}
-%%% \nl \lIf {$|M| > |\AlgVar{incumbent}|$}{$\AlgVar{incumbent} \gets M$} \label{StoreIncumbent}
-%%% %\nl \lIf {$\AlgVar{future} = \emptyset$}{return}
-%%% \medskip
-%%% \nl $\AlgVar{bound} \gets |M|  + \sum_{\langle \setG,\setH \rangle \in \AlgVar{future}} \min(|\setG|,|\setH|)$ \label{CalcBound} \;
-%%% \nl \lIf {$\AlgVar{bound} \leq |\AlgVar{incumbent}|$}{\KwSty{return}} \label{PruneSearch}
-%%% \medskip
-%%% \nl $\langle \setG,\setH \rangle \gets \FuncSty{SelectLabelClass}(\AlgVar{future})$ \label{SelectClass} \;
-%%% \nl $v \gets \FuncSty{SelectVertex}(\setG)$ \label{SelectVertex} \;
-%%% \nl \For {$w \in \setH$ \label{WLoop}} {
-%%% \nl    $\AlgVar{future'} \gets \emptyset$ \label{NewFuture} \;
-%%% \nl    \For {$\langle \setG',\setH'\rangle \in future$ \label{InnerLoop}}{
-%%% \nl        $\setG'' \gets \setG' \cap \N(\graphG, v) \setminus \{v\}$ \label{NewPWithEdge} \;
-%%% \nl        $\setH'' \gets \setH' \cap \N(\graphH, w) \setminus \{w\}$ \;
-%%% \nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$\label{IfNonEmpty}}{
-%%% \nl            $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG'' , \setH'' \rangle\}$ \label{AddToFutureWithEdge}}
-%%% \nl        $\setG'' \gets \setG' \cap \invN(\graphG, v) \setminus \{v\}$ \label{NewPWithoutEdge}  \;
-%%% \nl        $\setH'' \gets \setH' \cap \invN(\graphH, w) \setminus \{w\}$ \;
-%%% \nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$\label{IfNonEmpty2}}{
-%%% \nl            $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG'' , \setH'' \rangle\}$} \label{InnerLoopEnd}
-%%%        }
-%%% \nl   $\FuncSty{Search}(\AlgVar{future'},M\cup \{(v,w)\})$ \label{ExpandWithV} \;
-%%%   }
-%%% \nl $\setG' \gets \setG \setminus \{v\}$ \label{RemoveV} \;
-%%% \nl $\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle \setG,\setH \rangle\}$\;
-%%% \nl \lIf {$\setG' \neq \emptyset$} {$\AlgVar{future} \gets \AlgVar{future} \cup \{\langle \setG',\setH \rangle \}$}
-%%% \nl $\FuncSty{Search}(\AlgVar{future},M)$ \label{ExpandWithoutV} \;
-%%% }
-%%% \;
 \nl $\FuncSty{IsInducedUniversal}(\calF,G)$ \label{BruteForceFun} \;
@@
 that the `almost random' strategy performed better than any of the other
-strategies.
-\Cref{tab:graphs-by-autom-count}
+strategies; our expectation had been that this simple strategy would
+give some but not all of the benefit of the first two strategies.
+We conjecture that the almost-random strategy is effective because it
+quickly checks a diverse range of graphs of graphs.  Consider, for example,
+\Cref{tab:graphs-by-autom-count}, which shows the graphs of order 5 listed
+in descending order of automorphism count.  This is the order in which they
+are checked under the automorphisms strategy (with the order within each
+row chosen randomly).  Note that the two graphs in the second row are
+similar to the graphs in the first row; the first graphs in each of the first
+two rows, for example, have a maximum common induced subgraph of size 4.
+It seems likely that if a graph contains both graphs in the first row of
+the table as induced subgraphs, then it may also contain the graphs in the
+second row. The `almost random' strategy is likely to choose graphs
+that are quite different from a clique and an independent set after the first
+two graphs, perhaps increasing the probability that one of these graphs
+will not be contained in the larger graph.
 
@@
 
-TODO abcdef
+While the `almost random' strategy was very effective when searching
+for induced universal graphs of \emph{all} graphs of a given order,
+it seems likely that it will be less effective if $\calF$ is a strict
+subset of $\calF(k)$, and in particular if it does not include one
+or both of $I_k$ and $K_k$.  \Cref{fig:second-experiment-using-sample}
+examines this claim.  We generated 25 (TODO 50) subsets of $\calF(5)$,
+with each subset containing exactly half of the elements of $\calF(5)$.
+For each of these, we ran $\FuncSty{AllInducedUniversalGraphs}(\calF,9)$
+using each of the four strategies.  TODO say what happened
 
@@
 
-To give an upper bound of 14 on $f(6)$, Figure~\ref{fig:adjmat14} shows the
-adjacency matrix of a 14-vertex graph that is induced universal for the family
-of all graphs on six vertices.  This was generated using a simple local search
-algorithm.  We begin by generating a random graph on 14 vertices as follows.
-We number the vertices from zero; the generator makes the
+To find an induced universal graph for $\calF(6)$ in order to give an upper bound
+on $f(6)$, we use the following local search
+algorithm, which is shown in \Cref{alg:heuristic}.
+The entry point to the algorithm is function $\FuncSty{FindInducedUniversalGraph}()$,
+which takes as its parameter the maximum number of attempted improvement steps to
+take before restarting from scratch.  The main search function, which is called
+until timeout, is $\FuncSty{FindInducedUniversalGraph'}()$.
+
+\begin{algorithm}[h!]
+\DontPrintSemicolon
+\nl $\FuncSty{FlipEdge}(G, v, w)$  \;
+\nl \Begin{
+\nl      \If {$G$ has edge $\{v,w\}$}{
+    \nl           $\KwSty{return}$~$(V(G), E(G) \setminus \{v, w\})$ \;
+}
+\nl \Else {
+    \nl           $\KwSty{return}$~$(V(G), E(G) \cup \{v, w\})$ \;
+}
+}
+\medskip
+\nl $\FuncSty{Score}(G)$  \;
+\nl \Begin{
+    \nl $\KwSty{return}$~$|\{H \in \calF(6) \mid H \textrm{ is an induced subgraph of } G\}|$ \;
+}
+\medskip
+    \nl $\FuncSty{FindInducedUniversalGraph'}(\AlgVar{maxIter})$  \;
+\nl \Begin{
+\nl    $G \gets (\{0, \dots, 13\}, \emptyset)$ \LeftComment{A graph on 14 vertices with no edges}\;
+\nl    Add edges to make vertices $\{0, \dots, 5\}$ of $G$ a clique \;
+\nl    Add to $G$ each possible edge marked $\boldsymbol{\cdot}$ in \Cref{fig:heuristic-regions} with probability $1/2$ \;
+    \nl    \For {$i \in \{1, \dots \AlgVar{maxIter}\}$\label{ImproveLoop}}{
+\nl      $v, w \gets$ vertices corresponding to a randomly-selected position marked $\boldsymbol{\cdot}$ in \Cref{fig:heuristic-regions} \;
+    \nl      $G' \gets \FuncSty{FlipEdge}(G, v, w)$ \label{FlipStep} \;
+    \nl      \lIf {$\FuncSty{Score}(G') = |\calF(6)|$}{$\KwSty{return}$~$G'$ \LeftComment{Success!}}
+    \nl      \lIf {$\FuncSty{Score}(G') \geq \FuncSty{Score}(G)$}{$G \gets G'$\label{ImproveLoopEnd}}
+       }
+\nl   $\KwSty{return}$~$\AlgVar{null}$ \;
+%%%\nl    \For {$G \in \calF(n)$}{
+%%%\nl      \lIf {$\FuncSty{IsInducedUniversal}(\calF,G)$}{$\calG \gets \calG \cup \{G\}$}
+%%%       }
+%%%\nl   $\KwSty{return}$~$\calG$ \;
+    }
+\medskip
+    \nl $\FuncSty{FindInducedUniversalGraph}(\AlgVar{maxIter})$  \;
+\nl \Begin{
+\nl    \While {time limit is not exceeded}{
+    \nl    $G \gets \FuncSty{FindInducedUniversalGraph'}(\AlgVar{maxIter})$ \;
+    \nl    \lIf {$G \not= \AlgVar{null}$}{$\KwSty{return}$~$G$}
+       }
+\nl   $\KwSty{return}$~$\AlgVar{null}$ \;
+%%%\nl    \For {$G \in \calF(n)$}{
+%%%\nl      \lIf {$\FuncSty{IsInducedUniversal}(\calF,G)$}{$\calG \gets \calG \cup \{G\}$}
+%%%       }
+%%%\nl   $\KwSty{return}$~$\calG$ \;
+    }
+\caption{A hill-climbing algorithm to find an induced universal graph for family $\calF(6)$}
+\label{alg:heuristic}
+\end{algorithm}
+
+The function $\FuncSty{FindInducedUniversalGraph'}()$
+begins by generating a random graph $G$ on 14 vertices as follows
+(see \Cref{fig:heuristic-regions}).
+Numbering the vertices from zero, we make the
 $k$ vertices numbered $0$ to $k-1$ a clique, and the $k$ vertices numbered $k-1$ to $2k-2$ an
-independent set.\footnote{The clique and the independent set thus have one vertex
+independent set.\footnote{The
+clique and the independent set thus have one vertex
 in common.  The proof of Proposition~\ref{f6proposition} can be modified straightforwardly
@@
 induced subgraphs with disjoint vertex sets.}
-Each possible edge that is not involved in either the clique
-or the independent set is added with probability $1/2$.
-We then repeatedly ``flip'' the status of a random edge from present to absent
-or vice versa, but always leave the large clique and independent set intact.
-After each flip, we count the number of 156 graphs on 6 vertices that are isomorphic
-to a subgraph of our 14-vertex graph.  If the most recent flip decreased
-this number, we revert it.  After each 1000 flips, we restart the algorithm
-with a new random graph.
+
+\begin{figure}[h!]
+    \centering
+    \footnotesize
+%    \tt
+       \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \tikzstyle{every node}=[]
+          \input{img/heuristic-regions}
+      \end{tikzpicture}
+    \caption{The adjacency matrix used in our heuristic algorithm when searching
+        for an induced universal graph of order 14 for $\calF(6)$.  The blue
+        region (a 6-vertex clique) and the yellow region (a 6-vertex independent
+        set) remain unchanged as the algorithm runs.  The remaining possible edges,
+        marked $\boldsymbol{\cdot}$, are initially set to 0 or 1 at random,
+        and are flipped between 0 and 1 as the algorithm progresses.}
+\label{fig:heuristic-regions}
+\end{figure}
+
+Each possible edge that is not involved in either the clique or the independent
+set --- shown in the figure with a dot --- is added with probability $1/2$.
+The loop in \linerangeref{ImproveLoop}{ImproveLoopEnd} then carries out up to
+$\AlgVar{maxIter}$ iterations of a step that randomly modifies graph $G$,
+and accepts the modification if it causes the number of graphs in $\calF(6)$ that
+are induced subgraphs of $G$ to increase.
+The attempted-improvement step chooses a random pair of vertices $\{v, w\}$ that
+are not involved in either the clique or the independent set
+highlighted in \Cref{fig:heuristic-regions}, and
+``flips'' the status of the edge between these two vertices: the edge is added if it
+not present, and removed otherwise.
+After each flip, we count the number of graphs on 6 vertices (from a total
+of 156 graphs) that are
+isomorphic to an induced subgraph of our 14-vertex graph.  If the most recent flip
+increase this number or left it unchanged, we accept it.  After $\AlgVar{maxIter}$
+flips, we restart the algorithm with a new random graph.
+
+Our implementation has two optimisations that are not shown in \Cref{alg:heuristic}.
+First, scores are cached in a dictionary data structure, so that we can quickly look
+up the score of a graph that has already been visited. This data structure is
+cleared on return from $\FuncSty{FindInducedUniversalGraph'}$.
+(An alternative approach, which we have not implemented, would be avoid
+revisiting graphs using a tabu list \cite{DBLP:books/daglib/0093574}.)
+
+The second optimisation often avoids the need to visit every graph in $\calF(6)$
+during calls to $\FuncSty{Score}()$.  Rather than beginning our count at 0
+and incrementing it for every graph in $\calF(6)$ that is an induced subgraph
+of $G$, we begin our count at $|\calF(6)|$ and decrement it for every
+graph that is \emph{not} an induced subgraph of $G$.  This allows us to return
+early from the function with a score of $-1$ as soon as the counter variable
+falls below the score of $G$ prior to the most recent edge-flip, since any
+score below that value results in the most recent change being rejected.
+
+By running this search algorithm, we quickly found the graph of order 14 whose adjacency
+matrix is shown in \Cref{fig:adjmat14}.  This is induced universal for $\calF(6)$;
+therefore, $f(6) \leq 14$.  Combining our two bounds, we have $f(6) = 14$.
 
@@
 \centering
-\small
+\footnotesize
 \verb|0 1 1 1 1 1 0 1 1 0 0 0 1 0| \\
@@
 
-\begin{figure}[htb]
-    \centering
-      \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
-        \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.1cm]
-      \end{tikzpicture}
-\label{fig:heuristic-regions}
-\end{figure}
-
 \section{Bounds on \texorpdfstring{$f(7)$}{f(7)}}\label{sec:f7}
@@
 the heuristic described in Section~\ref{sec:f6}, with two modifications.
-First, 10000 rather than 1000
+First, 10,000 rather than 1000
 edge-flips were permitted before each restart, as this was found to be more effective
@@
 \centering
-\small
+\footnotesize
 \verb|0 1 1 1 1 1 1 1 1 1 0 1 1 1 1 1 0 0| \\
@@
 
+In order to generate induced universal graphs for families of
+all trees on $k$ vertices, we ran \Cref{alg:brute-force}
+using lists of trees published by Brendan
+McKay.\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/trees.html}}
+
 Table~\ref{tab:treeresults} gives the order $t(k)$ of a minimal induced universal graph for
@@
 shows one of the 66 minimal induced universal graphs for the family of 6-vertex trees.
+
+\section{Verification}\label{sec:verification}
+
+To verify correctness of the results in
+\Cref{tab:graphresults} and \Cref{tab:treeresults}, we repeated every call to
+the McSplit subgraph isomorphism solver using the LAD algorithm \cite{DBLP:journals/ai/Solnon10}
+as implemented in the \texttt{igraph} library \cite{igraph}.
+
+We verified that the graphs shown as adjacency matrices in \Cref{fig:adjmat14}
+and \Cref{fig:adjmat18} are induced universal graphs for their corresponding families
+using a shell script that calls the Glasgow Subgraph Solver to check for the
+inclusion of every graph of $k$ vertices \cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/gg/McCreeshP020}.
+This script shares no code with the Python program that generated the graphs.
+
+TODO: verification of small instances from the tables using Glasgow Subgraph Solver
 
@@
     \centering
-\begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+\begin{tikzpicture}[>=latex',line join=bevel,scale=.35]
   \pgfsetlinewidth{.5bp}
@@
 
-\caption{A universal graph for the family of all
+\caption{An induced universal graph for the family of all
 trees with 6 vertices}

### 2021-10-19 automorphism-experiment/analyse-second-experiment.sh
Re-run second experiment and add parallel coordinate plots

@@ -5,7 +5,9 @@ set -eu -o pipefail
 #echo "Results for automorphism and degree pattern orders:"
 #cat results-of-second-experiment.txt
 #echo
-for type in graph sample-of-graphs; do
+for type in graph sample-of-graphs tree; do
+    echo "------------------------------"
+    echo "Type:" $type
     echo "Results for "automorphisms" pattern order:"
     cat results-of-second-experiment_automorphisms_$type.txt | cut -d' ' -f1 | \
         datamash --header-out min 1 max 1 mean 1 | column -t
@@ -23,12 +25,21 @@ for type in graph sample-of-graphs; do
         datamash --header-out min 1 max 1 mean 1 | column -t
     echo
     echo
-done
+done | tee second-experiment-summary.txt
 
 Rscript reshape.R
 
 gnuplot second-experiment-scatter.gnuplot
 latexmk -pdf second-experiment-plot
 
+# Make .gnuplot file for parallel coordinates plot
+cat gnuplot-parts/second-experiment-scatter-using-sample.gnuplot.START > second-experiment-scatter-using-sample.gnuplot
+for run_num in $(seq 1 $(tail -n1 results-of-second-experiment_sample-of-graphs.tsv | awk '{print $1}')); do
+    echo "\"< awk 'NR==1 || \$1==$run_num' results-of-second-experiment_sample-of-graphs.tsv\"" 'u 5:($3/1000) w l lc rgb "#aa1f77b4" lw .3 notitle, \' >> second-experiment-scatter-using-sample.gnuplot
+done
+cat gnuplot-parts/second-experiment-scatter-using-sample.gnuplot.END >> second-experiment-scatter-using-sample.gnuplot
 gnuplot second-experiment-scatter-using-sample.gnuplot
 latexmk -pdf second-experiment-plot-using-sample
+
+gnuplot second-experiment-scatter-using-trees.gnuplot
+latexmk -pdf second-experiment-plot-using-trees

### 2021-10-19 automorphism-experiment/reshape.R

@@ -1,6 +1,6 @@
 library(tidyverse)
 
-for (type in c("graph", "sample-of-graphs")) {
+for (type in c("graph", "sample-of-graphs", "tree")) {
     autom <- read.delim(paste0("results-of-second-experiment_automorphisms_", type, ".txt"), header=FALSE, col.names=c("calls", "X"), sep=" ")$calls
     degree <- read.delim(paste0("results-of-second-experiment_degree_", type, ".txt"), header=FALSE, col.names=c("calls", "X"), sep=" ")$calls
     almost_random <- read.delim(paste0("results-of-second-experiment_almost-random_", type, ".txt"), header=FALSE, col.names=c("calls", "X"), sep=" ")$calls
@@ -8,8 +8,11 @@ for (type in c("graph", "sample-of-graphs")) {
 
     d <- data.frame(autom, degree, almost_random, r)
 
+    d$run_num <- 1:nrow(d)
+
     e <- d %>%
-        pivot_longer(everything(), names_to = "method", values_to = "calls")
+        #pivot_longer(everything(), names_to = "method", values_to = "calls")
+        pivot_longer(c("autom", "degree", "almost_random", "r"), names_to = "method", values_to = "calls")
 
     e$methodnum <- match(e$method, c("autom", "degree", "almost_random", "r"))
     e$jittered_methodnum <- e$methodnum + runif(nrow(e), -.0, .0)

### 2021-10-19 automorphism-experiment/run-second-experiment.sh

@@ -12,19 +12,41 @@ cd ..
 for type in graph sample-of-graphs; do
     for i in $(seq 1 $iters); do
         echo i = $i
-        time pypy3 find_universal_graphs.py $type $p $t >/dev/null 2>> automorphism-experiment/results-of-second-experiment_automorphisms_$type.txt
+        time pypy3 find_universal_graphs.py --seed=$i $type $p $t >/dev/null 2>> automorphism-experiment/results-of-second-experiment_automorphisms_$type.txt
     done
     for i in $(seq 1 $iters); do
         echo i = $i
-        time pypy3 find_universal_graphs.py $type $p $t --degree-pattern-order >/dev/null 2>> automorphism-experiment/results-of-second-experiment_degree_$type.txt
+        time pypy3 find_universal_graphs.py --seed=$i $type $p $t --degree-pattern-order >/dev/null 2>> automorphism-experiment/results-of-second-experiment_degree_$type.txt
     done
     for i in $(seq 1 $iters); do
         echo i = $i
-        time pypy3 find_universal_graphs.py $type $p $t --random-pattern-order >/dev/null 2>> automorphism-experiment/results-of-second-experiment_random_$type.txt
+        time pypy3 find_universal_graphs.py --seed=$i $type $p $t --random-pattern-order >/dev/null 2>> automorphism-experiment/results-of-second-experiment_random_$type.txt
     done
     for i in $(seq 1 $iters); do
         echo i = $i
-        time pypy3 find_universal_graphs.py $type $p $t --almost-random-pattern-order >/dev/null 2>> automorphism-experiment/results-of-second-experiment_almost-random_$type.txt
+        time pypy3 find_universal_graphs.py --seed=$i $type $p $t --almost-random-pattern-order >/dev/null 2>> automorphism-experiment/results-of-second-experiment_almost-random_$type.txt
     done
+done
+)
+(
+cd ..
+type=tree
+p=6
+t=8
+for i in $(seq 1 $iters); do
+    echo i = $i
+    time pypy3 find_universal_graphs.py --seed=$i $type $p $t >/dev/null 2>> automorphism-experiment/results-of-second-experiment_automorphisms_$type.txt
+done
+for i in $(seq 1 $iters); do
+    echo i = $i
[…]

### 2021-10-19 find_universal_graphs_heur.py
Print less to stderr from heuristic programs

@@ -75,7 +75,8 @@ def start():
                 best = score
                 if best > overall_best:
                     overall_best = best
-            sys.stderr.write("{} {} {}  {}\n".format(score, best, overall_best, len(patterns)))
+                    sys.stderr.write("* {} {}\n".format(overall_best, len(patterns)))
+            #sys.stderr.write("{} {} {}  {}\n".format(score, best, overall_best, len(patterns)))
             if score == len(patterns):
                 for row in T._adj_mat:
                     print(" ".join(str(int(x)) for x in row))

### 2021-10-19 find_universal_graphs_heur_for_trees.py

@@ -73,8 +73,8 @@ def start():
             elif score > best:
                 best = score
                 if best > overall_best:
-                    sys.stderr.write("* {} {}\n".format(overall_best, len(patterns)))
                     overall_best = best
+                    sys.stderr.write("* {} {}\n".format(overall_best, len(patterns)))
             if score == len(patterns):
 #                print()
 #                for i in range(nT):

### 2021-10-19 run_all.sh
Run checks when running program

@@ -7,6 +7,6 @@ mkdir results
 
 cat exhaustive_tests.txt | while read type p t; do
     echo $type $p $t
-    pypy3 find_universal_graphs.py $type $p $t > results/$type-$p-$t.txt
+    pypy3 find_universal_graphs.py $type $p $t --check > results/$type-$p-$t.txt
     pypy3 graph6_to_dot.py results/$type-$p-$t.txt dot-graphs/$type-$p-$t-{}.dot
 done

### 2021-10-19 run_heur.sh
Add more heuristic runs

@@ -1,4 +1,9 @@
 #!/bin/bash
 
-time pypy3 find_universal_graphs_heur.py 6 14 1000 1
-time pypy3 find_universal_graphs_heur.py 7 18 10000 1
+time pypy3 find_universal_graphs_heur.py 6 14 1000 1 | tee heur-graphs/graph6-14.txt
+time pypy3 find_universal_graphs_heur.py 7 18 10000 1 | tee heur-graphs/graph7-18.txt
+
+time pypy3 find_universal_graphs_heur_for_trees.py 7 11 1000 1 | tee heur-graphs/tree7-11.txt
+time pypy3 find_universal_graphs_heur_for_trees.py 8 13 10000 1 | tee heur-graphs/tree9-13.txt
+time pypy3 find_universal_graphs_heur_for_trees.py 9 15 10000 1 | tee heur-graphs/tree11-15.txt
+time pypy3 find_universal_graphs_heur_for_trees.py 10 18 10000 1 | tee heur-graphs/tree12-18.txt

### 2021-10-19 run_all.sh
Use CPython

@@ -7,6 +7,6 @@ mkdir results
 
 cat exhaustive_tests.txt | while read type p t; do
     echo $type $p $t
-    pypy3 find_universal_graphs.py $type $p $t --check > results/$type-$p-$t.txt
-    pypy3 graph6_to_dot.py results/$type-$p-$t.txt dot-graphs/$type-$p-$t-{}.dot
+    python3 find_universal_graphs.py $type $p $t --check > results/$type-$p-$t.txt
+    python3 graph6_to_dot.py results/$type-$p-$t.txt dot-graphs/$type-$p-$t-{}.dot
 done

### 2021-10-19 run_heur.sh

@@ -1,9 +1,9 @@
 #!/bin/bash
 
-time pypy3 find_universal_graphs_heur.py 6 14 1000 1 | tee heur-graphs/graph6-14.txt
-time pypy3 find_universal_graphs_heur.py 7 18 10000 1 | tee heur-graphs/graph7-18.txt
+time python3 find_universal_graphs_heur.py 6 14 1000 1 | tee heur-graphs/graph6-14.txt
+time python3 find_universal_graphs_heur.py 7 18 10000 1 | tee heur-graphs/graph7-18.txt
 
-time pypy3 find_universal_graphs_heur_for_trees.py 7 11 1000 1 | tee heur-graphs/tree7-11.txt
-time pypy3 find_universal_graphs_heur_for_trees.py 8 13 10000 1 | tee heur-graphs/tree9-13.txt
-time pypy3 find_universal_graphs_heur_for_trees.py 9 15 10000 1 | tee heur-graphs/tree11-15.txt
-time pypy3 find_universal_graphs_heur_for_trees.py 10 18 10000 1 | tee heur-graphs/tree12-18.txt
+time python3 find_universal_graphs_heur_for_trees.py 7 11 1000 1 | tee heur-graphs/tree7-11.txt
+time python3 find_universal_graphs_heur_for_trees.py 8 13 10000 1 | tee heur-graphs/tree9-13.txt
+time python3 find_universal_graphs_heur_for_trees.py 9 15 10000 1 | tee heur-graphs/tree11-15.txt
+time python3 find_universal_graphs_heur_for_trees.py 10 18 10000 1 | tee heur-graphs/tree12-18.txt

### 2021-10-19 paper/paper.tex
More writing

@@
 that are induced universal for $f(k)$.
-To give an example, we have $f(3)=5$ and $F(3)=5$. All five of the minimal induced universal graphs
+To give an example, we have $f(3)=5$ and $F(3)=5$; all five of the minimal
+induced universal graphs
 for $\calF(3)$ are shown in \Cref{fig:univ3a}. \Cref{fig:univ3b} shows the four
@@
     }
-\caption{Induced universal graphs for the family of all graphs on 3 vertices}
-\label{fig:graphs}
+\caption{Minimal induced universal graphs for the family of all graphs on 3 vertices}
+\label{fig:univ3}
 \end{figure}
@@
 This paper uses a brute-force approach similar to that of Preen to find minimal
-induced universal graphs for families of small \emph{general} graphs.
-Section~\ref{sec:method} describes our computer search method, including an
-optimisation to reduce run times.  Sections~\ref{sec:results5} presents exact
-values of $f(k)$ for $k \leq 5$ along with
-the corresponding counts of minimal induced universal subgraphs.
+induced universal graphs for families of small graphs.  Our contributions
+are (1) an ordering strategy to reduce the number of subgraph isomorphism
+calls required by the brute force algorithm; (2) exact values for $f(k)$ $(4 \leq k \leq 6)$
+and $F(k)$ $(1 \leq k \leq 5)$; (3) an upper bound of 18 for $f(7)$; (4) a hill-climbing
+search algorithm to find small (but not necessarily optimal) induced universal graphs
+(5) exact results for families of small trees.
+
+The sequel is structured as follows.
+Section~\ref{sec:method} describes the basic brute force method.
+Section~\ref{sec:iteration-order} compares four methods for sorting
+the graphs in $\calF$, and compares their effect on the number of subgraph
+isomorphism calls made by the brute-force algorithm.
+Sections~\ref{sec:results5} presents exact
+values of $f(k)$ and $F(k)$ for $k \leq 5$.
 Section~\ref{sec:f6} gives the value of $f(6)$ and and
@@
 Given $\calF$ and $n$,
-we use the brute-force approach described by Preen \cite{preen_math_se} and shown
-in \Cref{alg:brute-force} to find the set of
+we use the brute-force approach
+shown in \Cref{alg:brute-force} --- which is essentially the same as the method
+described by Preen \cite{preen_math_se} ---
+to find the set of
 $n$-vertex graphs that are induced universal for the family $\calF$.
@@
 for $\calF$.  The function $\FuncSty{IsInducedUniversal}()$ tests
-whether a graph $G$ is induced universal for $\calF$ by checking that
-$H$ is isomorphic to a subgraph of $G$ for each $G \in \calF$.
+whether a graph $G$ is induced universal for $\calF$ by checking for
+each $H \in \calF$ that $H$ is isomorphic to a subgraph of $G$.
 
 For the collection $\calF(n)$ of candidate graphs on \lineref{TryEachG}, our
-implementation uses trees generated using Brendan McKay's \texttt{geng}
+implementation uses graphs generated using Brendan McKay's \texttt{geng}
 program \cite{DBLP:journals/jal/McKay98}
@@
 
-For the call $\FuncSty{InducedSubIso}()$ to an algorithm for the induced
-subgraph isomorphism decision problem on line \lineref{CallSubIso},
-our program uses a Python implementation of the McSplit
+Our implementation is written purely in Python and run using the CPython
+interpreter.
+On line \lineref{CallSubIso}, the function $\FuncSty{InducedSubIso}()$
+calls an algorithm for the induced
+subgraph isomorphism decision problem; for this,
+our program uses an implementation by the author of the McSplit
 algorithm \cite{DBLP:conf/ijcai/McCreeshPT17}.
@@
 modification: we backtrack when the calculated upper bound
-on the order of a common subgraph is less the order of graph $H$.
+on the order of a common subgraph is less the order of either
+of the two input graphs.
 This method for solving
@@
 While McSplit is suitable for our purposes in this paper,
-we do not claim that it is the the fastest induced subgraph isomorphism
+we do not claim that it is the fastest induced subgraph isomorphism
 solver for very small graphs; it would be useful future work to perform
@@
 \DontPrintSemicolon
-\nl $\FuncSty{IsInducedUniversal}(\calF,G)$ \label{BruteForceFun} \;
+\nl $\FuncSty{IsInducedUniversal}(\calF,G)$ \;
 \nl \KwData{A family of graphs $\calF$ and a graph $G$.}
@@
 The full set of experiments described in this paper, run sequentially,
-took less than an hour to complete on a laptop with an Intel Core i5-6200U CPU
+took less than a day to complete on a laptop with an Intel Core i5-6200U CPU
 and 8 GB RAM.\footnote{The code and
@@
 $\calF$ better than others?  For a graph $G$ that contains a copy of each
-element of $\calF$, the iteration order is of no importance; we much check
-every element of $\calF$ before returning $\AlgVar{true}$.  For a graph $G$ that
+element of $\calF$, the iteration order is of no importance;
+every element of $\calF$ must be checked before $\AlgVar{true}$ is returned.
+For a graph $G$ that
 does \emph{not} contain a copy of each element of $\calF$, however, we would like
@@
 subgraph isomorphism test are checked early; that way, we can quickly return
-$\AlgVar{false}$.
-
-We begin by considering the case where $\calF = \calF(k)$ for some $k$; that is,
+$\AlgVar{false}$ and avoid unnecessary calls to the subgraph isomorphism algorithm.
+
+We begin by considering the case where our family of graphs $\calF$
+is $\calF(k)$ for some $k$; that is,
 $\calF$ is the set of all graphs of order $k$.
@@
 This led me
-to consider two possible orderings in which $K_k$ and its complement $I_k$ come
+to consider two orderings in which $K_k$ and its complement $I_k$ come
 first.  The first of these approaches is to the sort $\calF(k)$ such that
@@
 On the horizontal axes we have each of our two measures: size of automorphism group
-and `extremeness' of edge count.  The first of these measures has a strong correlation
+in the first plot and `extremeness' of edge count in the second plot.
+As we can see from the first plot, the automorphism measure has a strong correlation
 with the number of graphs containing $G$ as an induced subgraph, suggesting that
-it is a good strategy to use when sorting the list of graphs.
+sorting by this measure could be a good strategy.
 
@@
 finding all induced universal graphs of order 9 for the family of
-all graphs of order 5.  We consider four strategies.
-The first two of these are the automorphism-count and `degree extremeness'
+all graphs of order 5.  We consider four strategies:
+
+The first two of these are the automorphism-count and `edge-count extremeness'
 measures described above.  The fourth strategy is a random order.
@@
 the success of the first two strategies was purely due to having
-$I_5$ and $K_5$ at the start of the list of graphs.
+$I_5$ and $K_5$ at the start of the list of graphs. \Cref{tab:ordering-strategies}
+summarises these ordering strategies.
+
+\begin{table}[h!]
+\centering
+ \begin{tabular}{p{0.25\linewidth} p{0.55\linewidth}}
+ \toprule
+    Name & Description of strategy \\ [0.5ex]
+ \midrule
+    Automorphisms & Sort by size of the automorphism group \\
+    Edges & Sort by number of edges \\
+    Almost-random & Shuffle the list of graphs in $\calF$, then move $K_k$ and $I_k$ to the start (in random order) \\
+    Random & Shuffle the list of graphs in $\calF$ \\
+ \bottomrule
+\end{tabular}
+\caption{Ordering strategies tested for the brute-force algorithm}
+\label{tab:ordering-strategies}
+\end{table}
+
 For each of these strategies, we ran the program 50 times.
@@
 The results are shown in \Cref{fig:second-experiment}.  Over 50 runs,
-the `automorphisms` strategy required a mean of 304,758 calls to the subgraph
-isomorphism solver, while the `degree' strategy required a very similar
-number of calls:  a mean of 304,701.  Surprisingly, the `almost random'
+the automorphisms strategy required a mean of 304,746 calls to the subgraph
+isomorphism solver, while the edges strategy required marginally fewer
+calls:  a mean of 304,701.  Surprisingly, the almost-random
 strategy had the lowest average number of calls of all the strategies,
-at 304,431.  The `random' strategy fared poorly, requiring 822,649
-calls on average and a minimum of 451,287 calls in the best of the 50 runs.
-The variance in number of calls was tiny for all but the random strategy,
+at 304,387.  The random strategy fared poorly, requiring over 800,000
+calls on average.
+The variance in number of calls was small for all but the random strategy,
 with between 304,000 and 305,000 calls being required for each run.
 
-\begin{figure}[htb]
+\begin{figure}[h!]
     \centering
@@
     \caption{Using each of our four order strategies, we ran
-        $\FuncSty{AllInducedUniversalGraphs}(\calF(5),9)$ 50 times. The figure
+        $\FuncSty{AllInducedUniversalGraphs}(\calF(5),9)$ 50 times. The plot
         shows the number calls to the subgraph isomorphism algorithm, in thousands.
-        Horizontal lines show mean number of calls.}
+        Horizontal lines show average (mean) number of calls.}
 \label{fig:second-experiment}
@@
 
-We note, first, that `degree' strategy appears to be at least as effective
-as the `automorphisms' strategy for this instance, despite the lower
+We note, first, that edges strategy appears to be at least as effective
+as the automorphisms strategy for this instance, despite the lower
 correlation observed in \Cref{fig:scatter}.  Perhaps most surprising is
-that the `almost random' strategy performed better than any of the other
+that the almost-random strategy performed better than either of these two
 strategies; our expectation had been that this simple strategy would
@@
 We conjecture that the almost-random strategy is effective because it
-quickly checks a diverse range of graphs of graphs.  Consider, for example,
+quickly checks a diverse range of graphs.  Consider, for example,
 \Cref{tab:graphs-by-autom-count}, which shows the graphs of order 5 listed
@@
 the table as induced subgraphs, then it may also contain the graphs in the
-second row. The `almost random' strategy is likely to choose graphs
+second row. The almost-random strategy is likely to choose graphs
 that are quite different from a clique and an independent set after the first
@@
 
-\subsection{Testing the strategies: $\calF \subset \calF{5}$}
-
-While the `almost random' strategy was very effective when searching
+\subsection{Testing the strategies: $\calF \subset \calF(5)$}
+
+While the almost-random strategy was very effective when searching
 for induced universal graphs of \emph{all} graphs of a given order,
-it seems likely that it will be less effective if $\calF$ is a strict
-subset of $\calF(k)$, and in particular if it does not include one
-or both of $I_k$ and $K_k$.  \Cref{fig:second-experiment-using-sample}
-examines this claim.  We generated 25 (TODO 50) subsets of $\calF(5)$,
+it is plausible that it will be less effective if $\calF$ is a strict
+subset of $\calF(k)$, and in particular if it includes
+neither $I_k$ nor $K_k$.  \Cref{fig:second-experiment-using-sample}
+examines this claim.  We generated 50 subsets of $\calF(5)$,
 with each subset containing exactly half of the elements of $\calF(5)$.
 For each of these, we ran $\FuncSty{AllInducedUniversalGraphs}(\calF,9)$
-using each of the four strategies.  TODO say what happened
+using each of the four strategies.  The `automorphisms' strategy
+was the best overall in this case, requiring on average $8\%$, $27\%$, and $52\%$
+fewer calls to the subgraph isomorphism function than the edges, almost-random,
+and random strategies respectively.
+
+To help us understand why the almost-random
+strategy performs poorly for these instances, we connect the four solver runs on
+each instance together on the plot with blue lines.  The almost-random
+strategy, on the instances where it performs poorly, requires
+exactly the same number of subgraph isomorphism calls as the random strategy.
+These are the instances in which $\calF$ contains neither $K_5$ nor $I_5$;
+thus, the almost-random and random strategies are equivalent.  Notably,
+for these instances it is also the case that the automorphisms strategy
+tends to outperform the edges strategy.
 
@@
         $\FuncSty{AllInducedUniversalGraphs}(\calF,9)$ 50 times, where
-        $\calF$ is a random subset of $\calF(5)$ containing 17 graphs. The figure
+        $\calF$ is a random subset of $\calF(5)$ containing 17 graphs. This
+        parallel coordinates plot
         shows the number calls to the subgraph isomorphism algorithm, in thousands.
-        Horizontal lines show mean number of calls.}
+        Horizontal black lines show average (mean) number of calls.  The four runs for
+        each instance are connected by blue lines.}
 \label{fig:second-experiment-using-sample}
 \end{figure}
+
+\subsection{Testing the strategies: a family of trees}
+
+As a final test case, we let $\calF$ be the family of all trees on 6 vertices,
+and searched for all graphs of order 8 that are induced universal for $\calF$.
+\Cref{fig:second-experiment-using-trees} shows the results. The edges, almost-random,
+and random strategies had identical results, since all members of $\calF$ have five
+edges and the family contains neither $K_6$ nor $I_6$.  The automorphisms strategy
+requires fewer subgraph isomorphism calls on average than the other three strategies,
+although on seven of the 50 instances it does require marginally more calls than
+the other strategies.
+
+\begin{figure}[htb]
+    \centering
+
+    \includegraphics*[width=0.7\textwidth]{img/second-experiment-plot-using-trees}
+
+    \caption{Using each of our four order strategies, we ran
+        $\FuncSty{AllInducedUniversalGraphs}(\calF,8)$ 50 times, where
+        $\calF$ is the set of all trees of order 6. The plot
+        shows the number calls to the subgraph isomorphism algorithm, in thousands.
+        Horizontal black lines show average (mean) number of calls.}
+\label{fig:second-experiment-using-trees}
+\end{figure}
+
+The `automorphisms' strategy performs consistently well across our three test
+cases; therefore, we use this strategy for the remainder of the paper.
 

### 2021-10-19 paper/paper.tex
Paper version 2.0

@@
 We write $F(k)$ to denote the number of non-isomorphic graphs of order $f(k)$
-that are induced universal for $f(k)$.
+that are induced universal for $\calF(k)$.
 To give an example, we have $f(3)=5$ and $F(3)=5$; all five of the minimal
@@
 are (1) an ordering strategy to reduce the number of subgraph isomorphism
-calls required by the brute force algorithm; (2) exact values for $f(k)$ $(4 \leq k \leq 6)$
+calls required by the brute force algorithm; (2) exact values for $f(k)$ $(5 \leq k \leq 6)$
 and $F(k)$ $(1 \leq k \leq 5)$; (3) an upper bound of 18 for $f(7)$; (4) a hill-climbing
@@
 
-Table~\ref{tab:graphresults} shows, for $0 \leq k \leq 5$, the value of $f(k)$.
-The table also shows $F(k)$, the number of minimal induced universal
-graphs for the family of all $k$-vertex graphs.  To my knowledge, the value of
+Recall that $f(k)$ is the order of a minimal induced universal graph for
+$\calF(k)$, and $F(k)$ is the number of non-isomorphic graphs of that order
+for $\calF(k)$ that are induced universal for $\calF(k)$.  This section
+presents values of $f(k)$ and $F(k)$ for $0 \leq k \leq 5$, computed
+using \Cref{alg:brute-force}. The results are shown in 
+Table~\ref{tab:graphresults}.  To my knowledge, the value of
 $f(5)$ and the values of $F(k)$ have not been published previously.
@@
 Figure~\ref{fig:graphs} shows examples of minimal induced universal graphs
-for the families of all graphs with three, four and five vertices
+for the families of all graphs with four and five vertices
 respectively.
@@
 
-\begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
-  \pgfsetlinewidth{.5bp}
-%%
-\pgfsetcolor{black}
-  % Edge: 3 -- 0
-  \draw [] (51.166bp,32.248bp) .. controls (44.492bp,40.742bp) and (36.044bp,51.496bp)  .. (29.344bp,60.024bp);
-  % Edge: 4 -- 0
-  \draw [] (71.737bp,82.653bp) .. controls (60.758bp,80.98bp) and (46.814bp,78.855bp)  .. (35.882bp,77.189bp);
-  % Edge: 4 -- 1
-  \draw [] (104.4bp,96.268bp) .. controls (113.37bp,102.97bp) and (124.85bp,111.55bp)  .. (133.86bp,118.28bp);
-  % Edge: 4 -- 3
-  \draw [] (82.931bp,68.405bp) .. controls (78.728bp,58.107bp) and (73.39bp,45.028bp)  .. (69.205bp,34.773bp);
-  % Node: 0
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (18.0bp,74.46bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 1
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (148.63bp,129.31bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 2
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (148.0bp,57.0bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 3
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (62.36bp,18.0bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 4
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (89.87bp,85.42bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-%
-\end{tikzpicture}
-\qquad
-\qquad
 \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
@@
 \end{tikzpicture}
-\caption{Minimal induced universal graphs for the families of all
-graphs with 3, 4, and 5 vertices}
+\caption{Examples of minimal induced universal graphs for $\calF(4)$ and $\calF(5)$}
 \label{fig:graphs}
@@
     on $k$ vertices, and that $G$ has no more than $2k + 1$ vertices.  Graph
-    $G$ must have $K_k$ and $I_k$ as induced sugraphs.  This clique and
+    $G$ must have $K_k$ and $I_k$ as induced sugraphs.  Clearly, this clique and
     independent set may overlap by no more than one vertex, so their union must
@@
 are induced subgraphs of $G$ to increase.
-The attempted-improvement step chooses a random pair of vertices $\{v, w\}$ that
-are not involved in either the clique or the independent set
-highlighted in \Cref{fig:heuristic-regions}, and
+The attempted-improvement step chooses a random pair of vertices $\{v, w\}$,
+both of which are not in the same coloured square of 
+in \Cref{fig:heuristic-regions}, and
 ``flips'' the status of the edge between these two vertices: the edge is added if it
@@
 cleared on return from $\FuncSty{FindInducedUniversalGraph'}$.
-(An alternative approach, which we have not implemented, would be avoid
-revisiting graphs using a tabu list \cite{DBLP:books/daglib/0093574}.)
+(An alternative approach, which we have not implemented, would be to
+use a tabu list \cite{DBLP:books/daglib/0093574} to avoid revisiting graphs
+that have been visited recently.)
 
@@
 and \Cref{fig:adjmat18} are induced universal graphs for their corresponding families
-using a shell script that calls the Glasgow Subgraph Solver to check for the
-inclusion of every graph of $k$ vertices \cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/gg/McCreeshP020}.
+using a shell script that calls the Glasgow Subgraph Solver \cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/gg/McCreeshP020} to check for the
+inclusion of every graph of $k$ vertices.
 This script shares no code with the Python program that generated the graphs.
 
-TODO: verification of small instances from the tables using Glasgow Subgraph Solver
+Finally, we verified all but the final row in each of \Cref{tab:graphresults}
+and \Cref{tab:treeresults} using a second shell script that calls the Glasgow
+Subgraph Solver.  Again, this shares no code with the program used to generate
+the results in the tables.
 
@@
     \centering
-\begin{tikzpicture}[>=latex',line join=bevel,scale=.35]
+\begin{tikzpicture}[>=latex',line join=bevel,scale=.28]
   \pgfsetlinewidth{.5bp}

### 2021-10-20 paper/paper.tex
Clarify table of strategies

@@
  \midrule
-    Automorphisms & Sort by size of the automorphism group \\
-    Edges & Sort by number of edges \\
-    Almost-random & Shuffle the list of graphs in $\calF$, then move $K_k$ and $I_k$ to the start (in random order) \\
+    Automorphisms & Sort in descending order of automorphism group size \\
+    Edges & Sort in descending order of $\big|{2|E(G)| - {|V(G)| \choose 2}}\big|$ 
+            (thus, give priority to graphs that are very sparse or very dense)\\
+    Almost-random & Shuffle the list of graphs in $\calF$, then move $K_k$ and $I_k$ to the start if they are present \\
     Random & Shuffle the list of graphs in $\calF$ \\

### 2021-10-22 find_universal_graphs_heur_for_trees.py
Embed star in graph in heuristic program for trees

@@ -23,10 +23,23 @@ def read_all_graphs(n):
 
     
 def random_graph(n, nP):
+    # This function is not used
     G = Graph(n)
     for v in range(n):
         for w in range(v):
             if random.random() < .1:
+                G.add_edge(v, w)
+    return G
+
+    
+def initial_graph(n, nP):
+    G = Graph(n)
+    for v in range(n):
+        for w in range(v):
+            if v < nP and w < nP:
+                if w == 0:
+                    G.add_edge(v, w)
+            elif random.random() < .1:
                 G.add_edge(v, w)
     return G
 
@@ -56,7 +69,7 @@ def start():
     overall_best = -1
     while True:
         score_cache = {}
-        T = random_graph(nT, nP)
+        T = initial_graph(nT, nP)
         best = -1
         for _ in range(iters_per_T):
             v, w = choose_edge_to_change(T, nP)
@@ -118,8 +131,11 @@ def choose_edge_to_change(G, nP):
     while True:
         v = random.randrange(G.number_of_nodes())
         w = random.randrange(G.number_of_nodes())
-        if v != w:
-            return v, w
+        if v == w:
+            continue
+        if v < nP and w < nP:
+            continue
+        return v, w
 
 
 if __name__ == "__main__":

### 2021-10-23 find_universal_graphs_backtracking_for_trees.py
Add exact backtracking program for families of trees

@@ -1,0 +1,91 @@
+import sys
+
+from universalgraphs import Graph, induced_subgraph_isomorphism
+
+def read_all_trees(n):
+    with open("graphs/tree{}.all.txt".format(n), "r") as f:
+        for line in f:
+            tokens = [int(tok) for tok in line.strip().split()]
+            edges = zip(tokens[::2], tokens[1::2])
+            G = Graph(n)
+            for v, w in edges:
+                G.add_edge(v, w)
+            yield G
+
+    
+def initial_graph(n, nP):
+    G = Graph(n)
+    for v in range(1, nP):
+        G.add_edge(0, v)
+    return G
+
+    
+def iso(P, T):
+    return induced_subgraph_isomorphism(P, T)
+
+
+def isomorphic(G, H):
+    if G[1] != H[1]:  # For speed, compare degree sequences first
+        return False
+    return iso(G[0], H[0])
+
+
+def try_adding_universal_graph(T, universal_graphs):
+    if not any(isomorphic(G, T) for G in universal_graphs):
+        universal_graphs.append(T)
+
+
+def deg_seq(G):
+    return sorted([sum(row) for row in G._adj_mat])
+
+
+def search(nT, nP, patterns, universal_graphs, vals):
+    if len(vals) <= 2:
+        print(vals)
+    n = len(vals)
+    if n == nT - 1:
+        T = initial_graph(nT, nP)
+        for v, ww in enumerate(vals):
+            if v == 0:
+                for i in range(ww):
+                    T.add_edge(v, nT - i - 1)
+            else:
+                for i in range(nT - nP):
+                    if (ww & (1 << i)):
+                        T.add_edge(v, nT - i - 1)
+        if all(iso(P, T) for P in patterns):
+            try_adding_universal_graph((T, deg_seq(T)), universal_graphs)
+        return
+    if n == 0:
+        top = nT - nP + 1
+    elif n == 1:
+        top = 1 << (nT - nP)
+    elif n < nP:
+        top = vals[-1] + 1
+    else:
+        top = 1 << (nT - n - 1)
+    for i in range(top):
+        vals.append(i)
+        search(nT, nP, patterns, universal_graphs, vals)
+        vals.pop()
+
+
+def start():
+    if len(sys.argv) < 3:
+        print("Usage: python3 {} NP NT".format(sys.argv[0]))
+        exit(1)
+
+    nP = int(sys.argv[1])
[…]

### 2021-10-25 find_universal_graphs_backtracking_for_trees.py
Use list of small canonical graphs in completion algorithm for trees

@@ -1,6 +1,13 @@
 import sys
 
-from universalgraphs import Graph, induced_subgraph_isomorphism
+from universalgraphs import Graph, induced_subgraph_isomorphism, from_graph6_bytes
+
+def read_all_graphs(n):
+    with open("graphs/graph{}.g6".format(n), "rb") as f:
+        for line in f:
+            g = line.strip()
+            yield from_graph6_bytes(g)
+
 
 def read_all_trees(n):
     with open("graphs/tree{}.all.txt".format(n), "r") as f:
@@ -39,34 +46,33 @@ def deg_seq(G):
     return sorted([sum(row) for row in G._adj_mat])
 
 
-def search(nT, nP, patterns, universal_graphs, vals):
+def search(nT, nP, patterns, small_graphs, universal_graphs, vals):
     if len(vals) <= 2:
         print(vals)
     n = len(vals)
-    if n == nT - 1:
+    if n == nP + 1:
         T = initial_graph(nT, nP)
-        for v, ww in enumerate(vals):
-            if v == 0:
-                for i in range(ww):
+        for v, ww in enumerate(vals[:-1]):
+            for i in range(nT - nP):
+                if (ww & (1 << i)):
                     T.add_edge(v, nT - i - 1)
-            else:
-                for i in range(nT - nP):
-                    if (ww & (1 << i)):
-                        T.add_edge(v, nT - i - 1)
+        small_graph = small_graphs[vals[-1]]
+        for v in range(nT - nP):
+            for w in range(v):
+                if small_graph._adj_mat[v][w]:
+                    T.add_edge(v + nP, w + nP)
         if all(iso(P, T) for P in patterns):
             try_adding_universal_graph((T, deg_seq(T)), universal_graphs)
         return
-    if n == 0:
-        top = nT - nP + 1
-    elif n == 1:
+    if n <= 1:
         top = 1 << (nT - nP)
     elif n < nP:
         top = vals[-1] + 1
     else:
-        top = 1 << (nT - n - 1)
+        top = len(small_graphs)
     for i in range(top):
         vals.append(i)
-        search(nT, nP, patterns, universal_graphs, vals)
+        search(nT, nP, patterns, small_graphs, universal_graphs, vals)
         vals.pop()
 
 
@@ -82,8 +88,10 @@ def start():
 
     patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
 
[…]

### 2021-10-25 find_universal_graphs_backtracking_for_trees-naive.py
Add naive completion program for trees

@@ -1,0 +1,83 @@
+import sys
+
+from universalgraphs import Graph, induced_subgraph_isomorphism
+
+def read_all_trees(n):
+    with open("graphs/tree{}.all.txt".format(n), "r") as f:
+        for line in f:
+            tokens = [int(tok) for tok in line.strip().split()]
+            edges = zip(tokens[::2], tokens[1::2])
+            G = Graph(n)
+            for v, w in edges:
+                G.add_edge(v, w)
+            yield G
+
+    
+def initial_graph(n, nP):
+    G = Graph(n)
+    for v in range(1, nP):
+        G.add_edge(0, v)
+    return G
+
+    
+def iso(P, T):
+    return induced_subgraph_isomorphism(P, T)
+
+
+def isomorphic(G, H):
+    if G[1] != H[1]:  # For speed, compare degree sequences first
+        return False
+    return iso(G[0], H[0])
+
+
+def try_adding_universal_graph(T, universal_graphs):
+    if not any(isomorphic(G, T) for G in universal_graphs):
+        universal_graphs.append(T)
+
+
+def deg_seq(G):
+    return sorted([sum(row) for row in G._adj_mat])
+
+
+def search(nT, nP, patterns, universal_graphs, vals):
+    if len(vals) <= 2:
+        print(vals)
+    n = len(vals)
+    if n == nP + 1:
+        T = initial_graph(nT, nP)
+        for v, ww in enumerate(vals):
+            for i in range(nT - nP):
+                if (ww & (1 << i)):
+                    T.add_edge(v, nT - i - 1)
+        if all(iso(P, T) for P in patterns):
+            try_adding_universal_graph((T, deg_seq(T)), universal_graphs)
+        return
+    if n < nP:
+        top = 1 << (nT - nP)
+    else:
+        top = 1 << (nT - n - 1)
+    for i in range(top):
+        vals.append(i)
+        search(nT, nP, patterns, universal_graphs, vals)
+        vals.pop()
+
+
+def start():
+    if len(sys.argv) < 3:
+        print("Usage: python3 {} NP NT".format(sys.argv[0]))
+        exit(1)
+
+    nP = int(sys.argv[1])
+    nT = int(sys.argv[2])
+
+    patterns = [G for G in read_all_trees(nP)]
+
+    patterns.sort(key=lambda G: -len(induced_subgraph_isomorphism(G, G, True)))
+
+    universal_graphs = []
[…]

### 2021-10-25 make-backtracking-tree-algorithm-fig.py
Add scripts to make two figures

@@ -1,0 +1,39 @@
+n = 11
+k = 7
+
+adjmat = [[0] * n for _ in range(n)]
+
+for i in range(1, k):
+    adjmat[i][0] = "\\textbf{1}"
+    adjmat[0][i] = "\\textbf{1}"
+
+for i in range(n):
+    for j in range(i):
+        if i == j:
+            continue
+        if i < k and j < k:
+            continue
+        if i >= k and j >= k:
+            continue
+        a = j + 1
+        b = i - k + 1
+        adjmat[i][j] = "$x_{{{}{}}}$".format(a, b)
+        adjmat[j][i] = "$x_{{{}{}}}$".format(a, b)
+
+for i in range(n):
+    for j in range(i):
+        if i == j:
+            continue
+        if i >= k and j >= k:
+            adjmat[i][j] = "$\\boldsymbol{\\cdot}$"
+            adjmat[j][i] = "$\\boldsymbol{\\cdot}$"
+
+print("\draw[uofgcobalt,fill=uofgcobalt,fill opacity=0.15, thick] (-.4,-6.4) rectangle (6.4,.4);")
+print("\draw[uofgpumpkin,fill=uofgpumpkin,fill opacity=0.15, thick] (6.6,-6.4) rectangle (10.4,-.4);")
+#print("\draw[uofgpumpkin,fill=uofgpumpkin,fill opacity=0.15, thick] (.4,-10.4) rectangle (6.4,-6.6);")
+print("\draw[uofgslate,fill=uofgslate,fill opacity=0.15, thick] (6.6,-10.4) rectangle (10.4,-6.6);")
+
+for i in range(n):
+    for j in range(n):
+        print("\\node ({}_{}) at ({},{}) {{{}}};".format(i, j, i, -j, adjmat[i][j]))
+

### 2021-10-25 make-heuristic-regions-fig-for-trees.py

@@ -1,0 +1,24 @@
+n = 9
+k = 6
+
+adjmat = [[0] * n for _ in range(n)]
+
+for i in range(1, k):
+    adjmat[i][0] = "\\textbf{1}"
+    adjmat[0][i] = "\\textbf{1}"
+
+for i in range(n):
+    for j in range(n):
+        if i == j:
+            continue
+        if i < k and j < k:
+            continue
+        adjmat[i][j] = "$\\boldsymbol{\\cdot}$"
+
+print("\draw[uofgcobalt,fill=uofgcobalt,fill opacity=0.15, thick] (-.5,-5.5) rectangle (5.5,.5);")
+#print("\draw[uofgpumpkin,fill=uofgpumpkin,fill opacity=0.15, thick] (4.5,-10.5) rectangle (10.5,-4.5);")
+
+for i in range(n):
+    for j in range(n):
+        print("\\node ({}_{}) at ({},{}) {{{}}};".format(i, j, i, -j, adjmat[i][j]))
+

### 2021-10-25 find_universal_graphs_backtracking_for_trees.py
Make completion code nicer

@@ -20,10 +20,18 @@ def read_all_trees(n):
             yield G
 
     
-def initial_graph(n, nP):
-    G = Graph(n)
+def make_target_graph(nT, nP, vals, small_graph):
+    G = Graph(nT)
     for v in range(1, nP):
         G.add_edge(0, v)
+    for v, ww in enumerate(vals):
+        for i in range(nT - nP):
+            if (ww & (1 << i)):
+                G.add_edge(v, nT - i - 1)
+    for v in range(nT - nP):
+        for w in range(v):
+            if small_graph._adj_mat[v][w]:
+                G.add_edge(v + nP, w + nP)
     return G
 
     
@@ -50,30 +58,17 @@ def search(nT, nP, patterns, small_graphs, universal_graphs, vals):
     if len(vals) <= 2:
         print(vals)
     n = len(vals)
-    if n == nP + 1:
-        T = initial_graph(nT, nP)
-        for v, ww in enumerate(vals[:-1]):
-            for i in range(nT - nP):
-                if (ww & (1 << i)):
-                    T.add_edge(v, nT - i - 1)
-        small_graph = small_graphs[vals[-1]]
-        for v in range(nT - nP):
-            for w in range(v):
-                if small_graph._adj_mat[v][w]:
-                    T.add_edge(v + nP, w + nP)
-        if all(iso(P, T) for P in patterns):
-            try_adding_universal_graph((T, deg_seq(T)), universal_graphs)
-        return
     if n <= 1:
-        top = 1 << (nT - nP)
+        for i in range(1 << (nT - nP)):
+            search(nT, nP, patterns, small_graphs, universal_graphs, vals + [i])
     elif n < nP:
-        top = vals[-1] + 1
+        for i in range(vals[-1] + 1):
+            search(nT, nP, patterns, small_graphs, universal_graphs, vals + [i])
     else:
-        top = len(small_graphs)
-    for i in range(top):
-        vals.append(i)
-        search(nT, nP, patterns, small_graphs, universal_graphs, vals)
-        vals.pop()
+        for small_graph in small_graphs:
+            T = make_target_graph(nT, nP, vals, small_graph)
+            if all(iso(P, T) for P in patterns):
+                try_adding_universal_graph((T, deg_seq(T)), universal_graphs)
 
 
 def start():

### 2021-10-25 paper/paper.tex
Tidy
More about trees

@@
 search algorithm to find small (but not necessarily optimal) induced universal graphs
-(5) exact results for families of small trees.
+(5) exact values of the order of a minimal induced universal graph
+for the families of all trees on $k$ vertices, for $1 \leq k \leq 8$.
 
@@
 bound is proven and a graph obtained by heuristic search demonstrating the
-upper bound is given.  Finally, Section~\ref{sec:trees} gives results for
-families of trees.
+upper bound is given.  Finally, Section~\ref{sec:trees} describes a specialised
+algorithm for families of trees, and gives results for
+these families.
 
@@
 $n$-vertex graphs that are induced universal for the family $\calF$.
-The function $\FuncSty{AllInducedUniversalGraphs}()$ is the entry point.
+The entry point is the function $\FuncSty{AllInducedUniversalGraphs}()$.
 \Lineref{TryEachG} tests in turn each candidate graph $G$ from the family
@@
 \centering
-\footnotesize
+\scriptsize
 \verb|0 1 1 1 1 1 0 1 1 0 0 0 1 0| \\
@@
 \centering
-\footnotesize
+\scriptsize
 \verb|0 1 1 1 1 1 1 1 1 1 0 1 1 1 1 1 0 0| \\
@@
 
-In order to generate induced universal graphs for families of
-all trees on $k$ vertices, we ran \Cref{alg:brute-force}
-using lists of trees published by Brendan
+\begin{figure}[h!]
+    \centering
+    \footnotesize
+%    \tt
+       \begin{tikzpicture}[>=latex',line join=bevel,scale=.6]
+        \tikzstyle{every node}=[]
+          \input{img/regions-for-trees}
+      \end{tikzpicture}
+    \caption{The adjacency-matrix regions used in the 
+        ``symmetry breaking completion'' method. For this example, $n=11$ and $k=7$.}
+\label{fig:regions-for-trees}
+\end{figure}
+
+To generate induced universal graphs for families of
+all trees on $k$ vertices, we tried two approaches.
+The first approach was to run \Cref{alg:brute-force}, with the family
+$\calF$ of order-$k$ trees loaded from
+a list of trees published by Brendan
 McKay.\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/trees.html}}
+The second approach, described in the following subsection, generates
+only candidate graphs that contain a star --- one of the elements
+of $\calF$ ---as an induced subgraph.  The final part of this section
+gives results generated using these two methods, along with additional
+upper bounds generated using a specialised version of our hill-climbing
+algorithm.
+
+\subsection{The completion method}
+
+Our second approach was to embed the star $S_{k-1}$ --- which must be
+present since it is a tree on $k$ vertices --- in the top left position
+of the adjacency matrix, then systematically try all possible ways to complete
+the adjacency matrix.  For each completed adjacency matrix, we test whether
+the graph contains as an induced subgraph each tree in our family.
+If it does, and if the graph is not isomorphic to any induced universal
+graph that has already been found, we add it to our collection of induced
+universal graphs.
+We will refer to this as the ``naive completion'' method.
+
+Because the naive completion method has to test a huge number of graphs,
+many pairs of which are isomorphic, we created an improved version of the
+algorithm, which we will refer to as the ``symmetry breaking completion''
+method.  This adds two symmetry-breaking constraints on the adjacency matrix.
+The first of these constraints concerns the grey shaded region at the bottom-right
+of the adjacency matrix, which has $n - k$ rows and columns.  Rather than
+trying all $2^{n-k \choose 2}$ possible entries for this region of the
+adjacency matrix, we use a list containing a canonical adjacency matrix
+for each possible graph of order $n-k$, and we require that the shaded
+region be equal to one of these adjacency matrices.
+
+The second symmetry-breaking constraint concerns the yellow shaded region
+in \Cref{fig:regions-for-trees}. Each row of this region has $n-k$ entries,
+named $x_{i1}, \dots, x_{i,n-k}$.  For our symmetry-breaking constraint, we
+view each row as the binary number $x_{i1}, \dots, x_{i,n-k}$
+(with $x_{i,n-k}$ as the least significant digit), and we insist
+that these numbers are in nondecreasing order; thus
+$x_{21}, \dots, x_{2,n-k} \leq \dots \leq x_{k1}, \dots, x_{k,n-k}$.
+In other words, we impose a lexicographic ordering constraint
+on the rows of this region.
+
+The algorithm with these symmetry-breaking constraints is shown
+in \Cref{alg:completion}.  The call to $\FuncSty{MakeGraph}()$ on
+\lineref{MakeGraphLine} constructs a graph by building its
+adjacency matrix as shown in \Cref{fig:regions-for-trees}.  The
+first $k$ vertices are connected in the form of a star;
+for each $i$ such that $1 \leq i \leq j$, the digits of the binary
+representation of $x_i$
+are assigned to $x_{i1}, \dots, x_{i,n-k}$, and these values
+are placed in the adjacency matrix as shown in the figure.  Finally,
+the small adjacency matrix $M$ is copied to the bottom-right of
+the adjacency matrix of the constructed graph.
+
+For the collection of graphs $\calG$ defined on \lineref{DefineSetG},
+isomorphic graphs are considered identical.  Thus, we only add a graph
+to $\calG$ if it is not isomorphic to any existing member of the
+collection.\footnote{An efficient method to ensure that isomorphic
+graphs are not stored in $\calG$ would be label the vertices
+of a graph $G$ canonically before adding it to the collection
+\cite{DBLP:journals/jsc/McKayP14}, and to store the collection as a
+hash set.  Because this part of the prograph is not a performance
+bottleneck, and to miminise software dependencies, our implementation
+uses the simpler approach of testing whether $G$ is isomorphic
+to each existing member of $\calG$, in turn.}
+
+\begin{algorithm}[h!]
+\DontPrintSemicolon
+\nl $\FuncSty{Search}(n,k,(x_1, \dots, x_j))$ \;
+\nl \KwData{Natural numbers $n$ and $k$, and a sequence
+    $(x_1, \dots, x_j))$ of nonnegative integers such that the binary
+    representation of each $x_i$ represents entries $x_{i1}, \dots, x_{i,n-k}$
+    of an adjacency matrix, as shown in \Cref{fig:regions-for-trees}}
+\nl \KwResult{The set of all order-$n$ graphs that are induced universal
+    for the family of trees of order $k$, and have an adjacency matrix
+    given by $\FuncSty{MakeGraph}(n,k,X,M)$ for some completion
+    $X$ of $(x_1, \dots, x_j)$ and some $M \in \mathcal{M}(n-k)$}
+\nl \Begin{
+\nl    $\calG \gets \emptyset$ \label{DefineSetG} \;
+\nl    \If {$j \leq 1$}{
+\nl      \For {$i \in \{0, \dots, 2^{n-k} - 1\}$}{
+\nl        $\calG \gets \calG \cup \FuncSty{Search}(n,p,(x_1, \dots, x_j, i))$
+  }
+}
+\nl    \ElseIf {$j \leq k$}{
+\nl      \For {$i \in \{0, \dots, x_j\}$}{
+\nl        $\calG \gets \calG \cup \FuncSty{Search}(n,p,(x_1, \dots, x_j, i))$
+  }
+}
+\nl    \Else {
+\nl      \For {$M \in \mathcal{M}(n-k)$}{
+\nl        $G \gets \FuncSty{MakeGraph}(n, k, (x_1, \dots, x_j), M)$ \label{MakeGraphLine}\;
+\nl        \If {$G$ contains every tree of order $k$ as an induced subgraph}{
+\nl          $\calG \gets \calG \cup \{G\}$ \;
+  }
+}
+}
+\nl    $\KwSty{return}$~$\calG$ \;
+    }
+\medskip
+\nl $\FuncSty{Start}(n,k)$ \;
+\nl \KwData{Natural numbers $n$ and $k$ such that $k<n$}
+\nl \KwResult{The set of all graphs of order $n$ that are induced universal for
+    the family of trees of order $k$.}
+\nl \Begin{
+\nl    $\KwSty{return}$~$\FuncSty{Search}(n,k,())$ \;
+    }
+\caption{The symmetry-breaking completion algorithm for finding all graphs of order $n$
+    that are induced universal for the family of all order-$k$ trees}
+\label{alg:completion}
+\end{algorithm}
+
+Because of the following proposition, the symmetry breaking completion
+method finds, up to isomorphism, all induced universal graphs for the family
+of trees on $k$ vertices.
+
+\begin{proposition}
+    Let $n, k$ be given, and let $G$ be a graph on $n$ vertices that is
+    induced universal for the family of trees of order $k$.  Let $\mathcal{M}$
+    be a set of canonical adjacency matrices for $\calF(n-k)$.  There is
+    a graph on $n$ vertices numbered $1, \dots, n$ that is isomorphic
+    to $G$ and satisfies the two symmetry-breaking constraints described
+    above.
+\end{proposition}
+\begin{proof}
+    Since $G$ is induced universal for the family of trees of order $k$, it
+    contains the star $S_{k-1}$ as an induced subgraph.  Arbitrarily
+    choose one such star in $G$ and give the label 1 to its vertex of degree
+    $k-1$.  Its remaining vertices will be given the labels $\{2, \dots, k\}$,
+    in an order we will specify later.
+    The remaining $n-k$ vertices of $G$ induce a subgraph isomorphic
+    to one of the adjacency matrices in $\mathcal{M}$; call this matrix $M$.
+    By choosing an appropriate relabelling of the $n-k$ vertices of $G$,
+    the region of the adjacency matrix corresponding to these vertices
+    can be made equal to $M$.  Thus, the first symmetry-breaking constraint
+    is satisfied.
+
+    The second symmetry-breaking constraint can now be satisfied by assigning
+    labels $\{2, \dots, k\}$ to the degree-1 vertices of the star $S_{k-1}$
+    in such an order that their adjacencies to the vertices outside the star,
+    viewed as binary numbers, are in lexicographic order.
+\end{proof}
+
+\subsection{Results for families of trees}
+
+The symmetry-breaking completion method is much faster than \Cref{alg:brute-force}.
+For $n=9$ and $k=6$, for example, the completion method takes 9 seconds whereas
+the brute-force algorithm takes 46 seconds. The difference in run-times becomes
+even greater for larger values of $n$ and $k$.
+Using the symmetry-breaking completion method,
+we were able to find the values of $t(k)$ and $T(k)$ for $k \leq 7$. We were also
+able to show that $t(8) > 12$.  We confirmed
+the results for $k \leq 6$ using \Cref{alg:brute-force}.
 
@@
 shows one of the 66 minimal induced universal graphs for the family of 6-vertex trees.
+
+To obtain upper bounds on $t(k)$ for larger values of $k$, we use a version of
+\Cref{alg:heuristic} with minor modifications.  Instead of initially embedding
+a clique and an independent set as in \Cref{fig:heuristic-regions}, we embed
+a single tree --- the $k$-vertex star $S_{k-1}$ --- as shown in
+\Cref{fig:heuristic-regions-for-trees}.  Each possible edge marked with a dot
+in the figure is added with probability 0.1.
+
+\begin{figure}[h!]
+    \centering
+    \footnotesize
+%    \tt
+       \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \tikzstyle{every node}=[]
+          \input{img/heuristic-regions-for-trees}
+      \end{tikzpicture}
+    \caption{The adjacency matrix used in our heuristic algorithm when searching
+        for an induced universal graph of order 9 for the set of all trees
+        of order 6.  The blue
+        region, a star with 6 vertices, remains unchanged as the algorithm runs.
+        The remaining possible edges, marked $\boldsymbol{\cdot}$,
+        are flipped between 0 and 1 as the algorithm progresses.}
+\label{fig:heuristic-regions-for-trees}
+\end{figure}
+
+Using this heuristic algorithm, we were able to find induced universal graphs
+to demonstrate that $t(8) \leq 13$, $t(9) \leq 15$ and $t(10) \leq 18$.
+Since we have matching lower and upper bounds for $t(8)$, we can state
+that $t(8) = 13$.
 
@@
  6 & 9 & 66 \\
+ 7 & 11 & 687 \\
  \bottomrule

### 2021-10-25 find_universal_graphs_heur_for_trees-with-viz.py
Add better tree example figure

@@ -1,0 +1,180 @@
+import sys
+import random
+
+from universalgraphs import Graph, induced_subgraph_isomorphism, from_graph6_bytes
+
+counter = [0]
+
+tikz_template = r"""{{\begin{{tikzpicture}}[>=latex',line join=bevel]
+  \tikzstyle{{every node}}=[draw, circle, inner sep=1pt, minimum size=.1cm]
+  \pgfsetlinewidth{{.5bp}}
+  {}
+  {}
+\end{{tikzpicture}}}}"""
+
+tikz_node_template = "\\node[{}] at ({},{}) ({}) {{}};\n"
+
+tikz_edge_template = "\\draw[{}] ({}) -- ({});\n"
+
+highlight_colour = "uofgheather"
+
+def read_all_trees(n):
+    with open("graphs/tree{}.all.txt".format(n), "r") as f:
+        for line in f:
+            tokens = [int(tok) for tok in line.strip().split()]
+            edges = zip(tokens[::2], tokens[1::2])
+            G = Graph(n)
+            for v, w in edges:
+                G.add_edge(v, w)
+            yield G
+
+def read_all_graphs(n):
+    with open("graphs/graph{}.g6".format(n), "rb") as f:
+        for line in f:
+            g = line.strip()
+            yield (from_graph6_bytes(g), g)
+
+    
+def random_graph(n, nP):
+    # This function is not used
+    G = Graph(n)
+    for v in range(n):
+        for w in range(v):
+            if random.random() < .1:
+                G.add_edge(v, w)
+    return G
+
+    
+def initial_graph(n, nP):
+    G = Graph(n)
+    for v in range(n):
+        for w in range(v):
+            if v < nP and w < nP:
+                if w == 0:
+                    G.add_edge(v, w)
+            elif random.random() < .1:
+                G.add_edge(v, w)
+    return G
+
+    
+def iso(P, T):
+    counter[0] += 1
+    return induced_subgraph_isomorphism(P, T)
+
+
+def start():
+    if len(sys.argv) < 6:
+        print("Usage: python3 {} NP NT iters_per_T seed step".format(sys.argv[0]))
+        exit(1)
+
+    nP = int(sys.argv[1])
+    nT = int(sys.argv[2])
+    iters_per_T = int(sys.argv[3])
+    random.seed(int(sys.argv[4]))
+    step_num = int(sys.argv[5])
+
+    patterns = [G for G in read_all_trees(nP)]
+
+#    for P in patterns:
[…]

### 2021-10-25 make-tree-viz.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+
+set -eu -o pipefail
+
+python3 find_universal_graphs_heur_for_trees-with-viz.py 7 11 10000 1 1
+dot tmp/tree.dot | grep -v '\[' | grep pos | sed 's/[^"]*"//' | sed 's/".*//' > tmp/tree.txt
+python3 find_universal_graphs_heur_for_trees-with-viz.py 7 11 10000 1 2 > paper/img/tree-example.tex

### 2021-10-25 paper/paper.tex
Edit paper

@@
 \begin{abstract}
+We present exact and heuristic algorithms that find, for a given family
+of graphs, a graph that contains each member of the family as an induced
+subgraph.
 For $0 \leq k \leq 6$, we give the minimum number of vertices $f(k)$ in a graph containing all
@@
 whether a graph $G$ is induced universal for $\calF$ by checking for
-each $H \in \calF$ that $H$ is isomorphic to a subgraph of $G$.
+each $H \in \calF$ that $H$ is isomorphic to an induced subgraph of $G$.
 
@@
 implementation uses graphs generated using Brendan McKay's \texttt{geng}
-program \cite{DBLP:journals/jal/McKay98}
-\footnote{The program \texttt{geng} is distributed
-with Nauty; we used version 27.r3.\url{https://pallini.di.uniroma1.it/}}.
+program \cite{DBLP:journals/jal/McKay98}.\footnote{The program \texttt{geng} is distributed
+with Nauty; we used version 27.r3.\url{https://pallini.di.uniroma1.it/}}
 These are read in \texttt{graph6}
@@
 interpreter.
-On line \lineref{CallSubIso}, the function $\FuncSty{InducedSubIso}()$
+On \lineref{CallSubIso}, the function $\FuncSty{InducedSubIso}()$
 calls an algorithm for the induced
@@
 modification: we backtrack when the calculated upper bound
-on the order of a common subgraph is less the order of either
+on the order of a common subgraph is less the order of the smaller
 of the two input graphs.
@@
 \nl \KwData{A family of graphs $\calF$ and a graph $G$.}
-\nl \KwResult{A logical value indicating whether $G$ is induced universal for $\calF$.}
+\nl \KwResult{A boolean value indicating whether $G$ is induced universal for $\calF$.}
 \nl \Begin{
@@
 
-When deciding on \lineref{HLoop} of the algorithm whether a graph $G$ contains
+When determining on \lineref{HLoop} of \Cref{alg:brute-force} whether a graph $G$ contains
 every element of $\calF$ as an induced subgraph, are some iteration orders of
@@
 to consider two orderings in which $K_k$ and its complement $I_k$ come
-first.  The first of these approaches is to the sort $\calF(k)$ such that
-graphs with large automorphism groups appear first.  (Intuitively, if a graph
+first.  The first of these approaches is to sort $\calF(k)$ in descending
+order of automorphism count, so that
+graphs with large automorphism groups appear first in the list.  (Intuitively, if a graph
 has few automorphisms then we can generate many different labelled graphs
@@
 with the number of graphs containing $G$ as an induced subgraph, suggesting that
-sorting by this measure could be a good strategy.
+sorting by this measure could be a good strategy for reducing the number of
+subgraph isomorphism calls in \Cref{alg:brute-force}.
 
@@
 finding all induced universal graphs of order 9 for the family of
-all graphs of order 5.  We consider four strategies:
+all graphs of order 5.  We consider four strategies.
 
@@
 measures described above.  The fourth strategy is a random order.
-Finally, we used one further strategy, which we call ``almost random''.
+Finally, we used one additional strategy that we call ``almost random''.
 Under this strategy, $I_5$ and $K_5$ are placed at the start of the list,
@@
 two rows, for example, have a maximum common induced subgraph of size 4.
-It seems likely that if a graph contains both graphs in the first row of
+It seems likely that if a graph contains both of the graphs in the first row of
 the table as induced subgraphs, then it may also contain the graphs in the
@@
 neither $I_k$ nor $K_k$.  \Cref{fig:second-experiment-using-sample}
-examines this claim.  We generated 50 subsets of $\calF(5)$,
+examines this claim.  We generated 50 subsets $\calF$ of $\calF(5)$ by random selection,
 with each subset containing exactly half of the elements of $\calF(5)$.
@@
 
-Recall that $f(k)$ is the order of a minimal induced universal graph for
+Recall that $\calF(k)$ is the family of all graphs of order $k$, $f(k)$ is
+the order of a minimal induced universal graph for
 $\calF(k)$, and $F(k)$ is the number of non-isomorphic graphs of that order
-for $\calF(k)$ that are induced universal for $\calF(k)$.  This section
+that are induced universal for $\calF(k)$.  This section
 presents values of $f(k)$ and $F(k)$ for $0 \leq k \leq 5$, computed
@@
 This section shows that $f(6) = 14$.  We begin with the lower bound.
-For $k \geq 6$, we can increase by 2 the lower bound by ``bof''.  In particular,
-this means that $f(6) \geq 14$.
+For $k \geq 6$, we can increase by 2 the lower bound --- $f(k) \geq 2k$
+for $k \geq 4$ --- cited in the previous section.
 
@@
     Suppose that $G$ is an induced universal graph for the family of all graphs
-    on $k$ vertices, and that $G$ has no more than $2k + 1$ vertices.  Graph
-    $G$ must have $K_k$ and $I_k$ as induced sugraphs.  Clearly, this clique and
+    on $k$ vertices, and that $G$ has no more than $2k + 1$ vertices.  Graphs
+    $K_k$ and $I_k$ are both members of $\calF(k)$, and therefore must
+    be induced sugraphs of $G$.  Clearly, this clique and
     independent set may overlap by no more than one vertex, so their union must
@@
     \centering
-
-% \begin{tikzpicture}[>=latex',line join=bevel]
-%   \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.5cm]
-%   \pgfsetlinewidth{.5bp}
-%   \node at (0,2.9) (1) {};
-%   \node at (0,2) (2) {};
-%   \node at (0,1.1) (3) {};
-%   \node at (1.1,2.9) (4) {};
-%   \node at (1.1,2) (5) {};
-%   \node at (1.1,1.1) (6) {};
-%   \draw (1) -- (2) -- (3);
-%   \draw (3) to [out=140,in=220] (1);
-%   \draw (4) -- (5) -- (6);
-%   \draw (6) to [out=40,in=-40] (4);
-%   \draw (1) -- (4);
-%   \draw (1) -- (5);
-%   \draw (1) -- (6);
-%   \draw (2) -- (4);
-%   \draw (2) -- (5);
-%   \draw (2) -- (6);
-%   \draw (3) -- (4);
-%   \draw (3) -- (5);
-%   \draw (3) -- (6);
-% \end{tikzpicture}
-% \qquad \qquad
-% \begin{tikzpicture}
-%   \tikzstyle{every node}=[draw, circle, inner sep=1pt, minimum size=.5cm]
-%   \pgfsetlinewidth{.5bp}
-%   \node at (0,2.9) (1) {};
-%   \node at (0,2) (2) {};
-%   \node at (0,1.1) (3) {};
-%   \node at (1.1,2.9) (4) {};
-%   \node at (1.1,2) (5) {};
-%   \node at (1.1,1.1) (6) {};
-% \end{tikzpicture}
-% \qquad \qquad
 \begin{tikzpicture}[>=latex',line join=bevel]
@@
 clique $K_k$, the independent set $I_k$, and
-the two graphs shown. See the the proof of Proposition~\ref{f6proposition}}.
+the two graphs shown. See the the proof of Proposition~\ref{f6proposition}.}
 \label{fig:boundproof}
@@
 \end{proof}
+
+\Cref{f6proposition} implies that $f(6) \geq 14$.
 
@@
 The attempted-improvement step chooses a random pair of vertices $\{v, w\}$,
-both of which are not in the same coloured square of 
+both of which are not in the same coloured square
 in \Cref{fig:heuristic-regions}, and
 ``flips'' the status of the edge between these two vertices: the edge is added if it
-not present, and removed otherwise.
+is not present, and removed otherwise.
 After each flip, we count the number of graphs on 6 vertices (from a total
@@
 isomorphic to an induced subgraph of our 14-vertex graph.  If the most recent flip
-increase this number or left it unchanged, we accept it.  After $\AlgVar{maxIter}$
-flips, we restart the algorithm with a new random graph.
+increased this number or left it unchanged, we accept the modification
+to the graph.  After $\AlgVar{maxIter}$
+flips, we restart the algorithm with a new randomly-generated graph.
 
@@
 up the score of a graph that has already been visited. This data structure is
-cleared on return from $\FuncSty{FindInducedUniversalGraph'}$.
-(An alternative approach, which we have not implemented, would be to
+cleared on return from $\FuncSty{FindInducedUniversalGraph'}$, in order to
+bound the program's memory use.
+(An alternative approach, which we have not implemented but which
+may be useful in order to explore more of the space of possible graphs, would be to
 use a tabu list \cite{DBLP:books/daglib/0093574} to avoid revisiting graphs
@@
 of $G$, we begin our count at $|\calF(6)|$ and decrement it for every
-graph that is \emph{not} an induced subgraph of $G$.  This allows us to return
+graph that is \emph{not} an induced subgraph of $G$.  We can then return
 early from the function with a score of $-1$ as soon as the counter variable
-falls below the score of $G$ prior to the most recent edge-flip, since any
-score below that value results in the most recent change being rejected.
+falls below the score obtained by the graph prior to the most recent edge-flip.
+This does not affect the behaviour of the hill-climbing search, since any
+score below the previous graph's score results in the most recent change being rejected.
 
@@
 
-\begin{figure}[h!]
-    \centering
-    \footnotesize
-%    \tt
-       \begin{tikzpicture}[>=latex',line join=bevel,scale=.6]
-        \tikzstyle{every node}=[]
-          \input{img/regions-for-trees}
-      \end{tikzpicture}
-    \caption{The adjacency-matrix regions used in the 
-        ``symmetry breaking completion'' method. For this example, $n=11$ and $k=7$.}
-\label{fig:regions-for-trees}
-\end{figure}
-
 To generate induced universal graphs for families of
@@
 $\calF$ of order-$k$ trees loaded from
-a list of trees published by Brendan
+a list published by Brendan
 McKay.\footnote{\url{https://users.cecs.anu.edu.au/~bdm/data/trees.html}}
@@
 
-Our second approach was to embed the star $S_{k-1}$ --- which must be
+In our second approach to finding induced universal graphs for the
+family of all trees on $k$ vertices, we embed the star $S_{k-1}$ --- which must be
 present since it is a tree on $k$ vertices --- in the top left position
@@
 If it does, and if the graph is not isomorphic to any induced universal
-graph that has already been found, we add it to our collection of induced
-universal graphs.
+graph that has already been found, we add it to our collection of graphs.
 We will refer to this as the ``naive completion'' method.
@@
 algorithm, which we will refer to as the ``symmetry breaking completion''
-method.  This adds two symmetry-breaking constraints on the adjacency matrix.
+method.  Our description refers to the adjacence matrix in
+\Cref{fig:regions-for-trees}.
+This adds two symmetry-breaking constraints on the adjacency matrix.
 The first of these constraints concerns the grey shaded region at the bottom-right
@@
 region be equal to one of these adjacency matrices.
+
+\begin{figure}[h!]
+    \centering
+    \footnotesize
+%    \tt
+       \begin{tikzpicture}[>=latex',line join=bevel,scale=.6]
+        \tikzstyle{every node}=[]
+          \input{img/regions-for-trees}
+      \end{tikzpicture}
+    \caption{The adjacency-matrix regions used in the 
+        ``symmetry breaking completion'' method. For this example, $n=11$ and $k=7$.}
+\label{fig:regions-for-trees}
+\end{figure}
 
@@
 \cite{DBLP:journals/jsc/McKayP14}, and to store the collection as a
-hash set.  Because this part of the prograph is not a performance
+hash set.  Because this part of the program is not a performance
 bottleneck, and to miminise software dependencies, our implementation
@@
 \medskip
-\nl $\FuncSty{Start}(n,k)$ \;
+\nl $\FuncSty{FindAllInducedUniversalGraphsForTrees}(n,k)$ \;
 \nl \KwData{Natural numbers $n$ and $k$ such that $k<n$}
@@
 
-Because of the following proposition, the symmetry breaking completion
-method finds, up to isomorphism, all induced universal graphs for the family
-of trees on $k$ vertices.
+We claim that although the symmetry breaking completion method does not
+visit every adjacency matrix visited by the naive completion method,
+it finds, up to isomorphism, all induced universal graphs for the family
+of trees on $k$ vertices.  This is shown by the following proposition.
 
@@
     $k-1$.  Its remaining vertices will be given the labels $\{2, \dots, k\}$,
-    in an order we will specify later.
+    in an order that will be specified in the next paragraph.
     The remaining $n-k$ vertices of $G$ induce a subgraph isomorphic
@@
 Table~\ref{tab:treeresults} gives the order $t(k)$ of a minimal induced universal graph for
-the family of $k$-vertex trees, and the number $T(k)$ of such graphs.  Figure~\ref{fig:trees}
-shows one of the 66 minimal induced universal graphs for the family of 6-vertex trees.
-
-To obtain upper bounds on $t(k)$ for larger values of $k$, we use a version of
-\Cref{alg:heuristic} with minor modifications.  Instead of initially embedding
-a clique and an independent set as in \Cref{fig:heuristic-regions}, we embed
-a single tree --- the $k$-vertex star $S_{k-1}$ --- as shown in
-\Cref{fig:heuristic-regions-for-trees}.  Each possible edge marked with a dot
-in the figure is added with probability 0.1.
-
-\begin{figure}[h!]
-    \centering
-    \footnotesize
-%    \tt
-       \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
-        \tikzstyle{every node}=[]
-          \input{img/heuristic-regions-for-trees}
-      \end{tikzpicture}
-    \caption{The adjacency matrix used in our heuristic algorithm when searching
-        for an induced universal graph of order 9 for the set of all trees
-        of order 6.  The blue
-        region, a star with 6 vertices, remains unchanged as the algorithm runs.
-        The remaining possible edges, marked $\boldsymbol{\cdot}$,
-        are flipped between 0 and 1 as the algorithm progresses.}
-\label{fig:heuristic-regions-for-trees}
-\end{figure}
-
-Using this heuristic algorithm, we were able to find induced universal graphs
-to demonstrate that $t(8) \leq 13$, $t(9) \leq 15$ and $t(10) \leq 18$.
-Since we have matching lower and upper bounds for $t(8)$, we can state
-that $t(8) = 13$.
-
-\section{Verification}\label{sec:verification}
-
-To verify correctness of the results in
-\Cref{tab:graphresults} and \Cref{tab:treeresults}, we repeated every call to
-the McSplit subgraph isomorphism solver using the LAD algorithm \cite{DBLP:journals/ai/Solnon10}
-as implemented in the \texttt{igraph} library \cite{igraph}.
-
-We verified that the graphs shown as adjacency matrices in \Cref{fig:adjmat14}
-and \Cref{fig:adjmat18} are induced universal graphs for their corresponding families
-using a shell script that calls the Glasgow Subgraph Solver \cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/gg/McCreeshP020} to check for the
-inclusion of every graph of $k$ vertices.
-This script shares no code with the Python program that generated the graphs.
-
-Finally, we verified all but the final row in each of \Cref{tab:graphresults}
-and \Cref{tab:treeresults} using a second shell script that calls the Glasgow
-Subgraph Solver.  Again, this shares no code with the program used to generate
-the results in the tables.
-
-\section{Acknowledgements}
-
-I would like to thank Brendan McKay for helpful feedback on this paper, and Persi
-Diaconis for introducing me to induced universal graphs and for interesting email
-discussions on related topics.
+the family of $k$-vertex trees, and the number $T(k)$ of such graphs.  Figure~\ref{fig:tree-example}
+shows one of the 687 minimal induced universal graphs for the family of 7-vertex trees.
 
@@
     \centering
-\begin{tikzpicture}[>=latex',line join=bevel,scale=.28]
-  \pgfsetlinewidth{.5bp}
-%%
-\pgfsetcolor{black}
-  % Edge: 6 -- 0
-  \draw [] (33.468bp,122.33bp) .. controls (44.467bp,128.89bp) and (59.197bp,137.66bp)  .. (70.288bp,144.27bp);
-  % Edge: 6 -- 1
-  \draw [] (30.834bp,100.1bp) .. controls (39.774bp,91.035bp) and (51.646bp,78.993bp)  .. (60.571bp,69.941bp);
-  % Edge: 7 -- 0
-  \draw [] (108.35bp,211.0bp) .. controls (103.69bp,198.96bp) and (97.434bp,182.81bp)  .. (92.754bp,170.73bp);
-  % Edge: 7 -- 2
-  \draw [] (128.78bp,216.26bp) .. controls (138.81bp,207.93bp) and (152.29bp,196.74bp)  .. (162.31bp,188.42bp);
-  % Edge: 7 -- 3
-  \draw [] (111.85bp,245.92bp) .. controls (109.9bp,257.68bp) and (107.35bp,272.94bp)  .. (105.4bp,284.67bp);
-  % Edge: 8 -- 0
-  \draw [] (134.88bp,109.4bp) .. controls (124.44bp,118.9bp) and (110.07bp,131.98bp)  .. (99.64bp,141.47bp);
-  % Edge: 8 -- 1
-  \draw [] (132.46bp,88.608bp) .. controls (119.73bp,81.799bp) and (101.96bp,72.297bp)  .. (89.227bp,65.491bp);
-  % Edge: 8 -- 2
-  \draw [] (154.41bp,114.43bp) .. controls (159.1bp,127.91bp) and (165.58bp,146.51bp)  .. (170.24bp,159.91bp);
-  % Edge: 8 -- 4
-  \draw [] (166.38bp,93.965bp) .. controls (179.82bp,91.611bp) and (198.11bp,88.409bp)  .. (211.53bp,86.058bp);
-  % Edge: 8 -- 5
-  \draw [] (152.57bp,79.558bp) .. controls (155.7bp,66.447bp) and (159.96bp,48.609bp)  .. (163.09bp,35.509bp);
-  % Node: 0
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (86.17bp,153.73bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 1
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (73.34bp,56.99bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 2
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (176.16bp,176.91bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 3
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (102.42bp,302.6bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 4
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (229.48bp,82.92bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 5
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (167.27bp,18.0bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 6
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (18.0bp,113.12bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 7
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (114.87bp,227.81bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-  % Node: 8
-\begin{scope}
-  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
-  \pgfsetstrokecolor{strokecol}
-  \draw (148.38bp,97.12bp) ellipse (18.0bp and 18.0bp);
-\end{scope}
-%
-\end{tikzpicture}
-
+    \input{img/tree-example}
 \caption{An induced universal graph for the family of all
-trees with 6 vertices}
-\label{fig:trees}
+    trees on 7 vertices.  In each copy of the graph, one of the
+    11 trees of order 7 is highlighted.}.
+\label{fig:tree-example}
 \end{figure}
+
+To obtain upper bounds on $t(k)$ for larger values of $k$, we use a version of
+\Cref{alg:heuristic} with minor modifications.  Instead of initially embedding
+a clique and an independent set as in \Cref{fig:heuristic-regions}, we embed
+a single tree --- the $k$-vertex star $S_{k-1}$ --- as shown in
+\Cref{fig:heuristic-regions-for-trees}.  Each possible edge marked with a dot
+in the figure is added with probability 0.1.
+
+\begin{figure}[h!]
+    \centering
+    \footnotesize
+%    \tt
+       \begin{tikzpicture}[>=latex',line join=bevel,scale=.4]
+        \tikzstyle{every node}=[]
+          \input{img/heuristic-regions-for-trees}
+      \end{tikzpicture}
+    \caption{The adjacency matrix used in our heuristic algorithm when searching
+        for an induced universal graph of order 9 for the set of all trees
+        of order 6.  The blue
+        region, a star with 6 vertices, remains unchanged as the algorithm runs.
+        The remaining possible edges, marked $\boldsymbol{\cdot}$,
+        are flipped between 0 and 1 as the algorithm progresses.}
+\label{fig:heuristic-regions-for-trees}
+\end{figure}
+
+Using this heuristic algorithm, we were able to find induced universal graphs
+to demonstrate that $t(8) \leq 13$, $t(9) \leq 15$ and $t(10) \leq 18$.
+Since we have matching lower and upper bounds for $t(8)$, we can state
+that $t(8) = 13$.
+
+\section{Verification}\label{sec:verification}
+
+To verify correctness of the results in
+\Cref{tab:graphresults} and \Cref{tab:treeresults}, we repeated every call to
+the McSplit subgraph isomorphism solver using the LAD algorithm \cite{DBLP:journals/ai/Solnon10}
+as implemented in the \texttt{igraph} library \cite{igraph}.
+
+We verified that the graphs shown as adjacency matrices in \Cref{fig:adjmat14}
+and \Cref{fig:adjmat18} are induced universal graphs for their corresponding families
+using a shell script that calls the Glasgow Subgraph Solver \cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/gg/McCreeshP020} to check for the
+inclusion of every graph of $k$ vertices.
+This script shares no code with the Python program that generated the graphs.
+
+Finally, we verified all but the final row in each of \Cref{tab:graphresults}
+and \Cref{tab:treeresults} using a second shell script that calls the Glasgow
+Subgraph Solver.  Again, this shares no code with the program used to generate
+the results in the tables.
+
+\section{Acknowledgements}
+
+I would like to thank Brendan McKay for helpful feedback on this paper, and Persi
+Diaconis for introducing me to induced universal graphs and for interesting email
+discussions on related topics.
 

### 2021-10-25 glasgow-subgraph-solver-check/check.sh
Add Glasgow checker

@@ -1,0 +1,22 @@
+#!/bin/bash
+
+set -euf -o pipefail
+
+type=$1
+nP=$2
+nT=$3
+
+cat ../graphs/graph${nT}.g6 | while read -r T; do
+    echo $T | ../showg -e > target.txt
+    awk -f to_dimacs_format.awk < target.txt > target.dimacs
+    rm -f results.txt
+    cat ../graphs/${type}${nP}.g6 | while read -r P; do
+        echo $P | ../showg -e > pattern.txt
+        awk -f to_dimacs_format.awk < pattern.txt > pattern.dimacs
+        ./glasgow_subgraph_solver --induced --format dimacs pattern.dimacs target.dimacs | grep 'status =' >> results.txt
+    done
+
+    if ! grep -q 'false' results.txt; then
+        echo $T
+    fi
+done

### 2021-10-25 glasgow-subgraph-solver-check/check_adjmat.sh

@@ -1,0 +1,17 @@
+#!/bin/bash
+
+set -euf -o pipefail
+
+filename=$1
+type=$2
+nP=$3
+
+awk -f adjmat_to_dimacs.awk < $filename > target.dimacs 
+rm -f results.txt
+cat ../graphs/${type}${nP}.g6 | while read -r P; do
+    echo $P | ../showg -e > pattern.txt
+    awk -f to_dimacs_format.awk < pattern.txt > pattern.dimacs
+    ./glasgow_subgraph_solver --induced --format dimacs pattern.dimacs target.dimacs | grep 'status =' >> results.txt
+done
+
+cat results.txt | sort | uniq -c

### 2021-10-25 glasgow-subgraph-solver-check/check_exhaustive_tests.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+
+cat ../exhaustive_tests.txt | grep -v 'graph 5' | grep -v 'tree 7' | while read a b c; do
+    echo $a $b $c
+#    ./check.sh $a $b $c | tee results/$a-$b-$c.txt
+    diff results/$a-$b-$c.txt ../results/$a-$b-$c.txt
+done

### 2021-10-25 make-backtracking-tree-algorithm-fig.py
Tweak colours

@@ -28,10 +28,10 @@ for i in range(n):
             adjmat[i][j] = "$\\boldsymbol{\\cdot}$"
             adjmat[j][i] = "$\\boldsymbol{\\cdot}$"
 
-print("\draw[uofgcobalt,fill=uofgcobalt,fill opacity=0.15, thick] (-.4,-6.4) rectangle (6.4,.4);")
-print("\draw[uofgpumpkin,fill=uofgpumpkin,fill opacity=0.15, thick] (6.6,-6.4) rectangle (10.4,-.4);")
-#print("\draw[uofgpumpkin,fill=uofgpumpkin,fill opacity=0.15, thick] (.4,-10.4) rectangle (6.4,-6.6);")
-print("\draw[uofgslate,fill=uofgslate,fill opacity=0.15, thick] (6.6,-10.4) rectangle (10.4,-6.6);")
+print("\draw[uofgcobalt,fill=uofgcobalt,fill opacity=0.15, thick] (-.45,-6.45) rectangle (6.45,.45);")
+print("\draw[uofgpumpkin,fill=uofgpumpkin,fill opacity=0.15, thick] (6.55,-6.45) rectangle (10.45,-.45);")
+#print("\draw[uofgpumpkin,fill=uofgpumpkin,fill opacity=0.15, thick] (.45,-10.45) rectangle (6.45,-6.55);")
+print("\draw[uofgslate,fill=uofgslate,fill opacity=0.15, thick] (6.55,-10.45) rectangle (10.45,-6.55);")
 
 for i in range(n):
     for j in range(n):

### 2021-10-25 find_universal_graphs_backtracking_for_trees.py
IGraph check for tree algorithm

@@ -2,6 +2,11 @@ import sys
 
 from universalgraphs import Graph, induced_subgraph_isomorphism, from_graph6_bytes
 
+try:
+    import igraph
+except:
+    pass
+
 def read_all_graphs(n):
     with open("graphs/graph{}.g6".format(n), "rb") as f:
         for line in f:
@@ -35,8 +40,25 @@ def make_target_graph(nT, nP, vals, small_graph):
     return G
 
     
+def to_igraph_graph(G):
+    g = igraph.Graph()
+    g.add_vertices(G.number_of_nodes())
+    for v in range(G.number_of_nodes()):
+        for w in range(v):
+            if G.adj_row(v)[w]:
+                g.add_edges([(v, w)])
+    return g
+
+
+def igraph_induced_subgraph_isomorphism(G, H):
+    g = to_igraph_graph(G)
+    h = to_igraph_graph(H)
+    return h.subisomorphic_lad(g, induced=True)
+
 def iso(P, T):
-    return induced_subgraph_isomorphism(P, T)
+    result = bool(induced_subgraph_isomorphism(P, T))
+    assert result == igraph_induced_subgraph_isomorphism(P, T)
+    return result
 
 
 def isomorphic(G, H):
